### 2016-06-06 AnalysisCode/burden_mC.Rmd
first commit

---
title: "burden_mC"
author: "Tim Stuart"
date: "6 June 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import}
library(dplyr)
```

```{r load_data}
mC <- read.table("../data/c_dmr_mc.tsv", header = T)
TE_DMR <- read.table("../data/")
```

```{r functions}
gatherData <- function(insertions, ranks) {
  data <- vector()
  for(i in 1:nrow(insertions)){
    row <- insertions[i,]
    gene <- as.character(row[[1]])
    accessions <- unlist(strsplit(gsub("-", "_", as.character(row[[3]])), ","))
    accessions <- intersect(accessions, colnames(ranks))
    if(length(accessions) > 0){
      for(acc in accessions){
        data <- c(data, as.numeric(ranks[gene,acc]))
      }
    }
  }
  return(data)
}

permuteLabels <- function(ranks, insertions) {
  data <- vector()
  all_names <- colnames(ranks)
  for(i in 1:nrow(insertions)){
    row <- insertions[i,]
    gene <- as.character(row[[1]])
    accessions <- unlist(strsplit(gsub("-", "_", as.character(row[[3]])), ","))
    accessions <- intersect(accessions, colnames(ranks))
    if(length(accessions) > 0){
      rand <- sample(all_names, length(accessions), replace = FALSE)
      for(acc in rand){
        data <- c(data, as.numeric(ranks[gene,acc]))
      }
    }
  }
  return(data)
}
```

### 2016-06-06 AnalysisCode/closest_DMRs.sh

@@ -1,0 +1,7 @@
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
+| bedtools closest -b - -a ../RawData/c_dmrs.tsv.gz -d -wb \
+> ../ProcessedData/TE_C_DMR_distances.bed
+
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
+| bedtools closest -b - -a ../RawData/cg_dmrs.tsv.gz -d -wb \
+> ../ProcessedData/TE_CG_DMR_distances.bed

### 2016-06-06 DataAquisition/TEPID_master_script.sh

@@ -1,0 +1,39 @@
+# This script will run TEPID on all PE reads from Schmitz et al. 2013
+# and identify TE insertion and TE absences with repect to the reference genome
+
+cd ../RawData
+
+# download reference genome
+wget ftp://igenome:G3nom3s4u@ussd-ftp.illumina.com/Arabidopsis_thaliana/NCBI/TAIR10/Arabidopsis_thaliana_NCBI_TAIR10.tar.gz
+tar -xvzf Arabidopsis_thaliana_NCBI_TAIR10.tar.gz
+
+# build reference index
+bowtie2-build TAIR10/Sequence/WholGenomeFasta/genome.fa tair10
+yaha -g TAIR10/Sequence/WholGenomeFasta/genome.fa
+
+# using one core for mapping
+tepid-map -x tair10 -y genome -p 1 -s 200 -r -z
+
+# merge all calls
+python merge_deletions.py -f deletions
+python merge_insertions.py -f insertions
+python create_poly_te.py
+
+# run refine step
+map () {
+  for myfile in $(ls -d *.split.bam);do
+    fname=(${myfile//.split.bam/ })
+    tepy-refine -a ../../../accession_names.tsv -p 10 -i ../../insertions.bed -d ../../deletions.bed -n $fname -c "${fname}_filtered.bam" -s $myfile -t ../../../TAIR9_TE.bed.gz
+  done
+}
+
+for directory in ./*; do
+    if [ -d "$directory" ]; then
+        cd $directory
+        map
+        cd ..
+    fi
+done
+
+
+python genotype.py -i -a ambiguous -m merged_insertions.

### 2016-06-06 DataAquisition/build_tables.py

@@ -1,0 +1,102 @@
+#! /usr/bin/env python
+
+# run from directory containing mC_calls files
+
+import os
+from sys import argv
+import MySQLdb
+from argparse import ArgumentParser
+
+
+"""
+MySQL commands:
+create database population_epigenetics;
+use population_epigenetics;
+create table mC_calls_Aa_0 (assembly int(1), position int(9), strand char(1), class char(3), mc int(3), h int(3), uc int(3));
+insert into mC_calls_Aa_0 (assembly, position, strand, class, mc, h, uc) values (1, 10, '+', 'CHH', 4, 5, 1);
+...
+alter table mC_calls_Aa_0 add index (position);
+"""
+
+def create_database(options):
+    link = MySQLdb.connect(options.host, options.user, options.password)
+    cursor = link.cursor()
+    query = "create database {db};".format(db=options.database)
+    cursor.execute(query)
+    cursor.close()
+    link.close()
+
+def add_table(options, fname):
+    """Make new table and add data"""
+    with open(fname, 'r') as infile:
+        link = MySQLdb.connect(options.host, options.user, options.password)
+        cursor = link.cursor()
+        accession_name = fname.replace('calls_', '.').split('.')
+        accession_name = accession_name[1]
+        cursor.execute("use {db};".format(db=options.database))
+        build = """create table mC_calls_{name} (assembly int(1), position int(9), strand char(1), class char(3), mc int(3), h int(3), uc int(3));""".format(name=accession_name)
+        cursor.execute(build)
+        for line in infile:
+            line = line.rsplit()
+            chrom = line[0]
+            if chrom == 'chrom':
+                pass  # header
+            else:
+                chrom = int(chrom)
+                pos = int(line[1])
+                strand = line[2]
+                mc_class = line[3]
+                if mc_class[:2] == 'CG':
+                    mc_class = 'CG'
+                else:
+                    mc_class = mc_class[0] + mc_class[1:].replace('A', 'H').replace('T', 'H').replace('C', 'H')
+                mc = int(line[4])
+                h = int(line[5])
+                uc = h - mc
+                add_data = """insert into mC_calls_{name} (assembly, position, strand, class, mc, h, uc) values ({chrom}, {pos}, '{strand}', '{mc_class}', {mc}, {h}, {uc});""".format(name=accession_name,
+                                                                                                                                                                                       chrom=chrom,
+                                                                                                                                                                                       pos=pos,
+                                                                                                                                                                                       strand=strand,
[…]

### 2016-06-06 DataAquisition/source_PE_data.sh

@@ -1,0 +1,26 @@
+# This file will download PE sequencing data for all Arabidopsis accessions
+# in Schmitz et al. 2013, and extract the fastq files
+
+cd ../RawData
+while read acc; do
+  echo $acc
+  wget -r --no-parent --no-directories "ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByRun/sra/SRR/SRR492/${acc}/*"
+done < ./SraAccList.txt
+
+for myfile in ./*.sra; do
+  fname=(${myfile//.sra/ })
+  mkdir $fname
+  mv $myfile $fname
+done
+
+for directory in ./*; do
+    if [ -d "$directory" ]; then
+        cd $directory
+        for myfile in $(ls -d *.sra);do
+            fastq-dump --split-3 -v $myfile
+            rm -f $myfile
+        done
+        cd ..
+    fi
+done
+cd ../DataAquisition

### 2016-06-06 DataAquisition/source_mC_data.sh

@@ -1,0 +1,12 @@
+# this file will download DNA methylation data for accessions used in Schmitz et al. 2013
+# and create a MySQL database to hold the data
+# The script must be edited before running to include MySQL login details
+
+mkdir mC
+cd mC
+wget ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE43nnn/GSE43857/suppl/GSE43857_RAW.tar
+tar -x -f GSE43857_RAW.tar
+gunzip *.gz
+rm GSE43857_RAW.tar
+python ../build_tables.py -u [username] -p [mysql_password] -d population_epigenetics --host [host]
+cd ..

### 2016-06-06 ProcessingCode/classify_TE_absence.R

@@ -1,0 +1,16 @@
+library(dplyr)
+
+poly <- read_tsv("../ProcessedData/poly_te_genotyped.tsv", col_names = F)
+
+polarized <- poly %>%
+  rowwise() %>%
+  mutate(TE_present = length(unlist(strsplit(X5, ","))),
+         TE_absent = length(unlist(strsplit(X6, ","))),
+         Minor_allele = ifelse(TE_present < TE_absent, "presence", "absence"),
+         Absence_classification = ifelse(Minor_allele == "absence", "True deletion", "No insertion"),
+         MAF = min(TE_present, TE_absent) / 216) %>%
+  ungroup()
+
+colnames(polarized) <- c("chromosome", "start", "end", "TE", "Accessions_TE_present", "Accessions_TE_absent", "Count_TE_present", "Count_TE_absent", "Minor_allele", "Absence_classification", "MAF")
+write.table(polarized, file = "../RawData/tepav_regenerated.tsv",
+            quote = F, sep = "\t", na = "NA", row.names = F, col.names = T)

### 2016-06-06 DataAquisition/TEPID_master_script.sh
add discover step

@@ -12,7 +12,24 @@ bowtie2-build TAIR10/Sequence/WholGenomeFasta/genome.fa tair10
 yaha -g TAIR10/Sequence/WholGenomeFasta/genome.fa
 
 # using one core for mapping
-tepid-map -x tair10 -y genome -p 1 -s 200 -r -z
+tepid-map -x tair10 -y ./TAIR10/Sequence/WholeGenomeFasta/genome.X15_01_65525S -p 1 -s 200 -r -z
+
+
+discover () {
+  for myfile in $(ls -d *.split.bam);do
+    fname=(${myfile//.split.bam/ })
+    tepy-discover -n $fname -c "${fname}.bam" -s $myfile -t ../../../TAIR9_TE.bed.gz
+  done
+}
+
+# discover TEs
+for directory in ./*; do
+    if [ -d "$directory" ]; then
+        cd $directory
+        discover
+        cd ..
+    fi
+done
 
 # merge all calls
 python merge_deletions.py -f deletions
@@ -20,10 +37,10 @@ python merge_insertions.py -f insertions
 python create_poly_te.py
 
 # run refine step
-map () {
+refine () {
   for myfile in $(ls -d *.split.bam);do
     fname=(${myfile//.split.bam/ })
-    tepy-refine -a ../../../accession_names.tsv -p 10 -i ../../insertions.bed -d ../../deletions.bed -n $fname -c "${fname}_filtered.bam" -s $myfile -t ../../../TAIR9_TE.bed.gz
+    tepy-refine -a ../../../accession_names.tsv -i ../../insertions.bed -d ../../deletions.bed -n $fname -c "${fname}.bam" -s $myfile -t ../../../TAIR9_TE.bed.gz
   done
 }
 
@@ -34,6 +51,5 @@ for directory in ./*; do
         cd ..
     fi
 done
-
 
 python genotype.py -i -a ambiguous -m merged_insertions.

### 2016-06-06 DataAquisition/TEPID_master_script.sh
correct path

@@ -8,8 +8,8 @@ wget ftp://igenome:G3nom3s4u@ussd-ftp.illumina.com/Arabidopsis_thaliana/NCBI/TAI
 tar -xvzf Arabidopsis_thaliana_NCBI_TAIR10.tar.gz
 
 # build reference index
-bowtie2-build TAIR10/Sequence/WholGenomeFasta/genome.fa tair10
-yaha -g TAIR10/Sequence/WholGenomeFasta/genome.fa
+bowtie2-build Arabidopsis_thaliana/NCBI/TAIR10/Sequence/WholGenomeFasta/genome.fa tair10
+yaha -g Arabidopsis_thaliana/NCBI/TAIR10/Sequence/WholGenomeFasta/genome.fa
 
 # using one core for mapping
 tepid-map -x tair10 -y ./TAIR10/Sequence/WholeGenomeFasta/genome.X15_01_65525S -p 1 -s 200 -r -z

### 2016-06-06 AnalysisCode/gene_expression_burden.R
Added RNA expression analysis scripts

@@ -1,0 +1,167 @@
+library(dplyr)
+
+set.seed(1)
+
+# this data only has accession names with RNA data available
+expression <- read.table("../RawData/gene_expression.tsv.gz", sep = ""), header = T, row.names = 1)
+upstream <- read.table('../ProcessedData/upstream_insertion.tsv', sep = ""), header = F)
+exon <- read.table('../ProcessedData/exon_insertion.tsv', sep = ""), header = F)
+intron <- read.table("../ProcessedData/intron_insertion.tsv", sep = ""), header = F)
+utr5 <- read.table("../ProcessedData/utr5_insertion.tsv", sep = ""), header = F)
+utr3 <- read.table("../ProcessedData/utr3_insertion.tsv", sep = ""), header = F)
+downstream <- read.table("../ProcessedData/downstream_insertion.tsv", sep = ""), header = F)
+
+upstream <- upstream[!duplicated(upstream),]
+exon <- exon[!duplicated(exon),]
+intron <- intron[!duplicated(intron),]
+utr5 <- utr5[!duplicated(utr5),]
+utr3 <- utr3[!duplicated(utr3),]
+downstream <- downstream[!duplicated(downstream),]
+
+header <- c("gene", 'TE', 'pos_accessions', 'neg_accessions')
+colnames(upstream) <- header
+colnames(exon) <- header
+colnames(intron) <- header
+colnames(utr3) <- header
+colnames(utr5) <- header
+colnames(downstream) <- header
+
+upstream[] <- lapply(upstream, as.character)
+exon[] <- lapply(exon, as.character)
+utr3[] <- lapply(utr3, as.character)
+utr5[] <- lapply(utr5, as.character)
+intron[] <- lapply(intron, as.character)
+downstream[] <- lapply(downstream, as.character)
+
+find.len <- function(d) {
+  return(length(unlist(strsplit(d, ","))))
+}
+
+upstream %>%
+  rowwise() %>%
+  filter(find.len(pos_accessions) <= 4) -> upstream.five
+
+exon %>%
+  rowwise() %>%
+  filter(find.len(pos_accessions) <= 4) -> exon.five
+
+intron %>%
+  rowwise() %>%
+  filter(find.len(pos_accessions) <= 4) -> intron.five
+
+utr3 %>%
+  rowwise() %>%
+  filter(find.len(pos_accessions) <= 4) -> utr3.five
[…]

### 2016-06-06 AnalysisCode/gene_expression_volcano.R

@@ -1,0 +1,151 @@
+library(dplyr)	
+library(ggplot2)	
+library(qvalue)	
+
+data_path = "/Users/timstuart/Dropbox/PhD/Projects/Population_epigenetics/RNA/"	
+
+# this data only has accession names with RNA data available	
+expression <- read.table(paste(data_path, "expression.tsv", sep = ""), header = T, row.names = 1)	
+upstream <- read.table(paste(data_path, 'upstream_insertion.tsv', sep = ""), header = F)	
+exon <- read.table(paste(data_path, 'exon_insertion.tsv', sep = ""), header = F)	
+intron <- read.table(paste(data_path, "intron_insertion.tsv", sep = ""), header = F)	
+utr5 <- read.table(paste(data_path, "utr5_insertion.tsv", sep = ""), header = F)	
+utr3 <- read.table(paste(data_path, "utr3_insertion.tsv", sep = ""), header = F)	
+downstream <- read.table(paste(data_path, "downstream_insertion.tsv", sep = ""), header = F)	
+
+upstream <- upstream[!duplicated(upstream),]	
+exon <- exon[!duplicated(exon),]	
+intron <- intron[!duplicated(intron),]	
+utr5 <- utr5[!duplicated(utr5),]	
+utr3 <- utr3[!duplicated(utr3),]	
+downstream <- downstream[!duplicated(downstream),]	
+
+header <- c("gene", 'TE', 'pos_accessions', 'neg_accessions')	
+colnames(upstream) <- header	
+colnames(exon) <- header	
+colnames(intron) <- header	
+colnames(utr3) <- header	
+colnames(utr5) <- header	
+colnames(downstream) <- header	
+	
+upstream[] <- lapply(upstream, as.character)	
+exon[] <- lapply(exon, as.character)	
+utr3[] <- lapply(utr3, as.character)	
+utr5[] <- lapply(utr5, as.character)	
+intron[] <- lapply(intron, as.character)	
+downstream[] <- lapply(downstream, as.character)	
+	
+find.len <- function(d) {	
+  return(length(unlist(strsplit(d, ","))))	
+}	
+	
+# filter where insertion in at least 5 accessions (MAF >3%)	
+upstream %>%	
+  rowwise() %>%	
+  filter(find.len(pos_accessions) > 4) -> upstream.five	
+	
+exon %>%	
[…]

### 2016-06-06 AnalysisCode/resistance_phenotype.R

@@ -1,0 +1,43 @@
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(reshape2)
+library(RColorBrewer)
+
+download.file("http://journals.plos.org/plosgenetics/article/asset?unique&id=info:doi/10.1371/journal.pgen.0010060.sd002",
+              destfile = "../RawData/resistance_dataset.csv") 
+data <- read_csv("../RawData/resistance_dataset.csv", col_names = FALSE)
+tepav <- read_tsv("../RawData/TEPID_TEPAV.tsv.gz", col_names = TRUE)
+
+te.var <- tepav %>%
+  filter(TE == "AT3TE29460")
+
+pos.acc <- te.var[2,5][[1]]
+pos <- strsplit(pos.acc, ",")[[1]]
+
+names <- tepav[1,5:6]
+accessions <- strsplit(names[[2]], ",")
+all_schmitz <- gsub("_", "-", accessions[[1]])
+
+data %>%
+  filter(X1 %in% all_schmitz) %>%
+  mutate(TE = ifelse(X1 %in% pos, TRUE, FALSE)) -> filtered_data
+
+filtered_data <- transform(filtered_data, X6 = as.numeric(X6))
+filtered_data <- transform(filtered_data, X7 = as.numeric(X7))
+filtered_data <- transform(filtered_data, X8 = as.numeric(X8))
+filtered_data <- transform(filtered_data, X9 = as.numeric(X9))
+
+filtered_data %>%
+  group_by(TE) %>%
+  summarise(avrPpH3 = mean(X6, na.rm = T) * 100, avrRpm1 = mean(X7, na.rm = T) * 100,
+            avrRpt2 = mean(X8, na.rm = T) * 100, avrB = mean(X9, na.rm = T) * 100, sample_size = n()) -> stats
+
+reshaped <- head(melt(stats), -2)
+
+ggplot(reshaped, aes(variable, value, TE)) +
+  geom_bar(stat = "identity", aes(fill = TE), position = "dodge", color = "black", size = 0.2) + theme_bw() +
+  ylim(0, 100) + ylab("Accessions resistant (%)") + xlab("avr genes") +
+  scale_fill_brewer("Blues") + ggtitle("P. syringae resistance") + 
+  theme(legend.position = "null") +
+  ggsave(height=10,width=10, units = "cm", path="../Plots", filename="resistance.pdf", useDingbats=FALSE)

### 2016-06-07 AnalysisCode/mC_heatmap.R
Add annotation files and some analysis code

@@ -1,0 +1,56 @@
+library(dplyr)
+library(gplots)
+library(RColorBrewer)
+library(readr)
+
+mc_ins_2kb <- read_tsv("../ProcessedData/allC_ins_2kb.tsv", col_names = F)
+mc_del_2kb <- read_tsv("../ProcessedData/allC_true_del_2kb.tsv", col_names = F)
+
+
+### old code ###
+### re-write for updated analysis ###
+
+# sum rows for sort order
+mc_all_ins_2kb_dist.sums <- mutate(mc_all_ins_2kb_dist, sums=rowSums(mc_all_ins_2kb_dist[,2:41]))
+
+
+# categorical heatmap
+cat_hmap <- function(d) {
+  heatmap.2(d,Colv=NA,
+            #col=brewer.pal(9, "Dark2"),
+            col=brewer.pal(4, "Spectral"),
+            labRow = "",
+            labCol = "",
+            dendrogram = "none",
+            trace = "none",
+            par(cex.main=0.7,
+                cex.lab=0.7,
+                cex.axis=0.7
+            ),
+            cex.lab=0.7)
+}
+
+# heatmap
+hmap <- function(d) {
+  heatmap.2(as.matrix(d),Colv=NA,col=brewer.pal(9, "Reds"),
+            labRow = "",
+            labCol = "",
+            dendrogram = "none",
+            margins = c(5,2),
+            trace = "none",
+            par(cex.main=0.7,
+                cex.lab=0.7,
+                cex.axis=0.7
+            ),
+            cex.lab=0.7)
+}
+
+# scale values to max 0.5
+eu.floor <- eu > 0.5
+eu[eu.floor] <- 0.5
+het.floor <- het > 0.5
+het[het.floor] <- 0.5
+
+pdf("../Plots/heatmap_mc.pdf", height = 6, width = 3)
+hmap(data)
+dev.off()

### 2016-06-07 AnalysisCode/quadrana_overlap.sh

@@ -1,0 +1,7 @@
+# finds number of TE variants reported by Quadrana et al. are within 1 kb of our TE variants.
+
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz \
+| sed 1d \
+| bedtools closest -d -b - -a ../RawData/Quadrana_TEPAV.tsv.gz \
+| awk '$16 < 1000 {print $0}' \
+| wc -l

### 2016-06-07 DataAquisition/TEPID_master_script.sh
fix paths

@@ -2,6 +2,8 @@
 # and identify TE insertion and TE absences with repect to the reference genome
 
 cd ../RawData
+# in RawData
+# Assuming source_PE_data.sh has been run and the PE reads are in folders in the RawData folder
 
 # download reference genome
 wget ftp://igenome:G3nom3s4u@ussd-ftp.illumina.com/Arabidopsis_thaliana/NCBI/TAIR10/Arabidopsis_thaliana_NCBI_TAIR10.tar.gz
@@ -18,7 +20,7 @@ tepid-map -x tair10 -y ./TAIR10/Sequence/WholeGenomeFasta/genome.X15_01_65525S -
 discover () {
   for myfile in $(ls -d *.split.bam);do
     fname=(${myfile//.split.bam/ })
-    tepy-discover -n $fname -c "${fname}.bam" -s $myfile -t ../../../TAIR9_TE.bed.gz
+    tepy-discover -n $fname -c "${fname}.bam" -s $myfile -t ../TAIR9_TE.bed.gz
   done
 }
 
@@ -26,8 +28,10 @@ discover () {
 for directory in ./*; do
     if [ -d "$directory" ]; then
         cd $directory
+        # in RawData/Accession
         discover
         cd ..
+        # in RawData
     fi
 done
 
@@ -39,16 +43,19 @@ python create_poly_te.py
 # run refine step
 refine () {
   for myfile in $(ls -d *.split.bam);do
+    # in RawData/Accession
     fname=(${myfile//.split.bam/ })
-    tepy-refine -a ../../../accession_names.tsv -i ../../insertions.bed -d ../../deletions.bed -n $fname -c "${fname}.bam" -s $myfile -t ../../../TAIR9_TE.bed.gz
+    tepy-refine -a ../accession_names.tsv -i ../insertions.bed -d ../deletions.bed -n $fname -c "${fname}.bam" -s $myfile -t ../TAIR9_TE.bed.gz
   done
 }
 
 for directory in ./*; do
     if [ -d "$directory" ]; then
         cd $directory
-        map
+        # in RawData/Accession
+        refine
         cd ..
+        # in RawData
     fi
 done
 

### 2016-06-07 ProcessingCode/get_mc_DMRs.py
add DMR processing code

@@ -1,0 +1,71 @@
+#! /usr/bin/python
+
+from __future__ import division
+import MySQLdb
+from argparse import ArgumentParser
+
+
+parser = ArgumentParser(description='Get mC within coordinates for multiple samples')
+parser.add_argument('-x', '--host', help='host', required=True)
+parser.add_argument('-u', '--user', help='mySQL user', required=True)
+parser.add_argument('-p', '--password', help='mySQL password', required=True)
+parser.add_argument('-d', '--database', help='mySQL database name', required=True)
+parser.add_argument('-f', '--file', help='input coordinate file', required=True)
+options = parser.parse_args()
+
+class mcDatabase:
+
+    def __init__(self):
+        self.link = None
+
+    def setupSQL(self, host, user, password, database):
+        """ setup connection with mysql """
+        link = MySQLdb.connect(host, user, password)
+        self.link = link
+        cursor = link.cursor()
+        cursor.execute("use {};".format(database))
+        cursor.execute("show tables;")
+        results = cursor.fetchall()
+        self.tables = [row[0] for row in results]
+        cursor.close()
+
+    def gatherDataFromMySQL(self, line, context):
+        """ collect mC data from each table in database between coordinates in line """
+        cursor = self.link.cursor()
+        data = []
+        for i in self.tables:
+            query = "select sum(mc), sum(h) from {db} where assembly = {ch} and position between {s} and {e}".format(
+                db=str(i),
+                ch=line[0].strip('chr'),
+                s=line[1],
+                e=line[2])
+            cursor.execute(query)
+            results = cursor.fetchall()
+            mc = processResults(results)
+            data.append(mc)
+        cursor.close()
+        return data
+
+def processResults(results):
+    for row in results:
+        mc = row[0]
+        h = row[1]
+    if h == 0 or h is None:
+        return 0.
+    elif h > 0:
+        level = float(mc / h)
+        return round(level, 4)
+    else:
+        raise Exception('h values in table incorrect')
+
+mc = mcDatabase()
+mc.setupSQL(options.host, options.user, options.password, options.database)
+mc.tables.remove("snps")
+names =  [x.split('mC_calls_')[1] for x in mc.tables]
+print 'chr\tstart\tstop\t'+'\t'.join(names)  # header
+with open(options.file, 'r') as infile:
+    for line in infile:
+        line = line.rsplit()
+        data = mc.gatherDataFromMySQL(line, str(options.context))
[…]

### 2016-06-07 ProcessingCode/get_mc_DMRs_context.py

@@ -1,0 +1,73 @@
+#! /usr/bin/python
+
+from __future__ import division
+import MySQLdb
+from argparse import ArgumentParser
+
+
+parser = ArgumentParser(description='Get mC within coordinates for multiple samples')
+parser.add_argument('-x', '--host', help='host', required=True)
+parser.add_argument('-u', '--user', help='mySQL user', required=True)
+parser.add_argument('-p', '--password', help='mySQL password', required=True)
+parser.add_argument('-d', '--database', help='mySQL database name', required=True)
+parser.add_argument('-f', '--file', help='input coordinate file', required=True)
+parser.add_argument('-c', '--context', help='DNA methylation context', required=True)
+options = parser.parse_args()
+
+class mcDatabase:
+
+    def __init__(self):
+        self.link = None
+
+    def setupSQL(self, host, user, password, database):
+        """ setup connection with mysql """
+        link = MySQLdb.connect(host, user, password)
+        self.link = link
+        cursor = link.cursor()
+        cursor.execute("use {};".format(database))
+        cursor.execute("show tables;")
+        results = cursor.fetchall()
+        self.tables = [row[0] for row in results]
+        cursor.close()
+
+    def gatherDataFromMySQL(self, line, context):
+        """ collect mC data from each table in database between coordinates in line """
+        cursor = self.link.cursor()
+        data = []
+        for i in self.tables:
+            query = "select sum(mc), sum(h) from {db} where class in ('{context}') and assembly = {ch} and position between {s} and {e}".format(
+                db=str(i),
+                ch=line[0].strip('chr'),
+                context=context,
+                s=line[1],
+                e=line[2])
+            cursor.execute(query)
+            results = cursor.fetchall()
+            mc = processResults(results)
+            data.append(mc)
+        cursor.close()
+        return data
+
+def processResults(results):
+    for row in results:
+        mc = row[0]
+        h = row[1]
+    if h == 0 or h is None:
+        return 0.
+    elif h > 0:
+        level = float(mc / h)
+        return round(level, 4)
+    else:
+        raise Exception('h values in table incorrect')
+
+mc = mcDatabase()
+mc.setupSQL(options.host, options.user, options.password, options.database)
+mc.tables.remove("snps")
+names =  [x.split('mC_calls_')[1] for x in mc.tables]
+print 'chr\tstart\tstop\t'+'\t'.join(names)  # header
+with open(options.file, 'r') as infile:
[…]

### 2016-06-07 ProcessingCode/get_mc_DMRs.py
Allow reading from gzip file

@@ -3,7 +3,23 @@
 from __future__ import division
 import MySQLdb
 from argparse import ArgumentParser
+import gzip
 
+"""
+usage: get_mc_DMRs.py [-h] -x HOST -u USER -p PASSWORD -d DATABASE -f FILE
+
+Get mC within coordinates for multiple samples
+
+optional arguments:
+  -h, --help            show this help message and exit
+  -x HOST, --host HOST  host
+  -u USER, --user USER  mySQL user
+  -p PASSWORD, --password PASSWORD
+                        mySQL password
+  -d DATABASE, --database DATABASE
+                        mySQL database name
+  -f FILE, --file FILE  input coordinate file
+"""
 
 parser = ArgumentParser(description='Get mC within coordinates for multiple samples')
 parser.add_argument('-x', '--host', help='host', required=True)
@@ -63,9 +79,13 @@ mc.setupSQL(options.host, options.user, options.password, options.database)
 mc.tables.remove("snps")
 names =  [x.split('mC_calls_')[1] for x in mc.tables]
 print 'chr\tstart\tstop\t'+'\t'.join(names)  # header
-with open(options.file, 'r') as infile:
-    for line in infile:
-        line = line.rsplit()
-        data = mc.gatherDataFromMySQL(line, str(options.context))
-        data = (str(x) for x in data)
-        print '\t'.join(line) + '\t' + '\t'.join(data)
+if options.file.endswith(".gz"):
+    infile = gzip.open(options.file, "rb")
+else:
+    infile = open(options.file, "r")
+for line in infile:
+    line = line.rsplit()
+    data = mc.gatherDataFromMySQL(line, str(options.context))
+    data = (str(x) for x in data)
+    print '\t'.join(line) + '\t' + '\t'.join(data)
+infile.close()

### 2016-06-07 ProcessingCode/get_mc_DMRs_context.py

@@ -3,7 +3,26 @@
 from __future__ import division
 import MySQLdb
 from argparse import ArgumentParser
+import gzip
 
+"""
+usage: get_mc_DMRs_context.py [-h] -x HOST -u USER -p PASSWORD -d DATABASE -f
+                              FILE -c CONTEXT
+
+Get mC within coordinates for multiple samples
+
+optional arguments:
+  -h, --help            show this help message and exit
+  -x HOST, --host HOST  host
+  -u USER, --user USER  mySQL user
+  -p PASSWORD, --password PASSWORD
+                        mySQL password
+  -d DATABASE, --database DATABASE
+                        mySQL database name
+  -f FILE, --file FILE  input coordinate file
+  -c CONTEXT, --context CONTEXT
+                        DNA methylation context
+"""
 
 parser = ArgumentParser(description='Get mC within coordinates for multiple samples')
 parser.add_argument('-x', '--host', help='host', required=True)
@@ -65,9 +84,13 @@ mc.setupSQL(options.host, options.user, options.password, options.database)
 mc.tables.remove("snps")
 names =  [x.split('mC_calls_')[1] for x in mc.tables]
 print 'chr\tstart\tstop\t'+'\t'.join(names)  # header
-with open(options.file, 'r') as infile:
-    for line in infile:
-        line = line.rsplit()
-        data = mc.gatherDataFromMySQL(line, str(options.context))
-        data = (str(x) for x in data)
-        print '\t'.join(line) + '\t' + '\t'.join(data)
+if options.file.endswith(".gz"):
+    infile = gzip.open(options.file, "rb")
+else:
+    infile = open(options.file, "r")
+for line in infile:
+    line = line.rsplit()
+    data = mc.gatherDataFromMySQL(line, str(options.context))
+    data = (str(x) for x in data)
+    print '\t'.join(line) + '\t' + '\t'.join(data)
+infile.close()

### 2016-06-07 ProcessingCode/get_mc_dmrs.sh
add DMR processing script

@@ -1,0 +1,39 @@
+# This script gets the DNA methylation levels 
+# for C-DMRs and CG-DMRs in each accession out of the 
+# database created using previous scripts, for each
+# DNA methylation context, and for all contexts combined
+# MySQL database parameters must be added before running
+
+# C-DMR methylation levels
+python get_mc_DMRs.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz \
+> ../ProcessedData/c_dmr_allC.tsv
+
+python get_mc_DMRs_context.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CG \
+> ../ProcessedData/c_dmr_CG.tsv
+
+python get_mc_DMRs_context.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CHG \
+> ../ProcessedData/c_dmr_CHG.tsv
+
+python get_mc_DMRs_context.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CHH \
+> ../ProcessedData/c_dmr_CHH.tsv
+
+# CG-DMR methylation levels
+python get_mc_DMRs.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz \
+> ../ProcessedData/cg_dmr_allC.tsv
+
+python get_mc_DMR_context.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz -c CG \
+> ../ProcessedData/cg_dmr_CG.tsv
+
+python get_mc_DMR_context.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz -c CHG \
+> ../ProcessedData/cg_dmr_CHG.tsv
+
+python get_mc_DMR_context.py -x [host] \
+-u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz -c CHH \
+> ../ProcessedData/cg_dmr_CHH.tsv

### 2016-06-07 ProcessingCode/get_mc_dmrs.sh
fix error

@@ -6,34 +6,34 @@
 
 # C-DMR methylation levels
 python get_mc_DMRs.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz \
+-u [user] -p [password] -d [mC_database] -f ../RawData/c_dmrs.tsv.gz \
 > ../ProcessedData/c_dmr_allC.tsv
 
 python get_mc_DMRs_context.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CG \
+-u [user] -p [password] -d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CG \
 > ../ProcessedData/c_dmr_CG.tsv
 
 python get_mc_DMRs_context.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CHG \
+-u [user] -p [password] -d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CHG \
 > ../ProcessedData/c_dmr_CHG.tsv
 
 python get_mc_DMRs_context.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CHH \
+-u [user] -p [password] -d [mC_database] -f ../RawData/c_dmrs.tsv.gz -c CHH \
 > ../ProcessedData/c_dmr_CHH.tsv
 
 # CG-DMR methylation levels
 python get_mc_DMRs.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz \
+-u [user] -p [password] -d [mC_database] -f ../RawData/cg_dmrs.tsv.gz \
 > ../ProcessedData/cg_dmr_allC.tsv
 
 python get_mc_DMR_context.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz -c CG \
+-u [user] -p [password] -d [mC_database] -f ../RawData/cg_dmrs.tsv.gz -c CG \
 > ../ProcessedData/cg_dmr_CG.tsv
 
 python get_mc_DMR_context.py -x [host] \
--u [user] -p [password] - d [mC_database] -f ../RawData/cg_dmrs.tsv.gz -c CHG \
[…]

### 2016-06-07 ProcessingCode/classify_TE_absence.R
update deletion classification

@@ -7,7 +7,13 @@ polarized <- poly %>%
   mutate(TE_present = length(unlist(strsplit(X5, ","))),
          TE_absent = length(unlist(strsplit(X6, ","))),
          Minor_allele = ifelse(TE_present < TE_absent, "presence", "absence"),
-         Absence_classification = ifelse(Minor_allele == "absence", "True deletion", "No insertion"),
+         Absence_classification = if(TE_absent < 43) {
+           "True deletion"
+         } else if(TE_absent > 172) {
+           "No insertion"
+         } else {
+           "NA"
+         },
          MAF = min(TE_present, TE_absent) / 216) %>%
   ungroup()
 

### 2016-06-07 ProcessingCode/get_mc_flanking.py
add file to get flanking mC values from database

@@ -1,0 +1,131 @@
+#! /usr/bin/env python
+
+from __future__ import division
+import MySQLdb
+import numpy as np
+from argparse import ArgumentParser
+
+
+parser = ArgumentParser(description='gather mC data from mySQL database between given intervals')
+parser.add_argument('-a', '--host', help='host', required=True)
+parser.add_argument('-u', '--user', help='mySQL user', required=True)
+parser.add_argument('-p', '--password', help='mySQL password', required=True)
+parser.add_argument('-d', '--database', help='mySQL database name', required=True)
+parser.add_argument('-i', '--infile', help='input coordinate file', required=True)
+parser.add_argument('-o', '--outfile', help='output file', required=True)
+parser.add_argument('-c', '--context', help='mC context', required=False, default=False)
+parser.add_argument('-b', '--binsize', help='Bin size', required=False, default=200, type=int)
+parser.add_argument('-n', '--numberbins', help='Number of bins', required=False, default=20, type=int)
+parser.add_argument('-m', '--all_mc_tables', help='all mC table names', required=True)
+options = parser.parse_args()
+
+
+def get_data(chrom, start, stop, options, cursor, tables):
+    """
+    fetch mC data from table in database in upstream and downstream regions from deletion sites
+    input set of deletion coordinates with list of accessions containing deletion
+    sum methylation in bins upstream and downstream of deletion coordinates
+    return aggregate data for accessions containing and not containing TE deletion
+    """
+    data = []
+    window = options.binsize * options.numberbins  # default 4 kb
+    upstream = int(start - (window/2))
+    x = 0
+    levels = np.zeros(options.numberbins)
+    for table in tables:
+        x += 1
+        levels += query_region(upstream, stop, options, table, chrom, cursor)
+    levels = levels / x
+    return levels.tolist()
+
+
+def query_region(upstream, stop, options, table, chrom, cursor):
+    data = np.zeros(options.numberbins)
+    # need to do two parts, upstream and downstream
+    for x in xrange(options.numberbins):
+        if x < options.numberbins/2:
+            # upstream region
+            mc_start = upstream + (x*options.binsize)
+            mc_end = upstream + ((x+1)* options.binsize)
+        else:
+            # downstream region
+            a = int(x-(options.numberbins/2))
+            mc_start = stop + (a*options.binsize)
[…]

### 2016-06-07 ProcessingCode/get_mc_flanking.py
add gzipped input option

@@ -4,6 +4,7 @@ from __future__ import division
 import MySQLdb
 import numpy as np
 from argparse import ArgumentParser
+import gzip
 
 
 parser = ArgumentParser(description='gather mC data from mySQL database between given intervals')
@@ -104,7 +105,11 @@ def group_tables(pos_accessions, neg_accessions, options):
 
 def process_all(options):
     cursor, link = setup_sql(options)
-    with open(options.infile, 'r') as insertions, open(options.outfile, 'w+') as outf:
+    if options.infile.endswith(".gz"):
+        insertions = gzip.open(options.infile, "rb")
+    else:
+        insertions = open(options.infile, "r")
+    with open(options.outfile, 'w+') as outf:
         ins_bins = ['ins'+str(x) for x in range(int(options.numberbins))]
         no_ins_bins = ['no_ins'+str(x) for x in range(int(options.numberbins))]
         outf.write('coords'+"\t"+"\t".join(ins_bins)+"\t"+"\t".join(no_ins_bins)+"\n")  # header
@@ -115,8 +120,8 @@ def process_all(options):
             start = int(line[1])
             stop = int(line[2])
             coords = 'chr'+chrom+','+str(start)+','+str(stop)
-            pos_accessions = line[-2].replace('-', '_').split(',')
-            neg_accessions = line[-1].replace('-', '_').split(',')
+            pos_accessions = line[4].replace('-', '_').split(',')
+            neg_accessions = line[5].replace('-', '_').split(',')
             pos_tables, neg_tables = group_tables(pos_accessions, neg_accessions, options)
             if ((start-window) <= 0) or len(pos_tables) == 0 or len(neg_tables) == 0 or chrom == "Mt" or chrom == "Pt":
                 pass
@@ -127,5 +132,6 @@ def process_all(options):
                 outf.write(coords+"\t"+"\t".join(map(str,all_data))+"\n")
     cursor.close()
     link.close()
+    insertions.close()
 
 process_all(options)

### 2016-06-07 ProcessingCode/get_mc_flanking.py
Allow header in input file

@@ -115,21 +115,24 @@ def process_all(options):
         outf.write('coords'+"\t"+"\t".join(ins_bins)+"\t"+"\t".join(no_ins_bins)+"\n")  # header
         window = options.binsize * options.numberbins
         for line in insertions:
-            line = line.rsplit()
-            chrom = line[0].strip('chr')
-            start = int(line[1])
-            stop = int(line[2])
-            coords = 'chr'+chrom+','+str(start)+','+str(stop)
-            pos_accessions = line[4].replace('-', '_').split(',')
-            neg_accessions = line[5].replace('-', '_').split(',')
-            pos_tables, neg_tables = group_tables(pos_accessions, neg_accessions, options)
-            if ((start-window) <= 0) or len(pos_tables) == 0 or len(neg_tables) == 0 or chrom == "Mt" or chrom == "Pt":
-                pass
+            if line[0] == "chromosome":
+                pass # header
             else:
-                mc_values_accessions = get_data(chrom, start, stop, options, cursor, pos_tables)
-                mc_values_non_accessions = get_data(chrom, start, stop, options, cursor, neg_tables)
-                all_data = mc_values_accessions + mc_values_non_accessions
-                outf.write(coords+"\t"+"\t".join(map(str,all_data))+"\n")
+                line = line.rsplit()
+                chrom = line[0].strip('chr')
+                start = int(line[1])
+                stop = int(line[2])
+                coords = 'chr'+chrom+','+str(start)+','+str(stop)
+                pos_accessions = line[4].replace('-', '_').split(',')
+                neg_accessions = line[5].replace('-', '_').split(',')
+                pos_tables, neg_tables = group_tables(pos_accessions, neg_accessions, options)
+                if ((start-window) <= 0) or len(pos_tables) == 0 or len(neg_tables) == 0 or chrom == "Mt" or chrom == "Pt":
+                    pass
+                else:
+                    mc_values_accessions = get_data(chrom, start, stop, options, cursor, pos_tables)
+                    mc_values_non_accessions = get_data(chrom, start, stop, options, cursor, neg_tables)
+                    all_data = mc_values_accessions + mc_values_non_accessions
+                    outf.write(coords+"\t"+"\t".join(map(str,all_data))+"\n")
     cursor.close()
     link.close()
     insertions.close()

### 2016-06-07 ProcessingCode/get_mc_flanking.py
bug fix

@@ -115,10 +115,10 @@ def process_all(options):
         outf.write('coords'+"\t"+"\t".join(ins_bins)+"\t"+"\t".join(no_ins_bins)+"\n")  # header
         window = options.binsize * options.numberbins
         for line in insertions:
+            line = line.rsplit()
             if line[0] == "chromosome":
                 pass # header
             else:
-                line = line.rsplit()
                 chrom = line[0].strip('chr')
                 start = int(line[1])
                 stop = int(line[2])

### 2016-06-07 ProcessingCode/split_true_deletions.sh
Add TE absence classifications

@@ -1,0 +1,13 @@
+# This script will split TE variants into two categories
+# 1. TE absences that are due to the lack of an insertion in the ancestral genome
+# 2. TE absences that are due to a true deletion of the ancestral TE
+# Some TE variants will have an NA call, where the classification is ambiguous
+# these will not be in either file
+
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz \
+| sed 1d \
+| awk 'BEGIN {FS=OFS="\t"} {if ($10 == "True deletion") print $0}' -> ../ProcessedData/TEPAV_true_deletions.tsv
+
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz \
+| sed 1d \
+| awk 'BEGIN {FS=OFS="\t"} {if ($10 == "No insertion") print $0}' - > ../ProcessedData/TEPAV_no_insertion.tsv

### 2016-06-07 ProcessingCode/split_true_deletions.sh
gzip output

@@ -6,8 +6,12 @@
 
 gunzip -c ../RawData/TEPID_TEPAV.tsv.gz \
 | sed 1d \
-| awk 'BEGIN {FS=OFS="\t"} {if ($10 == "True deletion") print $0}' -> ../ProcessedData/TEPAV_true_deletions.tsv
+| awk 'BEGIN {FS=OFS="\t"} {if ($10 == "True deletion") print $0}' - \
+| gzip - \
+> ../ProcessedData/TEPAV_true_deletions.tsv.gz
 
 gunzip -c ../RawData/TEPID_TEPAV.tsv.gz \
 | sed 1d \
-| awk 'BEGIN {FS=OFS="\t"} {if ($10 == "No insertion") print $0}' - > ../ProcessedData/TEPAV_no_insertion.tsv
+| awk 'BEGIN {FS=OFS="\t"} {if ($10 == "No insertion") print $0}' - \
+| gzip - \
+> ../ProcessedData/TEPAV_no_insertion.tsv.gz

### 2016-06-07 AnalysisCode/closest_DMRs.sh
gzip output

@@ -1,7 +1,0 @@
-gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
-| bedtools closest -b - -a ../RawData/c_dmrs.tsv.gz -d -wb \
-> ../ProcessedData/TE_C_DMR_distances.bed
-
-gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
-| bedtools closest -b - -a ../RawData/cg_dmrs.tsv.gz -d -wb \
-> ../ProcessedData/TE_CG_DMR_distances.bed

### 2016-06-07 ProcessingCode/closest_DMRs.sh

@@ -1,0 +1,9 @@
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
+| bedtools closest -b - -a ../RawData/c_dmrs.tsv.gz -d -wb \
+| gzip - \
+> ../ProcessedData/TE_C_DMR_distances.bed.gz
+
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
+| bedtools closest -b - -a ../RawData/cg_dmrs.tsv.gz -d -wb \
+| gzip - \
+> ../ProcessedData/TE_CG_DMR_distances.bed.gz

### 2016-06-07 ProcessingCode/run_flanking_mc.sh
Add script to get flanking mC

@@ -1,0 +1,155 @@
+# No insertions
+# allC
+python get_mc_flanking.py \
+-a 202.8.39.33 \
+-p cr@ckl3\#3 \
+-u mako \
+-d population_epigenetics \
+-i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
+-o ../ProcessedData/flanking_allC_no_insertion.tsv \
+-b 100 \
+-n 20 \
+-m ../RawData/all_mc_tables.txt
+
+# CG
+python get_mc_flanking.py \
+-a 202.8.39.33 \
+-p cr@ckl3\#3 \
+-u mako \
+-d population_epigenetics \
+-i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
+-o ../ProcessedData/flanking_CG_no_insertion.tsv \
+-b 100 \
+-n 20 \
+-c CG \
+-m ../RawData/all_mc_tables.txt
+
+# CHG
+python get_mc_flanking.py \
+-a 202.8.39.33 \
+-p cr@ckl3\#3 \
+-u mako \
+-d population_epigenetics \
+-i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
+-o ../ProcessedData/flanking_CHG_no_insertion.tsv \
+-b 100 \
+-n 20 \
+-c CHG \
+-m ../RawData/all_mc_tables.txt
+
+# CHH
+python get_mc_flanking.py \
+-a 202.8.39.33 \
+-p cr@ckl3\#3 \
+-u mako \
+-d population_epigenetics \
+-i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
+-o ../ProcessedData/flanking_CHH_no_insertion.tsv \
+-b 100 \
+-n 20 \
+-c CHH \
+-m ../RawData/all_mc_tables.txt
+
+# True deletions
+# allC
+python get_mc_flanking.py \
+-a 202.8.39.33 \
+-p cr@ckl3\#3 \
+-u mako \
+-d population_epigenetics \
+-i ../ProcessedData/TEPAV_true_deletion.tsv.gz \
+-o ../ProcessedData/flanking_allC_true_deletion.tsv \
+-b 100 \
+-n 20 \
+-m ../RawData/all_mc_tables.txt
+
+# CG
+python get_mc_flanking.py \
+-a 202.8.39.33 \
+-p cr@ckl3\#3 \
+-u mako \
[…]

### 2016-06-07 ProcessingCode/run_flanking_mc.sh
remove mysql info

@@ -1,10 +1,10 @@
 # No insertions
 # allC
 python get_mc_flanking.py \
--a 202.8.39.33 \
--p cr@ckl3\#3 \
--u mako \
--d population_epigenetics \
+-a [host] \
+-p [password] \
+-u [user] \
+-d [database] \
 -i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
 -o ../ProcessedData/flanking_allC_no_insertion.tsv \
 -b 100 \
@@ -13,10 +13,10 @@ python get_mc_flanking.py \
 
 # CG
 python get_mc_flanking.py \
--a 202.8.39.33 \
--p cr@ckl3\#3 \
--u mako \
--d population_epigenetics \
+-a [host] \
+-p [password] \
+-u [user] \
+-d [database] \
 -i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
 -o ../ProcessedData/flanking_CG_no_insertion.tsv \
 -b 100 \
@@ -26,10 +26,10 @@ python get_mc_flanking.py \
 
 # CHG
 python get_mc_flanking.py \
--a 202.8.39.33 \
--p cr@ckl3\#3 \
--u mako \
--d population_epigenetics \
+-a [host] \
+-p [password] \
+-u [user] \
+-d [database] \
 -i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
 -o ../ProcessedData/flanking_CHG_no_insertion.tsv \
 -b 100 \
@@ -39,10 +39,10 @@ python get_mc_flanking.py \
 
 # CHH
 python get_mc_flanking.py \
--a 202.8.39.33 \
--p cr@ckl3\#3 \
--u mako \
--d population_epigenetics \
+-a [host] \
+-p [password] \
+-u [user] \
+-d [database] \
 -i ../ProcessedData/TEPAV_no_insertion.tsv.gz \
 -o ../ProcessedData/flanking_CHH_no_insertion.tsv \
 -b 100 \
@@ -53,10 +53,10 @@ python get_mc_flanking.py \
 # True deletions
 # allC
 python get_mc_flanking.py \
--a 202.8.39.33 \
--p cr@ckl3\#3 \
--u mako \
--d population_epigenetics \
+-a [host] \
+-p [password] \
+-u [user] \
+-d [database] \
 -i ../ProcessedData/TEPAV_true_deletion.tsv.gz \
[…]

### 2016-06-07 DataAquisition/get_DHS_data.sh
add DHS data

@@ -1,0 +1,1 @@
+wget ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM1289nnn/GSM1289359/suppl/GSM1289359%5FCol%2D0%2E7d%5FSeedling%2ENA%2ECONTROL%2EDS21094%2Epeaks%2Ebed%2Etxt%2Egz

### 2016-06-07 ProcessingCode/make_genomic_feature_files.sh
add genomic regions

@@ -1,0 +1,27 @@
+# This file will create coordinate files for gene upstream and downstream regions
+# and gene introns, and place the output in ProcessedData
+
+# Upstream
+gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
+| awk 'BEGIN {FS=OFS="\t"} {if ($3 == "gene") print $0}' - \
+| awk 'BEGIN {FS=OFS="\t"} {$3 = "upstream"; $5 = $4; $4 = $4 - 2000; print $0}' - \
+| gzip - \
+> ../ProcessedData/gene_upstream_regions.gff.gz
+
+# Downstream
+gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
+| awk 'BEGIN {FS=OFS="\t"} {if ($3 == "gene") print $0}' - \
+| awk 'BEGIN {FS=OFS="\t"} {$3 = "downstream"; $4 = $5; $5 = $5 + 2000; print $0}' - \
+| gzip - \
+> ../ProcessedData/gene_downstream_regions.gff.gz
+
+# Introns
+wget ftp://ftp.ensemblgenomes.org//pub/release-31/plants/gff3/arabidopsis_thaliana/Arabidopsis_thaliana.TAIR10.31.gff3.gz
+
+gt gff3 -addintrons Arabidopsis_thaliana.TAIR10.31.gff3.gz \
+| grep "intron" - \
+| awk 'BEGIN {FS=OFS="\t"} {$1 = "chr"$1; print $0}' - \
+| bedtools intersect -a - -b ../RawData/TAIR10_GFF3_genes.gff.gz -wb \
+| awk 'BEGIN {FS=OFS="\t"} {if ($12 == "gene") print $10, $11, $3, $4, $5, $15, $16, $17, $18}' - \
+| gzip - \
+> ../ProcessedData/introns.gff.gz

### 2016-06-07 ProcessingCode/add_feature_between.py
add gene features

@@ -1,0 +1,51 @@
+#! /usr/bin/env python
+
+""" Example GFF
+chr1    TAIR9   exon    3631    3913    .   +   .   Parent=AT1G01010.1
+chr1    TAIR9   exon    3996    4276    .   +   .   Parent=AT1G01010.1
+chr1    TAIR9   exon    4486    4605    .   +   .   Parent=AT1G01010.1
+chr1    TAIR9   exon    4706    5095    .   +   .   Parent=AT1G01010.1
+chr1    TAIR9   exon    5174    5326    .   +   .   Parent=AT1G01010.1
+chr1    TAIR9   exon    5439    5899    .   +   .   Parent=AT1G01010.1
+chr1    TAIR9   exon    8571    8737    .   -   .   Parent=AT1G01020.1
+chr1    TAIR9   exon    8417    8464    .   -   .   Parent=AT1G01020.1
+chr1    TAIR9   exon    8236    8325    .   -   .   Parent=AT1G01020.1
+chr1    TAIR9   exon    7942    7987    .   -   .   Parent=AT1G01020.1
+"""
+
+"""
+Usage:
+python add_feature_between.py infile.gff > output.gff
+Read gff file and coordinates of intervening sequences and label with given feature name
+Input must be position sorted
+sort -k1,1 -k4,4n file > out
+"""
+
+import pandas as pd
+import sys
+
+
+def add_introns(f):
+    with open(f, 'r') as f:
+        table = pd.read_csv(f, sep='\t', header=-1)
+    l = len(table)
+    for x in range(1,l):
+        if table.iloc[x][8] == table.iloc[x-1][8]:
+            start = table.iloc[x-1][4]
+            end = table.iloc[x][3]
+            strand = table.iloc[x][6]
+            chrom = table.iloc[x][0]
+            data = [
+                chrom,
+                table.iloc[x][1],
+                'intron',
+                start,
+                end,
+                table.iloc[x][5],
+                table.iloc[x][6],
+                table.iloc[x][7],
+                table.iloc[x][8]]
+            print("\t".join(map(str,data)))
+
+
+add_introns(sys.argv[1])

### 2016-06-07 ProcessingCode/make_genomic_feature_files.sh

@@ -6,22 +6,41 @@ gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
 | awk 'BEGIN {FS=OFS="\t"} {if ($3 == "gene") print $0}' - \
 | awk 'BEGIN {FS=OFS="\t"} {$3 = "upstream"; $5 = $4; $4 = $4 - 2000; print $0}' - \
 | gzip - \
-> ../ProcessedData/gene_upstream_regions.gff.gz
+> ../ProcessedData/GeneFeatures/gene_upstream_regions.gff.gz
 
 # Downstream
 gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
 | awk 'BEGIN {FS=OFS="\t"} {if ($3 == "gene") print $0}' - \
 | awk 'BEGIN {FS=OFS="\t"} {$3 = "downstream"; $4 = $5; $5 = $5 + 2000; print $0}' - \
 | gzip - \
-> ../ProcessedData/gene_downstream_regions.gff.gz
+> ../ProcessedData/GeneFeatures/gene_downstream_regions.gff.gz
+
+# Exons
+gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
+| awk '{if ($3 == "exon") print $0}' - \
+> ../ProcessedData/GeneFeatures/exons.gff
 
 # Introns
-wget ftp://ftp.ensemblgenomes.org//pub/release-31/plants/gff3/arabidopsis_thaliana/Arabidopsis_thaliana.TAIR10.31.gff3.gz
-
-gt gff3 -addintrons Arabidopsis_thaliana.TAIR10.31.gff3.gz \
-| grep "intron" - \
-| awk 'BEGIN {FS=OFS="\t"} {$1 = "chr"$1; print $0}' - \
-| bedtools intersect -a - -b ../RawData/TAIR10_GFF3_genes.gff.gz -wb \
-| awk 'BEGIN {FS=OFS="\t"} {if ($12 == "gene") print $10, $11, $3, $4, $5, $15, $16, $17, $18}' - \
+python add_feature_between.py ../ProcessedData/exons.gff \
 | gzip - \
-> ../ProcessedData/introns.gff.gz
+> ../ProcessedData/GeneFeatures/introns.gff.gz
+
+# 5' UTR
+gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
[…]

### 2016-06-07 ProcessingCode/gene_feature_intersections.sh
Add scripts and data for doing gene feature intersections

@@ -1,0 +1,19 @@
+# This script will intersect the TE variants with all genomic features, including
+# 2 kb upstream region
+# 5' UTR
+# Exon
+# Intron
+# 3' UTR
+# 2 kb downstream region
+# Pseudogenes
+# Other Col-0 TEs
+# DNase I hypersensitivity sites
+
+gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d > tepid_tepav.tsv
+
+# need to remove redundancy in gene models first
+gunzip -c ../ProcessedData/GeneFeatures/exons.gff.gz \
+| bedtools intersect -a - -b tepid_tepav.tsv -wb \
+> ../ProcessedData/GeneFeatures/exon_intersections.gff
+
+rm tepid_tepav.tsv

### 2016-06-07 ProcessingCode/make_genomic_feature_files.sh

@@ -18,10 +18,11 @@ gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
 # Exons
 gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
 | awk '{if ($3 == "exon") print $0}' - \
+| sort -k1,1 -k4,4n - \
 > ../ProcessedData/GeneFeatures/exons.gff
 
 # Introns
-python add_feature_between.py ../ProcessedData/exons.gff \
+python add_feature_between.py ../ProcessedData/GeneFeatures/exons.gff \
 | gzip - \
 > ../ProcessedData/GeneFeatures/introns.gff.gz
 
@@ -43,4 +44,8 @@ gunzip -c ../RawData/TAIR10_GFF3_genes.gff.gz \
 | gzip - \
 > ../ProcessedData/GeneFeatures/pseudogene.gff.gz
 
-gzip ../ProcessedData/GeneFeatures/exons.gff
+gff2bed < ../ProcessedData/GeneFeatures/exons.gff \
+| bedtools merge -c 10 -o distinct -i - \
+| gzip - \
+> ../ProcessedData/GeneFeatures/exons.bed.gz
+rm ../ProcessedData/GeneFeatures/exons.gff

### 2016-06-09 ProcessingCode/classify_TE_absence.R
Add TEPAV frequency classifications

@@ -1,6 +1,6 @@
 library(dplyr)
 
-poly <- read_tsv("../ProcessedData/poly_te_genotyped.tsv", col_names = F)
+poly <- read_tsv("../ProcessedData/poly_te_genotyped.tsv.gz", col_names = F)
 
 polarized <- poly %>%
   rowwise() %>%
@@ -14,9 +14,11 @@ polarized <- poly %>%
          } else {
            "NA"
          },
-         MAF = min(TE_present, TE_absent) / 216) %>%
+         MAF = min(TE_present, TE_absent) / 216,
+         Frequency_classification = ifelse(MAF < 0.03, "Rare", "Common")) %>%
   ungroup()
 
-colnames(polarized) <- c("chromosome", "start", "end", "TE", "Accessions_TE_present", "Accessions_TE_absent", "Count_TE_present", "Count_TE_absent", "Minor_allele", "Absence_classification", "MAF")
+colnames(polarized) <- c("chromosome", "start", "end", "TE", "Accessions_TE_present", "Accessions_TE_absent", "Count_TE_present", "Count_TE_absent", "Minor_allele", "Absence_classification", "MAF", "Frequency_classification")
 write.table(polarized, file = "../RawData/tepav_regenerated.tsv",
             quote = F, sep = "\t", na = "NA", row.names = F, col.names = T)
+system("gzip ../RawData/tepav_regenerated.tsv")

### 2016-06-09 AnalysisCode/TE_insertion_distribution.R
Added R code and plots
Code to generate plots of chromosomal distribution of
TE variants, length distribution of TE variants,
minor allele frequency

@@ -1,0 +1,108 @@
+library(readr)
+library(dplyr)
+library(ggplot2)
+library(RColorBrewer)
+
+# TEPID TE calls
+tepav <- read_tsv("../RawData/TEPID_TEPAV.tsv.gz", col_names = T)
+
+# 1. Plot minor allele frequency
+pdf("../Plots/MAF.pdf", height = 4, width = 4)
+hist(tepav$MAF*100, breaks = 100, col = "grey", xlab = "MAF", main = "Minor allele frequency distribution")
+dev.off()
+
+# 2. Plot TE variant length distribution
+all_te <- read_tsv("../RawData/TAIR9_TE.bed.gz", col_names = F) %>%
+  mutate(l = X3-X2)
+
+tepav_tes <- unlist(strsplit(tepav$TE, ","))
+lengths <- all_te %>%
+  mutate(pav = X5 %in% tepav_tes) %>%
+  filter(pav == T)
+
+pdf(file = "../Plots/length_distribution_histogram.pdf", width = 4, height = 4)
+hist(all_te$l, breaks = 100, col = "grey", main = "Length distribution of all Col-0 TEs", xlab = "Length (bp)")
+rug(all_te$l)
+
+hist(lengths$l, breaks = 100, col = "grey", main = "Length distribution of population-variable TEs", xlab = "Length (bp")
+rug(lengths$l)
+dev.off()
+
+# plot logs
+pdf("../Plots/TE_length_log.pdf", width = 5, height = 5)
+plot(density(log(lengths$l)), col=rgb(1, 0, 0,0.5), main = "TE length distribution", xlab = "log10 TE length")
+polygon(density(log(lengths$l)), col=rgb(1, 0, 0,0.5))
+polygon(density(log(all_te$l)), col = rgb(0, 0, 1,0.5))
+legend(8,0.2,legend = c("All", "TEPAV"), lty=c(1,1), col=c("blue", "red"), bty = "n")
+dev.off()
+
+# 3. Plot genomic distribution of TE variants, for rare, common, TE deletions vs no insertions, and all together
+
+rare_vs_common_colors <- c("#FC9272", "#67000D")
+ins_vs_del_colors <- c("#9ECAE1", "#08306B")
+
+ggplot(tepav, aes(start, color = Frequency_classification)) +
+  geom_density(adjust = 1/8) + scale_color_manual(values = rare_vs_common_colors) +
[…]

### 2016-06-09 AnalysisCode/mC_heatmap.R
update plots

@@ -2,17 +2,18 @@ library(dplyr)
 library(gplots)
 library(RColorBrewer)
 library(readr)
+library(ggplot2)
 
-mc_ins_2kb <- read_tsv("../ProcessedData/allC_ins_2kb.tsv", col_names = F)
-mc_del_2kb <- read_tsv("../ProcessedData/allC_true_del_2kb.tsv", col_names = F)
-
-
-### old code ###
-### re-write for updated analysis ###
+mc_no_insertion_2kb <- read_tsv("../ProcessedData/DNAmethylation/flanking_allC_no_insertion.tsv", col_names = T)
+mc_true_deletion_2kb <- read_tsv("../ProcessedData/DNAmethylation/flanking_allC_true_deletions.tsv", col_names = T)
 
 # sum rows for sort order
-mc_all_ins_2kb_dist.sums <- mutate(mc_all_ins_2kb_dist, sums=rowSums(mc_all_ins_2kb_dist[,2:41]))
+mc_no_insertion_2kb.sums <- mutate(mc_no_insertion_2kb, sums=rowSums(mc_no_insertion_2kb[,2:41]))
+mc_true_deletion_2kb.sums <- mutate(mc_true_deletion_2kb, sums=rowSums(mc_true_deletion_2kb[,2:41]))
 
+# Sort by total mC level
+mc_no_insertion_2kb.sort <- arrange(mc_no_insertion_2kb.sums, desc(sums))
+mc_true_deletion_2kb.sort <- arrange(mc_true_deletion_2kb.sums, desc(sums))
 
 # categorical heatmap
 cat_hmap <- function(d) {
@@ -30,27 +31,19 @@ cat_hmap <- function(d) {
             cex.lab=0.7)
 }
 
-# heatmap
-hmap <- function(d) {
-  heatmap.2(as.matrix(d),Colv=NA,col=brewer.pal(9, "Reds"),
-            labRow = "",
-            labCol = "",
-            dendrogram = "none",
-            margins = c(5,2),
-            trace = "none",
-            par(cex.main=0.7,
-                cex.lab=0.7,
-                cex.axis=0.7
-            ),
-            cex.lab=0.7)
+
+scale_max <- function(m, l){
+  m.floor <- m > l
+  m[m.floor] <- l
+  return(t(data.matrix(m)))
 }
 
-# scale values to max 0.5
-eu.floor <- eu > 0.5
-eu[eu.floor] <- 0.5
[…]

### 2016-08-08 ProcessingCode/long_range_ld.py
Add script to check for long-range LD between TE variants

@@ -1,0 +1,87 @@
+#! /usr/bin/env python
+
+from __future__ import division
+from argparse import ArgumentParser
+import gzip
+
+
+parser = ArgumentParser(description="Check for long-range LD in set of TE variants")
+parser.add_argument('-t', '--transposons', help='file containing the TE variants', required=True)
+parser.add_argument('-c', '--cutoff', help='fraction of matching accession names to be called similar', required=True)
+parser.add_argument('-o', '--outfile', help='name of output file', required=True)
+options = parser.parse_args()
+
+
+def read_file(options):
+    """ read TE variants into list """
+    if options.transposons.endswith(".gz"):
+        infile = gzip.open(options.transposons, "rb")
+    else:
+        infile = open(options.transposons, "r")
+    d = []
+    for line in infile:
+        line = line.rsplit()
+        if line[0] == "chromosome":
+            pass  # header
+        else:
+            pos_acc = line[4]
+            neg_acc = line[5]
+            ident = ",".join([line[0], line[1], line[2]])
+            d.append([pos_acc, neg_acc, ident])
+    infile.close()
+    return(d)
+
+
+def compare_lists(list1, list2):
+    """take two lists, return fraction of matches between the two lists"""
+    number_match = len(set(list1) & set(list2))
+    l1 = len(list1)
+    l2 = len(list2)
+    fraction = number_match / max(l1, l2)
+    return(fraction)
+
+
+def calc_distance(c1, c2):
+    """ Find the distance between coordinates if chromosome is the same
+        otherwise just return NA """
+    if c1[0] == c2[0]:  # chromosome is the same
+        d = abs(int(c1[1]) - int(c2[1]))
+    else:
+        d = "NA"
+    return(d)
+        
+
+def get_similar(data, cutoff):
+    """iterate over dictionary items, return items with similarity above cutoff"""
+    similar_variants = []  # holds coordinates of similar TE variants
+    # convert to only half the symmetrical matrix so we don't do 2x too many comparisons and get duplicate results
+    for i in xrange(0, len(data)):
+        for j in xrange(i, len(data)):
+            acc_pos_1, acc_neg_1, coords_1 = [x.split(",") for x in data[i]]
+            acc_pos_2, acc_neg_2, coords_2 = [x.split(",") for x in data[j]]
+            if (coords_1 == coords_2) or (len(acc_pos_1) < 5) or (len(acc_neg_1) < 5):  # same TEPAV or only one accession
+                pass
+            else:
[…]

### 2016-08-09 AnalysisCode/plot_trans_scatter.R
Modified graphs to show highest trans counts

@@ -48,12 +48,16 @@ pvals %>%
 
 ggsave("../Plots/pval_ins.png", height=5, width = 5, dpi = 600)
 
-dot_plot <- function(pval, chr, y) {
+dot_plot <- function(pval, chr, y, z) {
   filtered <- subset(pval, te_chr == chr)
   h <- hist(filtered$te_start, breaks = 10000, plot=F)
-  d <- data.frame(x = h$breaks[1:length(h$breaks) - 1], y = h$counts)
+  d <- data.frame(x = h$breaks[1:length(h$breaks) - 1], y = h$counts, z = NA)
   d <- d[d$y > 0,]
-  plot(d, pch=19, cex=0.5, ylim = c(0, y))
+  # add column with color
+  d[d$y > z, 3] <- 1
+  d[d$y <= z, 3] <- 2
+  plot(d$x, d$y, pch=19, cex=0.5, ylim = c(0, y), col=c("red", "grey")[d$z])
+  abline(h=z, col="grey")
 }
 
 ins <- filter(pvals, indel == "Insertion")
@@ -61,18 +65,18 @@ del <- filter(pvals, indel == "Deletion")
 
 pdf("../Plots/trans_dotplot_insertions.pdf", height=1.5, width=10, useDingbats = F)
 par(mfrow=c(1,5), mar = c(1,1,1,1))
-dot_plot(ins, "chr1", 400)
-dot_plot(ins, "chr2", 400)
-dot_plot(ins, "chr3", 400)
-dot_plot(ins, "chr4", 400)
-dot_plot(ins, "chr5", 400)
+dot_plot(ins, "chr1", 400, 200)
+dot_plot(ins, "chr2", 400, 200)
+dot_plot(ins, "chr3", 400, 200)
+dot_plot(ins, "chr4", 400, 200)
+dot_plot(ins, "chr5", 400, 200)
 dev.off()
 
 pdf("../Plots/trans_dotplot_deletions.pdf", height=1.5, width=10, useDingbats = F)
 par(mfrow=c(1,5), mar = c(1,1,1,1))
-dot_plot(del, "chr1", 700)
-dot_plot(del, "chr2", 700)
-dot_plot(del, "chr3", 700)
-dot_plot(del, "chr4", 700)
-dot_plot(del, "chr5", 700)
+dot_plot(del, "chr1", 700, 200)
+dot_plot(del, "chr2", 700, 200)
[…]

### 2016-08-09 ProcessingCode/filter_sig_pvals_correlated_tes.sh
Add code and data for ghost trans DMR associations

@@ -1,0 +1,31 @@
+#! /bin/bash
+
+# make a temp folder to hold the intermediate output
+mkdir ./temp
+
+# filter where distance between TEs is more than 5 kb (this might be a low cutoff)
+awk '$7 > 5000' ../ProcessedData/long_range_ld_te.tsv > ./temp/long_range_ld_5kb.tsv
+
+# switch order of columns and cat, so we can intersect with the significant pval coordinates
+awk 'BEGIN {FS=OFS="\t"} {print $4,$5,$6,$1,$2,$3,$7}' ./temp/long_range_ld_5kb.tsv \
+    | cat - ./temp/long_range_ld_5kb.tsv \
+    > ./temp/both_combinations_long_range_ld.tsv
+
+# first need to change column order and add a second coordinate to the pvals file
+awk 'BEGIN {FS=OFS="\t"} {$6 = $4 + 10; print $3,$4,$6,$1,$2,$5}' ../ProcessedData/cdmr_sig_pvals.tsv > ./temp/reordered_pvals.tsv
+
+# intersect with the significant pvalues
+bedtools intersect -a ./temp/reordered_pvals.tsv -b ./temp/both_combinations_long_range_ld.tsv -wb > ./temp/intersections_pval.tsv
+
+# now add column with distance between DMR and TE, NA if different chromosome
+# filter where distance is less than 2 kb
+
+# columns are coords of DMR-correlated TE insertion, insertion/deletion classification
+# chr start of correlated DMR, coords of TE intersecting the inserted TE, coords of TE found
+# in same accessions as the correlated TE, distance between the two TEs
+# found in the same accessions (NA if chromosomes are different)
+
+# if chromosomes different insert NA, else calculate the distance between correlated DMR and TE, print those where distance is less than 2000
+awk 'BEGIN {FS=OFS="\t"} {if ($4 != $10) $14 = "NA"; else $14 = sqrt(($5 - $11)^2); if ($14 < 2000) print $0}' ./temp/intersections_pval.tsv \
+    > ../ProcessedData/ghost_trans_associations.tsv
+

### 2016-08-09 ProcessingCode/filter_sig_pvals_correlated_tes.sh
Remove temp data

@@ -29,3 +29,4 @@ bedtools intersect -a ./temp/reordered_pvals.tsv -b ./temp/both_combinations_lon
 awk 'BEGIN {FS=OFS="\t"} {if ($4 != $10) $14 = "NA"; else $14 = sqrt(($5 - $11)^2); if ($14 < 2000) print $0}' ./temp/intersections_pval.tsv \
     > ../ProcessedData/ghost_trans_associations.tsv
 
+rm -r ./temp

### 2016-08-10 AnalysisCode/plot_trans_scatter.R
Add heatmaps showing TE density

@@ -80,3 +80,23 @@ dot_plot(del, "chr3", 700, 200)
 dot_plot(del, "chr4", 700, 200)
 dot_plot(del, "chr5", 700, 200)
 dev.off()
+
+# draw heatmaps of TE density to go along the edges of the plots
+te <- read_tsv("../RawData/TAIR9_TE.bed.gz", col_names = c("chr", "start", "stop", "strand", "TE", "Family", "Superfamily"))
+
+color <- colorRampPalette(brewer.pal(9, "YlGnBu"))(100)
+
+plot_heatmap <- function(d, chr, color) {
+  c <- d[d$chr == chr,]
+  h <- hist(c$start, breaks = 200, plot = F)
+  image(as.matrix(h$counts), col = color)
+}
+
+pdf("../Plots/te_density_heatmaps.pdf", height = 10, width = 5)
+par(mfrow = c(5,1))
+plot_heatmap(te, "chr1", color)
+plot_heatmap(te, "chr2", color)
+plot_heatmap(te, "chr3", color)
+plot_heatmap(te, "chr4", color)
+plot_heatmap(te, "chr5", color)
+dev.off()

### 2016-08-11 Paper/paper.tex
Add paper

\documentclass[12pt]{article}
\usepackage[labelfont=bf]{caption}
\usepackage{lmodern}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{newfloat}
\usepackage[flushleft]{threeparttable}
\usepackage{parskip}
\usepackage[
  backend=bibtex,
  style=numeric,
  sorting=none,
  doi=true,
  url=false,
  citestyle=numeric-comp
]{biblatex}

\AtEveryBibitem{%
  \clearlist{language}%
  }
\setlength\bibitemsep{\baselineskip}

\bibliography{references.bib}
\usepackage{doi}
\usepackage{authblk}

\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
\usepackage{amssymb,amsmath}
\usepackage[margin=2cm]{geometry}
\graphicspath{ {../Plots/FinalFigures/} }
\usepackage{lineno}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true,
            colorlinks=true,
            citecolor=blue
}
\urlstyle{same}  % don't use monospace font for urls
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\date{}

\begin{document}

\title{Population scale mapping of transposable element diversity reveals links to gene regulation and epigenomic variation}

\author[1]{Tim Stuart}
\author[2]{Steven R. Eichten}
\author[1]{Jonathan Cahn}
\author[1]{Yuliya Karpievitch}
\author[2]{Justin Borevitz}
\author[1]{Ryan Lister}
\affil[1]{ARC Centre of Excellence in Plant Energy Biology, The University of
Western Australia, Perth, Australia}
\affil[2]{ARC Centre of Excellence in Plant Energy Biology, The Australian
National University, Canberra, Australia}

\renewcommand\Authands{ and }

\maketitle

Corresponding author: Ryan Lister \href{mailto:ryan.lister@uwa.edu.au}{ryan.lister@uwa.edu.au} \\

\textbf{Author ORCID IDs:}

0000-0002-3044-0897 (TS)

0000-0003-2268-395X (SRE)

0000-0002-5006-741X (JC)

0000-0001-6637-7239 (RL)

\linenumbers

\section{Abstract}

Variation in the presence or absence of transposable elements (TEs) is a
major source of genetic variation between individuals. Here, we
identified 23,095 TE presence/absence variants between 216 wild
Arabidopsis accessions. Most variants are rare, and we find a burden of
rare variants associated with extremes of both gene expression and DNA
methylation levels within the population. Of the common alleles
identified, two thirds were not in linkage disequilibrium with nearby
SNPs, implicating these TE variants as a source of novel genetic
diversity. Nearly 200 common TE variants were associated with
significantly altered expression of nearby genes, with potential
functional consequences including decreased pathogen resistance. A major
fraction of inter-accession DNA methylation differences were associated
with nearby TE insertion variants, with DNA, indicating an important
role of TE variants in facilitating epigenomic variation.

\section{Introduction}

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and make up a large fraction of most
eukaryotic genomes. The two classes of TEs are retrotransposons (class
I elements), which transpose via an RNA intermediate requiring a
reverse transcription reaction, and DNA transposons (class II
elements), which transpose via either a cut-paste or, in the case of
Helitrons, a rolling circle mechanism with no RNA intermediate
\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
insertion may disrupt essential regions of the genome. Consequently,
safeguards have evolved in order to suppress this activity. These
safeguards include epigenetic transcriptional silencing mechanisms,
chiefly involving the methylation of cytosine nucleotides (DNA
methylation) to produce 5-methylcytosine (mC), a mark that can signal
transcriptional silencing of the methylated locus. In
\emph{Arabidopsis thaliana }(Arabidopsis), DNA methylation occurs in
three DNA sequence contexts: mCG, mCHG, and mCHH, where H is any base
but G. Establishment of DNA methylation marks can be carried out by
two distinct pathways -- the RNA-directed DNA methylation pathway
guided by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2
pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
methylation in Arabidopsis is in the transcriptional silencing of TEs.
Loss of DNA methylation due to mutations in genes essential for DNA
methylation establishment and maintenance leads to expression of
previously silent TEs, and in some cases transposition
\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm,
  Jeddeloh:1999gu, Zemach:2013dj}.

TEs are thought to play an important role in evolution, not only because
of the disruptive potential of their transposition. The release of
transcriptional and post-transcriptional silencing of TEs can lead to
bursts of TE activity, quickly generating new genetic diversity
\cite{Vitte:2014he}. TEs may carry regulatory information such as
promoters and transcription factor binding sites, and their mobilization
may lead to the creation or expansion of gene regulatory networks
\cite{Henaff:2014hg, Bolger:2014bn, Ito:2011dga, Makarevitch:2015ho}.
Furthermore, the transposase enzymes required and encoded by TEs have
frequently been domesticated and repurposed as endogenous proteins, such
as the \emph{DAYSLEEPER} gene in Arabidopsis, derived from a hAT
transposase enzyme \cite{Bundock:2005gp}. Clearly, the activity of TEs
can have widespread and unpredictable effects on the host genome.
However, the identification of TE presence/absence variants in genomes
has remained difficult to date. It is challenging to identify the
structural changes in the genome caused by TE mobilization using current
short-read sequencing technologies as these reads are typically mapped
to a reference genome, which has the effect of masking structural
changes that may be present. However, in terms of the number of base
pairs affected, a large fraction of genetic differences between
Arabidopsis accessions appear to be due to variation in TE content
\cite{Cao:2011cf, Quadrana:2016bi}. Therefore identification of TE
variants is essential in order to develop a more comprehensive
understanding of the genetic variation that exists between genomes, and
of the consequences of TE movement on genome and cellular function.

The tools developed previously for identification of novel TE insertion
sites have several limitations. They either require a library of active
TE sequences, cannot identify TE absence variants, or are not designed
with population studies in mind, or suffer from a high degree of
false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel algorithm,
TEPID (Transposable Element Polymorphism IDentification), which is
designed for population studies. We tested our algorithm using both
simulated and real Arabidopsis sequencing data, finding that TEPID is
able to accurately identify TE presence/absence variants with respect to
the Col-0 reference genome. We applied our TE analysis method to
existing genome resequencing data for 216 different wild Arabidopsis
accessions, and identified widespread TE variation amongst these
accessions \cite{Schmitz:2013iu}.

\section{Results}

\subsection{Computational identification of TE presence/absence variation}

We developed TEPID, an analysis pipeline capable of detecting TE
presence/absence variants from paired end DNA sequencing data. TEPID
integrates split and discordant read mapping information, read mapping
quality, sequencing breakpoints, as well as local variations in
sequencing coverage to identify novel TE presence/absence variants with
respect to a reference TE annotation (Figure \ref{fig1}; see methods). This
typically takes 5-10 minutes using Arabidopsis sequencing data at 20-40x
coverage (excluding the read mapping step). After TE variant discovery
has been performed, TEPID then includes a second refinement step
designed for population studies. This examines each region of the genome
where there was a TE insertion identified in any member of the group,
and checks for evidence of this insertion in all other members of the
group. In this way, TEPID leverages TE variant information for a group
of related samples to correct false negative calls within the group.
Testing of TEPID using simulated TE variants in the Arabidopsis genome
showed that it was able to reliably detect simulated TE variants at
sequencing coverage levels commonly used in genomics studies (Figure 1 -
figure supplement \ref{fig1s1}).

In order to further assess the sensitivity and specificity of TE variant
discovery using TEPID, we identified TE variants in the \emph{Landsberg}
\emph{erecta} (L\emph{er}) accession, and compared these with the
L\emph{er }genome assembly created using long PacBio sequencing reads
\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
using TEPID, enabling identification of 446 TE insertions (Figure 1 -
source data 1) and 758 TE absence variants (Figure 1 - source data 2)
with respect to the Col-0 reference TE annotation. Reads providing
evidence for these variants were then mapped to the L\emph{er} reference
genome that was generated by \emph{de novo} assembly using Pacific
Biosciences P5-C3 chemistry with a 20 kb insert library
\cite{Chin:2013iw}, using the same alignment parameters as was used to
map reads to the Col-0 reference genome. This resulted in 98.7\% of
reads being aligned concordantly to the L\emph{er} reference, whereas
100\% aligned discordantly or as split reads to the Col-0 reference
genome (Table \ref{table1}). To find whether reads mapped to homologous regions in
both the Col-0 and L\emph{er} reference genomes, we conducted a blast
search \cite{Camacho:2009fc} using the DNA sequence between read pair
mapping locations in the L\emph{er }genome against the Col-0 genome, and
found the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence variants
in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. We conclude that reads providing evidence for TE
variants map discordantly or as split reads when mapped to the Col-0
reference genome, but map concordantly to homologous regions of the
L\emph{er} \emph{de novo} assembled reference genome, indicating that
structural variation is present at the sites identified by TEPID, and
that this is resolved in the \emph{de novo} assembled genome.

To estimate the rate of false negative TE absence calls made using
TEPID, we compared our L\emph{er} TE absence calls to the set of TE
absences in L\emph{er} genome identified previously by aligning
full-length Col-0 TEs to the Ler reference using BLAT
\cite{Quadrana:2016bi}. We found that 89.6\% (173/193) of these TE
absences were also identified using TEPID, indicating a false negative
rate of around 10\% for TE absence calls. To determine the rate of false
negative TE insertion calls, we ran TEPID using 90 bp paired-end Col-0
reads (Col-0 control samples from \cite{Jiang:2014ih}), aligning reads
to the L\emph{er} PacBio assembly. As TEPID requires a high-quality TE
annotation to discover TE variants, which is not available for the
L\emph{er} assembly, we simply looked for discordant and split read
evidence at the known Col-0-specific TE insertion sites
\cite{Quadrana:2016bi}, and found evidence reaching the TEPID threshold
for a TE insertion call to be made at 89.6\% (173/193) of these sites,
indicating a false negative rate of 10\%. However, we note that this
estimate does not take into account the TEPID refinement step used on
large populations, and so the false negative rate for samples analyzed
in the population from Schmitz et al. (2013) is likely to be slightly
lower than this estimate, as each accession gained on average 4\% more
insertion calls following this refinement step (Figure 2 - figure
supplement \hyperref[fig2s1]{1}).

\subsection{Abundant TE positional variation among natural Arabidopsis
populations}

We used TEPID to analyze previously published 100 bp paired-end genome
resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions and 8,088 TE
absence variants relative to the Col-0 reference accession, totalling
23,095 unique TE variants. In most accessions we identified 300-500 TE
insertions (mean = 378) and 1,000-1,500 TE absence variants (mean =
1,279), the majority of which were shared by two or more accessions
(Figure 2 - figure supplement \hyperref[fig2s2]{2}). PCR validations were performed for a
random subset of 10 insertions and 10 absence variants in 14 accessions
(totalling 280 validations), and confirmed the high accuracy of TE
variant discovery using the TEPID package, with results similar to that
observed using simulated data and the L\emph{er} genome analysis (Figure 2 -
figure supplement \hyperref[fig2s3]{3}). The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement \hyperref[fig2s4]{4}A, B), indicating
that the sensitivity of TE absence calls is not limited by sequencing
depth, while TE insertion calls benefitted from high sequencing depth.
However, accessions with low coverage gained more TE insertion calls
during the TEPID refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
indicating that these false negatives were effectively reduced by
leveraging TE variant information for the whole population.

As TE insertion and TE absence calls represent an arbitrary comparison
to the Col-0 reference genome, we sought to remove these arbitrary
comparisons and classify each variant as a new TE insertion or true
deletion of an ancestral TE in the population. To do this, we examined
the minor allele frequency (MAF) of each variant in the population,
under the expectation that the minor allele is the derived allele. We
re-classified common TE absences relative to Col-0 as rare TE insertions
in Col-0, and common TE insertions relative to Col-0 as true TE
deletions in Col-0. Cases where the TE variant was at an high MAF
(\textgreater{}20\%) were assigned NA calls, as it could not be
determined if these were cases where the variant was most likely to be
due to a true TE deletion or due to a new TE insertion. While these
classifications are not definitive, as there will be rare cases where a
true TE deletion has spread through the population and becomes the
common allele, we will correctly classify most TE variants. Overall, we
found 72.3\% of the TE absence variants identified with respect to the
Col-0 reference genome were likely due to a true TE deletion in these
accessions, while 4.8\% were due to new insertions in Col-0 not shared
by others in the population (Table \ref{table2}). Overall, we identified 15,077 new
TE insertions, 5,856 true TE deletions, and 2,162 TE variants at a high
MAF that were unable to be classified as an insertion or deletion.

TE insertions and deletions were distributed throughout chromosome 1 in
a pattern that was similar to the distribution of all Col-0 TEs (Figure
\ref{fig2}A). TE deletions and common TE variants were found in similar
chromosomal regions, as deletion variants represent the rare loss of
common variants. TE deletions and common variants were more highly
enriched in the pericentromeric regions than rare variants or TE
insertions. The distribution of rare TE variants and TE insertions was
somewhat similar to that observed for regions of the genome previously
identified as being differentially methylated in all DNA methylation
contexts (mCG, mCHG, mCHH) between the wild accessions (population
C-DMRs), while population CG-DMRs, differentially methylated in the mCG
context, less frequently overlapped with all types of TE variants
identified \cite{Schmitz:2013iu}. Furthermore, TE variants were depleted
within genes and DNase I hypersensitivity sites \cite{Sullivan:2014ii},
while they were enriched in gene flanking regions and within other
annotated TEs or pseudogenes (Figure \ref{fig2}B). We found that TE deletions and
common TE variants were enriched within the set of TE variants found in
gene bodies (Figure \ref{fig2}C, D). We did not find any significant enrichment
of TE variants within the \emph{KNOT ENGAGED ELEMENT (KEE)} regions,
previously identified as regions that may act as a ``TE sink''
\cite{Grob:2014kg} (Figure 2 - figure supplement \hyperref[fig2s5]{5}). This may indicate
that these regions do not act as a ``TE sink'' as has been previously
proposed, or that the ``TE sink'' activity is restricted to very recent
insertions, as the insertions we analysed in this population were likely
older than those used in the study by Grob et al.

Among the identified TE variants, several TE superfamilies were over- or
under-represented compared to the number expected by chance given the
overall genomic frequency of different TE types (Figure \ref{fig2}E). In
particular, both TE insertions and deletions in the RC/Helitron
superfamily were less numerous than expected, with an 11.5\% depletion
of RC/Helitron elements in the set of TE variants. In contrast, TEs
belonging to the LTR/Gypsy superfamily were more frequently deleted than
expected, with a 17\% enrichment in the set of TE deletions. This was
unlikely to be due to a differing ability of our detection methods to
identify TE variants of different lengths, as the TE variants identified
had a similar distribution of lengths as all Arabidopsis TEs annotated
in the Col-0 reference genome (Figure 2 - figure supplement \hyperref[fig2s6]{6}). These
enrichments suggest that the RC/Helitron TEs have been relatively
dormant in recent evolutionary history, while the LTR/Gypsy, which are
highly enriched in the pericentromeric regions, are frequently lost from
the Arabidopsis genome. At the family level, we observed similar
patterns of TE variant enrichment or depletion (Figure 2 - figure
supplement \hyperref[fig2s7]{7}; source data 5).

We further examined Arabidopsis (Col-0) DNA sequencing data from a
transgenerational stress experiment to investigate the possible minimum
number of generations required for TE variants to arise
\cite{Jiang:2014ih}. We identified a single potential TE insertion in a
sample following 10 generations of single-seed descent under high
salinity stress conditions, and no TE variants in the control
single-seed descent set. However, without experimental validation it
remains unclear if this represents a true variant. Therefore, we
conclude that TE variants may arise at a rate less than 1 insertion in
30 generations under laboratory conditions. Further experimental work
will be required to precisely determine the rate of transposition in
Arabidopsis.

\subsection{Relationship between TE variants and single nucleotide
polymorphisms}

Although thousands of TE variants were identified, they may be linked
(i.e. `tagged') by the previously identified single nucleotide
polymorphisms (SNPs), or unlinked from SNPs across the accessions. We
tested how often common TE variants (\textgreater{}3 \% MAF; 7
accessions) were linked to adjacent SNPs to determine when they would
represent a previously unassessed source of genetic variation between
accessions. SNPs that were previously identified between the accessions
\cite{Schmitz:2013iu} were compared to the presence/absence of
individual TE variants. For the testable TE variants in the population,
the nearest 300 flanking SNPs upstream and downstream of the TE variant
site were analyzed for local linkage disequilibrium (LD, r\textsuperscript{2}; see
methods). TE variants were classified as being either `low', `mid', or
`high' LD variants by comparing ranked r\textsuperscript{2} values of TE variant to SNPs
against the median ranked r\textsuperscript{2} value for all between SNP comparisons
(SNP-SNP) to account for regional variation in the extent of SNP-SNP LD
(Figure \ref{fig3}A, B) due to recombination rate variation or selection
\cite{Horton:2012jh}. The majority (61\%) of testable TE variants had
low LD with nearby SNPs, and represent a source of genetic diversity not
previously assessed by current SNP-based genotype calling methods
(Figure \ref{fig3}C). 29\% of TE variants displayed high levels of LD and are
tagged by nearby SNPs, while only 10\% had intermediate levels of LD. We
observed a positive correlation between TE variant MAF and LD state,
with variants of a high minor allele frequency more often classified as
high-LD (Figure \ref{fig3}D). While the proportion of TE variants classified as
high, mid, or low-LD was mostly the same for both TE insertions and TE
deletions, TE variants with a high MAF (\textgreater{}20\%) that were
unable to be classified as either true deletions or as new insertions
had a much higher proportion of high-LD variants (Figure \ref{fig3}E). This was
consistent with the observation that the more common alleles were more
often in a high-LD state. TE variants displayed a similar distribution
over chromosome one regardless of linkage classification (Figure 3 -
figure supplement \hyperref[fig3s1]{1}). Overall, this analysis revealed an abundance of
previously unexplored genetic variation that exists amongst Arabidopsis
accessions caused by the presence or absence of TEs, and illustrates the
importance of identifying TE variants alongside other genetic diversity
such as SNPs.

\subsection{TE variants affect gene expression}

To determine whether the TE variants identified affected nearby gene
expression, we compared the steady state transcript abundance within
mature leaf tissue between accessions with and without TE insertions
or deletions, for genes with TE variants located in the 2 kb gene
upstream region, 5' UTR, exon, intron, 3' UTR or 2 kb downstream
region (Figure \ref{fig4}A). While the steady state transcript
abundance of most genes appeared to be unaffected by the presence of a
TE, 196 genes displayed significant differences in transcript
abundance linked with the presence of a TE variant, indicating a role
for these variants in the local regulation of gene expression (1\%
false discovery rate; greater than 2-fold change in transcript
abundance; Figure \ref{fig4}A, Figure 4 - source data 1). We did
not find any functional category enrichments in this set of
differentially expressed genes. As rare TE variants with a MAF less
than 3\% may also be associated with a difference in transcript
abundance, but were unable to be statistically tested due to their
rarity, we performed a burden test for enrichment of rare variants in
the extremes of expression \cite{Zhao:2016gc}. Briefly, this method
counts the frequency of rare variants within each gene expression rank
in the population, and aggregates this information over the entire
population to determine whether an enrichment of rare variants exists
within any gene expression rank. We found a strong enrichment for gene
expression extremes for TE variants in the all gene features tested
(Figure \ref{fig4}B). While TE variants in gene upstream regions
showed a strong enrichment of both high and low gene expression ranks,
TE variants in exons or gene downstream regions seemed to have a
stronger enrichment for low expression ranks than high
ranks. Randomization of the accession names removed these enrichments
completely (Figure 4 - figure supplement \hyperref[fig4s1]{1}), and
there was little difference between TE insertions and TE deletions in
the gene expression rank enrichments found (Figure 4 - figure
supplement \hyperref[fig4s2]{2}). This rare variant analysis further
indicated that TE variants may alter the transcript abundance of
nearby genes.

As both increases and decreases in transcript abundance of nearby
genes were observed for TE variants within each gene feature, it
appears to be difficult to predict the impact a TE variant may have on
nearby gene expression. Furthermore, gene-level transcript abundance
measurements may fail to identify the potential positional effect of
TE variants upon transcription. To more closely examine changes in
transcript abundance associated with TE variants among the accessions,
we inspected a subset of TE variant sites and identified TE variants
that appear to have an impact on transcriptional patterns beyond
changes in total transcript abundance from a nearby gene. For example,
the presence of a TE insertion within an exon of \emph{AtRLP18
}(AT2G15040) was associated with truncation of the transcripts at the
TE insertion site in accessions possessing the TE variant, as well as
silencing of a downstream gene encoding a leucine-rich repeat protein
(AT2G15042) (Figure \ref{fig5}A, B). Both genes had significantly
lower transcript abundance in accessions containing the TE insertion
(p \textless{} $5.8\times10^{-10}$, Mann-Whitney U
test). \emph{AtRLP18 }is reported to be involved in bacterial
resistance, with the disruption of this gene by T-DNA insertion
mediated mutagenesis resulting in increased susceptibility to the
bacterial plant pathogen \emph{Pseudomonas syringae}
\cite{Wang:2008km}. We examined pathogen resistance phenotype data
\cite{Aranzana:2005cq}, and found that accessions containing the TE
insertion in the \emph{AtRLP18} exon were more often sensitive to
infection by \emph{Pseudomonas syringae} transformed with
\emph{avrPpH3} genes (Figure \ref{fig5}C). This suggests that the
accessions containing this TE insertion within \emph{AtRLP18} may have
an increased susceptibility to certain bacterial pathogens.

We also observed some TE variants associated with increased expression
of nearby genes. For example, the presence of a TE within the upstream
region of a gene encoding a pentatricopeptide repeat (PPR) protein
(AT2G01360) was associated with higher steady state transcript abundance
of this gene (Figure \ref{fig5}D, E). Transcription appeared to begin at the TE
insertion point, rather than the transcriptional start site of the gene
(Figure \ref{fig5}D). Accessions containing the TE insertion had significantly
higher AT2G01360 transcript abundance than the accessions without the TE
insertion (p \textless{} $1.8\times10^{-7}$, Mann-Whitney U test). The apparent
transcriptional activation, linked with presence of a TE belonging to
the \emph{HELITRON1} family, indicates that this element may carry
sequences or other regulatory information that has altered the
expression of genes downstream of the TE insertion site. Importantly,
this variant was classified as a low-LD TE insertion, as it is not in LD
with surrounding SNPs, and therefore the associated changes in gene
transcript abundance would not be identified using only SNP data. This
TE variant was also upstream of \emph{QPT }(AT2G01350), involved in NAD
biosynthesis \cite{Katoh:2006if}, which did not show alterations in
steady state transcript abundance associated with the presence of the TE
insertion, indicating a potential directionality of regulatory elements
carried by the TE (Figure \ref{fig5}D, E). Overall, these examples demonstrate
that TE variants can have unpredictable, yet important, effects on the
expression of nearby genes, and these effects may be missed by studies
focused on genetic variation at the level of SNPs.

\subsection{TE variants explain many DNA methylation differences between
accessions}

As TEs are frequently highly methylated in Arabidopsis
\cite{Lister:2008bh, Cokus:2008fc, Zhang:2006eo, Zilberman:2006hx}, we
next assessed the DNA methylation state surrounding TE variant sites to
determine whether TE variants might be responsible for some of the
differences in DNA methylation patterns previously observed between the
wild accessions \cite{Schmitz:2013iu}. We found that TE variants were
often physically close to DMRs (Figure \ref{fig6}A). Furthermore, C-DMRs were
more often close to a TE insertion than expected, while they were rarely
near a TE deletion (Table \ref{table3}). CG-DMRs were rarely close to TE insertions
or TE deletions. Overall, we found 48\% of the 13,482 previously
reported population C-DMRs were located within 1 kb of a TE variant
(predominantly TE insertions), while only 15\% of CG-DMRs were within 1
kb of a TE variant (Table \ref{table3}). For C-DMRs, this was significantly more
than expected by chance, while it was significantly less than expected
for CG-DMRs (p \textless{} $1\times10^{-4}$, determined by resampling 10,000
times). To determine if DMR methylation levels were dependent on the
presence/absence of nearby TE variants, we calculated Pearson
correlation coefficients between the DNA methylation level at each DMR
and the presence/absence of the nearest TE variant. We observed a
negative correlation between the distance from a C-DMR to the nearest TE
insertion and the correlation between the DNA methylation level at the
C-DMR with the presence/absence of the TE insertion (Figure \ref{fig6}B). This
suggests a distance-dependent effect of TE insertion presence on C-DMR
methylation. In contrast, we found no such relationship for TE deletions
on C-DMRs, or for insertions or deletions on CG-DMRs (Figure \ref{fig6}B). DNA
methylation levels at C-DMRs located within 1 kb of a TE insertion
(TE-DMRs) were more often positively correlated with the
presence/absence of a TE insertion than the DNA methylation levels at
C-DMRs further than 1 kb from a TE insertion (non-TE-DMRs). This was
evident from the distribution of correlations for non-TE-DMRs being
centred around zero, whereas for TE-DMRs this distribution was skewed to
the right (Figure \ref{fig6}C, \emph{D}=0.23). For TE deletions, we did not
observe such a difference in the distributions of correlation
coefficients between TE-DMRs and non-TE-DMRs, nor for CG-DMRs and their
nearby TE insertions or deletions (Figure \ref{fig6}C, \emph{D=}0.07-0.10).
Furthermore, DNA methylation levels were often higher in the presence of
the nearby TE insertion, while this relationship was generally not
observed for C-DMRs further than 1 kb from a TE variant, for TE
deletions, or for CG-DMRs (Figure 6 - figure supplement \hyperref[fig6s1]{1}).

As the above correlations between TE presence/absence and DMR
methylation level rely on the TE variants having a high MAF, this
precludes an analysis of the effect of rare variants on DMR methylation
levels. To determine the effect that these rare TE variants may have on
DMR methylation levels, we performed a burden test for enrichment of DMR
methylation extremes at TE-DMRs, similar to as was done to test the
effect of rare variants on gene expression. We found a strong enrichment
for high C-DMR and CG-DMR methylation level ranks for TE insertions,
while TE deletions were associated with both high and low extremes of
DNA methylation levels at C-DMRs, and less so at CG-DMRs (Figure \ref{fig6}D).
This further indicates that the presence of a TE insertion is associated
with higher C-DMR methylation levels, while TE deletions seem to have
more variable effects on DMR methylation levels. This enrichment was
completely absent after repeating the analysis with randomized accession
names (Figure 6 - figure supplement \hyperref[fig6s2]{2}). We also observed a slight
enrichment for low DMR methylation ranks for TE insertions near CG-DMRs,
indicating that the insertion of a TE was sometimes associated with
reduced CG methylation in nearby regions (\textless{}1 kb from the TE).
We examined these TE insertions in a genome browser, and found that some
TE insertions were associated with decreased transcript abundance of
nearby genes, with a corresponding loss of gene body methylation,
offering a potential explanation for the decreased CG methylation
observed near some TE insertions (Figure 6 - figure supplement \hyperref[fig6s3]{3}).

We next examined levels of DNA methylation in regions flanking all TE
variants regardless of the presence or absence of a population DMR call.
While DNA methylation levels around pericentromeric TE insertions and
deletions (\textless{}3 Mb from a centromere) seemed to be unaffected by
the presence of a TE insertion (Figure \ref{fig7}A), TE insertions in the
chromosome arms were associated with an increase in DNA methylation
levels in all contexts (Figure \ref{fig7}A, B). In contrast, TE deletions in the
chromosome arms did not affect patterns of DNA methylation, as the
flanking methylation level in all contexts seemed to remain high
following deletion of the TE (Figure \ref{fig7}A, C). As the change in DNA
methylation levels around TE variant sites appeared to be restricted to
regions \textless{}200 bp from the insertion site, we correlated DNA
methylation levels in 200 bp regions flanking TE variants with the
presence/absence of TE the variants. DNA methylation levels were often
positively correlated with the presence of a TE insertion when the
insertion was distant from a centromere (Figure \ref{fig7}D). TE deletions were
more variably correlated with local DNA methylation levels, but also
showed a slight bias towards positive correlations for TE deletions
distant from the centromeres. As methylome data was available for both
leaf and bud tissue for 12 accessions, we repeated this analysis
comparing between tissue types, but did not observe any difference in
the patterns of methylation surrounding TE variant sites between the two
tissues (Figure 7 - figure supplement \hyperref[fig7s1]{1}).

These results indicate that local DNA methylation patterns are
influenced by the differential TE content between genomes, and that the
DNA methylation-dependent silencing of TEs may lead to formation of DMRs
between wild Arabidopsis accessions. TE insertions appear to be
important in defining local patterns of DNA methylation, while DNA
methylation levels often remain elevated following a TE deletion, and so
are independent from the presence or absence of TEs in these cases.
Importantly, the distance from a TE insertion to the centromere appears
to have a strong impact on whether an alteration of local DNA
methylation patterns will occur. This is likely due to flanking
sequences being highly methylated in the pericentromeric regions, and so
the insertion of a TE cannot further increase levels of DNA methylation.
Overall, a large fraction of the population C-DMRs previously identified
between wild accessions are correlated with the presence of local TE
insertions, but not TE deletions. CG-DMR methylation levels seem to be
mostly independent from the presence/absence of common TEs variants,
while rare TE variants have an impact on DNA methylation levels at both
C-DMRs and CG-DMRs.

\subsection{Genome-wide analysis of TE variant association with methylation
levels highlights distant and local control}

To better quantify the above results, an association scan was conducted
for all common TE variants (\textgreater{}3\% MAF) and all population
C-DMRs for the 124 accessions with both DNA methylation and TE variant
data available. To test the significance of each pairwise correlation,
we collected bootstrap p-value estimates based on 500 permutations of
accession labels. TE-DMR associations were deemed significant if they
had a real association more extreme than any of the 500 permutations (p
\textless{} 1/500). A band of significant associations was observed for
TE insertions and their nearby C-DMRs, signifying a local association
between TE insertion presence/absence and C-DMR methylation (Figure \ref{fig8}A).
This local association was not as strong for TE deletions (Figure \ref{fig8}B),
consistent with our above findings. While TE variants and DNA
methylation showed a local association, it is also possible that TE
variation may influence DNA methylation state more broadly in the
genome, perhaps through production of \emph{trans-}acting smRNAs or
inactivation of genes involved in DNA methylation establishment or
maintenance. To identify any potential enrichment of C-DMRs regulated in
\emph{trans}, we summed the total number of significant associations for
each TE variant across the whole genome (Figure \ref{fig8}A and B, top panels).
At many sites, we found far more significant associations than expected
due to the false positive rate alone. This suggested the existence of
many putative \emph{trans }associations between TE variants and
genome-wide C-DMR methylation levels. We further examined these C-DMRs
that appeared to be associated with a TE insertion in \emph{trans},
checking for TE insertions near these C-DMRs that were present in the
same accessions as the \emph{trans }associated TE, as these could lead
to a false \emph{trans }association. These were extremely rare, with
only 4 such cases for TE insertions, and 38 cases for TE deletions, and
so were unable to explain the high degree of \emph{trans }associations
found.

\section{Discussion}

Here we discovered widespread differential TE content between wild
Arabidopsis accessions, and explore the impact of these variants at the
level of individual accessions. Most TE variants were due to the
\emph{de novo }insertion of TEs, while a smaller subset was likely due
to the deletion of ancestral TE copies, mostly around the
pericentromeric regions. A subset (32\%) of TE variants with a minor
allele frequency above 3\% were able to be tested for linkage with
nearby SNPs. The majority of these TE variants were not in LD with SNPs,
indicating that they represent genetic variants currently overlooked in
genomic studies. We found a marked depletion of TE variants within gene
bodies and DNase I hypersensitivity sites (putative regulatory regions),
indicating that the more deleterious TE insertions may have been removed
from this population through selection. Of those TE variants found in
gene bodies, TE deletions were overrepresented, indicating that the loss
of ancestral TEs inserted within genes may be more frequent, or perhaps
less deleterious, than the \emph{de novo }insertion of TEs into genes.

The identification of a large number of TE variants in this population
gave an opportunity to form statistically robust correlations between TE
presence/absence and transcript abundance from nearby genes, as well as
genome-wide patterns of DNA methylation. We were able to identify
examples where TE variants appear to have an effect upon gene
expression, both in the disruption of transcription and in the spreading
or disruption of regulatory information leading to the transcriptional
activation of genes, indicating that these TE variants can have
important consequences upon the expression of protein coding genes
(Figure \ref{fig5}). In one case, these changes in gene expression could be
linked with phenotypic changes, with accessions containing a TE
insertion more frequently sensitive to bacterial infection. Further
experiments will be needed to establish a causal link between this TE
insertion and the associated phenotype. An analysis of rare TE variants,
present at a low MAF, further strengthened this relationship between TE
presence/absence and altered transcript abundance, as we were able to
identify a strong enrichment of rare TE variants in accessions with
extreme gene expression ranks in the population.

Perhaps most importantly, we provide evidence that differential TE
content between genomes of Arabidopsis accessions underlies a large
fraction of the previously reported population C-DMRs, in agreement with
recent similar findings \cite{Quadrana:2016bi}. Thus, the frequency of
pure epialleles, independent of underlying genetic variation, may be
even more rare than previously anticipated \cite{Richards:2006br}. We
did not find evidence of CG-DMR methylation level being altered by the
presence of common TE variants, but rather rare TE variants may be more
important in shaping patterns of DNA methylation at some CG-DMRs, though
the reasons for this distinction remain unclear. The level of local DNA
methylation changes associated with TE variants was also related to the
distance from a TE variant to the centromere, with variants in the
chromosome arms being more strongly correlated with DNA methylation
levels. This seems to be due to a higher baseline level of DNA
methylation at the pericentromeric regions, which prevent any further
increase in DNA methylation level following insertion of a TE.
Furthermore, we found an important distinction between TE insertions and
TE deletions in the effect that these variants have on nearby DNA
methylation levels. While flanking DNA methylation levels appeared to
increase following a TE insertion, the deletion of an ancestral TE was
often not associated with a corresponding decrease in flanking DNA
methylation levels (Figure \ref{fig7}). This indicates that high levels of DNA
methylation, once established, may be maintained in the absence of the
TE insertion that presumably triggered the original change in DNA
methylation level. It is then possible that TE variants explain more of
the variation in DNA methylation patterns than we find direct evidence
for, if some C-DMRs were formed by the insertion of an ancestral TE that
is now absent in all the accessions analysed here. These DMRs would then
represent the epigenetic ``scars'' of past TE insertions.

Finally, we performed a genome-wide scan of common TE variant
association with C-DMR methylation levels, and found further evidence of
a strong local association between TE insertion presence/absence and
C-DMR methylation level (Figure \ref{fig8}). We were also able to identify some
TE variants that appeared to be associated with changes in DNA
methylation levels at multiple loci throughout the genome, indicating a
possible \emph{trans} regulation of DNA methylation levels linked to
certain TE variants. Further experiments will be required to confirm and
examine the role of these TE variants in determining genome-wide
patterns of DNA methylation. Overall our results show that TE
presence/absence variants between wild Arabidopsis accessions not only
have important effects on nearby gene expression, but can also have a
role in determining local patterns of DNA methylation, and explain many
regions of differential DNA methylation previously observed in the
population.

\section{Methods}

\subsection{TEPID development}

\emph{Mapping}

FASTQ files are mapped to the reference genome using the `tepid-map'
algorithm (Figure \ref{fig1}). This first calls bowtie2 \cite{Langmead:2012jh}
with the following options: `--local', `--dovetail', `--fr', `-R5',
`-N1'. Soft-clipped and unmapped reads are extracted using Samblaster
\cite{Faust:2014hf}, and remapped using the split read mapper Yaha
\cite{Faust:2012ic}, with the following options: `-L 11', `-H 2000', `-M
15', `-osh'. Split reads are extracted from the Yaha alignment using
Samblaster \cite{Faust:2014hf}. Alignments are then converted to bam
format, sorted, and indexed using samtools \cite{Li:2009ka}.

\emph{TE variant discovery}

The `tepid-discover' algorithm examines mapped bam files generated by
the `tepid-map' step to identify TE presence/absence variants with
respect to the reference genome. Firstly, mean sequencing coverage, mean
library insert size, and standard deviation of the library insert size
is estimated. Discordant read pairs are then extracted, defined as mate
pairs that map more than 4 standard deviations from the mean insert size
from one another, or on separate chromosomes.

To identify TE insertions with respect to the reference genome, split
read alignments are first filtered to remove reads where the distance
between split mapping loci is less than 5 kb, to remove split reads due
to small indels, or split reads with a mapping quality (MAPQ) less than
5. Split and discordant read mapping coordinates are then intersected
using pybedtools \cite{Dale:2011cl, Quinlan:2010km} with the Col-0
reference TE annotation, requiring 80\% overlap between TE and read
mapping coordinates. To determine putative TE insertion sites, regions
are then identified that contain independent discordant read pairs
aligned in an orientation facing one another at the insertion site, with
their mate pairs intersecting with the same TE (Figure \ref{fig1}). The total
number of split and discordant reads intersecting the insertion site and
the TE is then calculated, and a TE insertion predicted where the
combined number of reads is greater than a threshold determined by the
average sequencing depth over the whole genome (1/10 coverage if
coverage is greater than 10, otherwise a minimum of 2 reads).
Alternatively, in the absence of discordant reads mapped in orientations
facing one another, the required total number of split and discordant
reads at the insertion site linked to the inserted TE is set higher,
requiring twice as many reads.

To identify TE absence variants with respect to the reference genome,
split and discordant reads separated \textgreater{}20 kb from one
another are first removed, as 99.9\% of Arabidopsis TEs are shorter than
20 kb, and this removes split reads due to larger structural variants
not related to TE diversity. Col-0 reference annotation TEs that
are located within the genomic region
spanned by the split and discordant reads are then identified. TE
absence variants are predicted where at least 80\% of the TE sequence is
spanned by a split or discordant read, and the sequencing depth within
the spanned region is \textless{}10\% the sequencing depth of the 2 kb
flanking sequence, and there are a minimum number of split and
discordant reads present, determined by the sequencing depth (1/10
coverage; Figure \ref{fig1}). A threshold of 80\% TE sequence spanned by split
or discordant reads is used, as opposed to 100\%, to account for
misannotation of TE sequence boundaries in the Col-0 reference TE
annotation, as well as TE fragments left behind by DNA TEs during
cut-paste transposition (TE footprints) that may affect the mapping of
reads around annotated TE borders \cite{Plasterk:1991vd}. This was found
to improve TE absence detection using simulated data. Furthermore, the
coverage within the spanned region may be more than 10\% that of the
flanking sequence, but in such cases twice as many split and discordant
reads are required. If multiple TEs are spanned by the split and
discordant reads, and the above requirements are met, multiple TEs in
the same region can be identified as absent with respect to the
reference genome. Absence variants in non-Col-0 accessions are
subsequently recategorized as TE insertions present in the Col-0 genome
but absent from a given wild accession.

\emph{TE variant refinement}

Once TE insertions are identified using the `tepid-map' and
`tepid-discover' algorithms, these variants can be refined if multiple
related samples are analysed. The `tepid-refine' algorithm is designed
to interrogate regions of the genome in which a TE insertion was
discovered in other samples but not the sample in question, and check
for evidence of that TE insertion in the sample using lower read count
thresholds compared to the `tepid-discover' step. In this way, the
refine step leverages TE variant information for a group of related
samples to reduce false negative calls within the group. This
distinguishes TEPID from other similar methods for TE variant discovery
utilizing short sequencing reads. A file containing the coordinates of
each insertion, and a list of sample names containing the TE insertion
must be provided to the `tepid-refine' algorithm, which this can be
generated using the `merge\_insertions.py' script included in the TEPID
package. Each sample is examined in regions where there was a TE
insertion identified in another sample in the group. If there is a
sequencing breakpoint within this region (no continuous read coverage
spanning the region), split reads mapped to this region will be
extracted from the alignment file and their coordinates intersected with
the TE reference annotation. If there are split reads present at the
variant site that are linked to the same TE as was identified as an
insertion at that location, this TE insertion is recorded in a new file
as being present in the sample in question. If there is no sequencing
coverage in the queried region for a sample, an ``NA'' call is made
indicating that it is unknown whether the particular sample contains the
TE insertion or not.

While the above description relates specifically to use of TEPID for
identification of TE variants in Arabidopsis in this study, this method
can be also applied to other species, with the only prerequisite being
the annotation of TEs in a reference genome and the availability of
paired-end DNA sequencing data.

\subsection{TE variant simulation}

To test the sensitivity and specificity of TEPID, 100 TE insertions (50
copy-paste transpositions, 50 cut-paste transpositions) and 100 TE
absence variants were simulated in the Arabidopsis genome using the
RSVSim R package, version 1.7.2 \cite{Bartenhagen:2013kl}, and synthetic
reads generated from the modified genome at various levels of sequencing
coverage using wgsim \cite{Li:2009ka} (\url{https://github.com/lh3/wgsim}).
These reads were then used to calculate the true positive, false
positive, and false negative TE variant discovery rates for TEPID at
various sequencing depths, by running `tepid-map' and `tepid-discover'
using the simulated reads with the default parameters (Figure 1 - figure
supplement \hyperref[fig1s1]{1}).

\subsection{Estimation of sensitivity}

Previously published 100 bp paired end sequencing data for L\emph{er}
(\url{http://1001genomes.org/data/MPI/MPISchneeberger2011/releases/current/Ler-1/Reads/}
; \cite{Schneeberger:2011ft}) was downloaded and analyzed with the TEPID
package to identify TE variants. Reads providing evidence for TE
variants were then mapped to the \emph{de novo} assembled L\emph{er}
genome \cite{Chin:2013iw}. To determine whether reads mapped to
homologous regions of the L\emph{er} and Col-0 reference genome, the
\emph{de novo }assembled L\emph{er }genome sequence between mate pair
mapping locations in L\emph{er} were extracted, with repeats masked
using RepeatMasker with RepBase-derived libraries and the default
parameters (version 4.0.5, \url{http://www.repeatmasker.org}). A blastn
search was then conducted against the Col-0 genome using the following
parameters: `-max-target-seqs 1', `-evalue 1e-6' \cite{Camacho:2009fc}.
Coordinates of the top blast hit for each read location were then
compared with the TE variant sites identified using those reads. To
estimate false negative rates for TEPID TE absence calls, L\emph{er} TE
absence calls were compared with a known set of Col-0-specific TE
insertions, absent in L\emph{er }\cite{Quadrana:2016bi}. For TEPID TE
insertion calls, we mapped Col-0 DNA sequencing reads
\cite{Jiang:2014ih} to the L\emph{er }PacBio assembly, and identified
sites with read evidence reaching the TEPID threshold for a TE insertion
call to be made.

\subsection{Arabidopsis TE variant discovery}

We ran the TEPID, including the insertion refinement step, on previously
published sequencing data for 216 different Arabidopsis populations
(NCBI SRA SRA012474; Schmitz et al. 2013), mapping to the TAIR10
reference genome and using the TAIR9 TE annotation. The `--mask' option
was also used to mask the mitochondrial and plastid genomes. We also ran
TEPID using previously published transgenerational data for salt stress
and control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), using
the `--mask' option to mask mitochondrial and plastid genomes, and the
`--strict' option for highly related samples.

\subsection{TE variant / SNP comparison}

SNP information for 216 Arabidopsis accessions was obtained from the
1001 genomes data center
(\url{http://1001genomes.org/data/Salk/releases/2013_24_01/}; \cite{Schmitz:2013iu}).
This was formatted into reference (Col-0 state), alternate,
or NA calls for each SNP. Accessions with both TE variant information
and SNP data were selected for analysis. Hierarchical clustering of
accessions by SNPs as well as TE variants were used to identify
essentially clonal accessions, as these would skew minor allele
frequency calculations. A single representative from each cluster of
similar accessions was kept, leading to a total of 187 accessions for
comparison. For each TE variant with minor allele frequency greater than
3\%, the nearest 300 upstream and 300 downstream SNPs with a minor
allele frequency greater than 3\% were selected. Pairwise genotype
correlations (r\textsuperscript{2} values) for all complete cases were obtained for
SNP-SNP and SNP-TE variant states. r\textsuperscript{2} values were then ordered by
decreasing rank and a median SNP-SNP rank value was calculated. For each
of the 600 ranked surrounding positions, the number of times the TE rank
was greater than the SNP-SNP median rank was calculated as a relative LD
metric of TE to SNP. TE variants with less than 200 ranks over the
SNP-SNP median were classified as low-LD insertions. TE variants with
ranks between 200 and 400 were classified as mid-LD, while TE variants
with greater than 400 ranks above their respective SNP-SNP median value
were classified as variants in high LD with flanking SNPs.

\subsection{PCR validations}

\emph{Selection of accessions to be genotyped}

To assess the accuracy of TE variant calls in accessions with a range of
sequencing depths of coverage, we grouped accessions into quartiles
based on sequencing depth of coverage and randomly selected a total of
14 accessions for PCR validations from these quartiles. DNA was
extracted for these accessions using Edward's extraction protocol
\cite{Edwards:1991wp}, and purified prior to PCR using AMPure beads.

\emph{Selection of TE variants for validation and primer design}

Ten TE insertion sites and 10 TE absence sites were randomly selected
for validation by PCR amplification. Only insertions and absence
variants that were variable in at least two of the fourteen accessions
selected to be genotyped were considered. For insertion sites, primers
were designed to span the predicted TE insertion site. For TE absence
sites, two primer sets were designed; one primer set to span the TE,
and another primer set with one primer annealing within the TE
sequence predicted to be absent, and the other primer annealing in the
flanking sequence (Figure 2 - figure supplement \hyperref[fig2s3]{3}). Primer
sequences were designed that did not anneal to regions of the genome
containing previously identified SNPs in any of the 216 accessions
\cite{Schmitz:2013iu} or small insertions and deletions, identified
using lumpy-sv with the default settings
\cite{Layer:2014ie}(\url{https://github.com/arq5x/lumpy-sv}), had an
annealing temperature close to 52ºC calculated based on nearest
neighbor thermodynamics (MeltingTemp submodule in the SeqUtils python
module; \cite{Cock:2009hj}), GC content between 40\% and 60\%, and
contained the same base repeated not more than four times in a
row. Primers were aligned to the TAIR10 reference genome using bowtie2
\cite{Langmead:2012jh} with the `-a' flag set to report all
alignments, and those with more than 5 mapping locations in the genome
were then removed.

\emph{PCR}

PCR was performed with 10 ng of extracted, purified Arabidopsis DNA
using Taq polymerase. PCR products were analysed by agarose gel
electrophoresis. Col-0 was used as a positive control, water was added
to reactions as a negative control.

\subsection{mRNA analysis}

Processed mRNA data for 144 wild Arabidopsis accessions were downloaded
from NCBI GEO GSE43858 \cite{Schmitz:2013iu}. To find differential gene
expression dependent on TE presence/absence variation, we first filtered
TE variants to include only those where the TE variant was shared by at
least 5 accessions with RNA data available. We then grouped accessions
based on TE presence/absence variants, and performed a Mann-Whitney U
test to determine differences in RNA transcript abundance levels between
the groups. We used q-value estimation to correct for multiple testing,
using the R qvalue package v2.2.2 with the following parameters: lambda
= seq(0, 0.6, 0.05), smooth.df = 4 \cite{Storey:2003cj}. Genes were
defined as differentially expressed where there was a greater than 2
fold difference in expression between the groups, with a q-value less
than 0.01. Gene ontology enrichment analysis was performed using PANTHER
(\url{http://pantherdb.org}).

\subsection{DNA methylation data analysis}

Processed base-resolution DNA methylation data for wild Arabidopsis
accessions were downloaded from NCBI GEO GSE43857 \cite{Schmitz:2013iu},
and used to construct MySQL tables in a database.

\subsection{Rare variant analysis}

To assess the effect of rare TE variants on gene expression or DMR DNA
methylation levels, we tested for a burden of rare variants in the
population extremes, essentially as described previously
\cite{Zhao:2016gc}. For each rare TE variant near a gene or DMR, we
ranked the gene expression level or DMR DNA methylation level for all
accessions in the population, and tallied the ranks of accessions
containing a rare variant. These rank counts were then binned to produce
a histogram of the distribution of ranks. We then fit a quadratic model
to the counts data, and calculated the R\textsuperscript{2} and p-value for the fit of
the model.

\subsection{TE variant and DMR genome-wide association analysis}

Accessions were subset to those with both leaf DNA methylation data and
TEPID calls. Pairwise correlations were performed for observed data
pairs for each TE variant and a filtered set of population C-DMRs, with
those C-DMRs removed where more than 15\% of the accessions had no
coverage. This amounted to a final set of 9,777 C-DMRs. Accession names
were then permuted to produce a randomized dataset, and pairwise
correlations again calculated. This was repeated 500 times to produce a
distribution of expected Pearson correlation coefficients for each
pairwise comparison. Correlation values more extreme than any of the 500
permutations were deemed significant.

\section{Data access}

TEPID source code can be accessed at
\url{https://github.com/ListerLab/TEPID}. Code and data needed to
reproduce this analysis can be found at
\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er }TE
variants are available in Figure 1 - source data 1 and 2. TE variants
identified among the 216 wild Arabidopsis accessions resequenced by
Schmitz et al. (2013) are available in Figure 2 - source data 1, 2 and
3.

\section{Acknowledgments}

This work was supported by the Australian Research Council (ARC) Centre
of Excellence program in Plant Energy Biology CE140100008 (J.B., R.L.).
R.L. was supported by an ARC Future Fellowship (FT120100862) and Sylvia
and Charles Viertel Senior Medical Research Fellowship, and work in the
laboratory of R.L. was funded by the Australian Research Council. T.S.
was supported by the Jean Rogerson Postgraduate Scholarship. S.R.E. was
supported by an Australian Research Council Discovery Early Career
Research Award (DE150101206). We thank Robert J. Schmitz, Mathew G.
Lewsey, Ronan C. O'Malley, and Ian Small for their critical reading of
the manuscript, and Kevin Murray for his helpful comments regarding the
development of TEPID.

\section{Author contributions}

R.L. and T.S. designed the research project. R.L. and J.B. supervised
research. T.S. developed and tested TEPID. J.C. performed PCR
validations of TE variants. T.S. and S.R.E. performed bioinformatic
analysis. Y.K. provided statistical guidance. R.L., T.S., J.B. and
S.R.E. prepared the manuscript.

\section{Competing financial interests}

The authors declare no competing financial interests.

\pagebreak

\printbibliography

%\bibliography{references}{}
%\bibliographystyle{ieeetr}

% Figure 1

\pagebreak

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\paperwidth]{Figure1/Figure1.png}
  \caption{TE variant discovery pipeline}
  \label{fig1}
\end{figure}

Principle of TE variant discovery using split and discordant read
mapping positions. Paired end reads are first mapped to the reference
genome using Bowtie2 \cite{Langmead:2012jh}. Soft-clipped or unmapped
reads are then extracted from the alignment and re-mapped using Yaha, a
split read mapper \cite{Faust:2012ic}. All read alignments are then used
by TEPID to discover TE variants relative to the reference genome, in
the `tepid-discover' step. When analyzing groups of related samples,
these variants can be further refined using the `tepid-refine' step,
which examines in more detail the genomic regions where there was a TE
variant identified in another sample, and calls the same variant for the
sample in question using lower read count thresholds as compared to the
`tepid-discover' step, in order to reduce false negative variant calls
within a group of related samples.

\pagebreak

% Figure 1 - figure supplement 1

%\setcounter{figure}[0]

\begin{suppfigure}[h]
  \centering
  \includegraphics[width=0.7\paperwidth]{Figure1/Figure1-figure_supplement1.png}
  \caption{figure supplement 1}
  \label{fig1s1}
\end{suppfigure}

Testing of the TEPID pipeline using simulated TE variants in the
Arabidopsis Col-0 genome (TAIR10), for a range of sequencing coverage
levels. TE insertions (A) and TE absence calls (B).

\pagebreak

% Figure 2

%\setcounter{figure}[2]

\begin{figure}[h]
  \centering
  \includegraphics[width=1\textwidth]{Figure2/Figure2.png}
  \caption{Extensive novel genetic diversity uncovered by TE variant analysis}
  \label{fig2}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Distribution of identified TE variants on chromosome 1, with
  distributions of all Col-0 genes, Col-0 TEs, and population DMRs.
\item
  Frequency of TE variants at different genomic features.
\item
  Proportion of TE variants within each genomic feature classified as
  deletions or insertions.
\item
  Proportion of TE variants within each genomic feature classified as
  rare or common.
\item
  Enrichment and depletion of TE variants categorized by TE superfamily
  compared to the expected frequency due to genomic occurrence.
\end{enumerate}

\pagebreak

% Figure 2 - figure supplement 1

\setcounter{suppfigure}{1}
\begin{suppfigure}
  \centering
  \includegraphics{Figure2/Figure2-figure_supplement_1.png}
  \caption{figure supplement 1}
  \label{fig2s1}
\end{suppfigure}

Number of additional TE insertion calls made due to the TEPID refinement
step for each accession in the population.

\pagebreak
% Figure 2 - figure supplement 2
\setcounter{suppfigure}{1}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure2/Figure2-figure_supplement_2.png}
  \caption{figure supplement 2}
  \label{fig2s2}
\end{suppfigure}

Minor allele frequency distribution for TE variants.

\pagebreak
% Figure 2 - figure supplement 3
\setcounter{suppfigure}{1}

\begin{suppfigure}
  \centering
  \includegraphics[width=1\textwidth]{Figure2/Figure2-figure_supplement_3.png}
  \caption{figure supplement 3}
  \label{fig2s3}
\end{suppfigure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  PCR validations for a TE absence variant. Accessions that were
  predicted to contain a TE insertion or TE absence are marked in bold.
  Two primer sets were used; forward (F) and reverse (R) or internal
  (I). Accessions with a TE absence will not produce the FI band and
  produce a shorter FR product, with the change in size matching the
  size of the deleted TE.
\item
  PCR validations for a TE insertion variant. One primer set was used,
  spanning the TE insertion site. A band shift of approximately 200 bp
  can be seen, corresponding to the size of the inserted TE.
\end{enumerate}

\pagebreak
% Figure 2 - figure supplement 4

\setcounter{suppfigure}{1}

\begin{suppfigure}
  \centering
  \includegraphics[width=1\textwidth]{Figure2/Figure2-figure_supplement_4.pdf}
  \caption{figure supplement 4}
  \label{fig2s4}
\end{suppfigure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Number of TE absence variants identified versus the sequencing depth
  of coverage for each accession.
\item
  Number of TE insertion variants identified versus the sequencing depth
  of coverage for each accession.
\item
  Number of additional TE insertion calls made due to the TEPID
  refinement step versus sequencing depth of coverage for all
  accessions.
\end{enumerate}

\pagebreak
% Figure 2 - figure supplement 5

\setcounter{suppfigure}{1}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.7\textwidth]{Figure2/Figure2-figure_supplement_5.png}
  \caption{figure supplement 5. Frequency of TE insertion in the \emph{KNOT} region}
  \label{fig2s5}
\end{suppfigure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Number of TE insertion variants within each 300 kb \emph{KNOT ENGAGED ELEMENT (KEE)},
  vertical lines) and the number of TE insertion variants
  found in 10,000 randomly selected 300 kb windows (histogram).
\item
  Table showing number of TE insertion variants within each \emph{KEE}
  region, and the associated p-value determined by resampling 10,000
  times.
\end{enumerate}

\pagebreak

% Figure 2 - figure supplement 6
\setcounter{suppfigure}{1}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.7\textwidth]{Figure2/Figure2-figure_supplement_6.png}
  \caption{figure supplement 6. Length distribution for all Col-0 TEs and all TE variants}
  \label{fig2s6}
\end{suppfigure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Histogram showing lengths of all annotated TEs in the Col-0 reference
  genome.
\item
  Histogram showing lengths of all TE variants.
\item
  Density distribution of log10 TE length for all Col-0 TEs (red) and TE
  variants (blue).
\end{enumerate}

\pagebreak

% Figure 2 - figure supplement 7
\setcounter{suppfigure}{1}

\begin{suppfigure}[!ht]
  \centering
  \includegraphics[width=1\textwidth]{Figure2/Figure2-figure_supplement_7.png}
  \caption{figure supplement 7}
  \label{fig2s7}
\end{suppfigure}

TE family enrichments and depletions for TE insertions and TE deletions.

\pagebreak

% Figure 3

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.5\textwidth]{Figure3/Figure3.png}
  \caption{Patterns of TE-SNP linkage}
  \label{fig3}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  r\textsuperscript{2} correlation matrices for individual representative high and low-LD
  TE variants showing the background level of SNP-SNP linkage.
\item
  Rank order plots for individual representative high and low-LD TE
  variants (matching those shown in A). Red line indicates the median r\textsuperscript{2}
  value for each rank across SNP-based values. Blue line indicates r\textsuperscript{2}
  values for TE-SNP comparisons. Grey lines indicate all individual
  SNP-SNP comparisons.
\item
  Histogram of the number of TE r\textsuperscript{2} ranks (0-600) that are above the
  SNP-based median r\textsuperscript{2} value for testable TE variants.
\item
  Boxplots showing distribution of minor allele frequencies for each LD
  category. Boxes represent the interquartile range (IQR) from quartile
  1 to quartile 3. Boxplot upper whiskers represent the maximum value,
  or the upper value of the quartile 3 plus 1.5 times the IQR (whichever
  is smaller). Boxplot lower whisker represents the minimum value, or
  the lower value of the quartile 1 minus 1.5 times the IQR (whichever
  is larger).
\end{enumerate}

(E). Proportion of TE insertions, TE deletions, and unclassified TE
variants in each LD category.

\pagebreak

% Figure 3 - figure supplement 1

\setcounter{suppfigure}{2}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.7\textwidth]{Figure3/Figure3-figure_supplement1.png}
  \caption{figure supplement 1}
  \label{fig3s1}
\end{suppfigure}

Distribution of TE variants across chromosome 1 for each LD category
(high, mid, low).

\pagebreak

% Figure 4

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{Figure4/Figure4.png}
  \caption{Differential transcript abundance associated with TE variant presence/absence}
  \label{fig4}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Volcano plots showing transcript abundance differences for genes
  associated with TE insertion variants at different positions,
  indicated in the plot titles. Genes with significantly different
  transcript abundance in accessions with a TE insertion compared to
  accessions without a TE insertion are colored blue (lower transcript
  abundance in accessions containing TE insertion) or red (higher
  transcript abundance in accessions containing TE insertion). Vertical
  lines indicate ±2 fold change in FPKM. Horizontal line indicates the
  1\% FDR.
\item
  Relationship between TE rare variant counts and gene expression rank.
  Plot shows the cumulative number of rare TE variants in equal-sized
  bins for gene expression ranks, from the lowest-ranked accession
  (left) to the highest-ranked accession (right). Lines indicate the fit
  of a quadratic model.
\end{enumerate}

\pagebreak

% Figure 4 - figure supplement 1

\setcounter{suppfigure}{3}

\begin{suppfigure}
  \centering
  \includegraphics[width=1\textwidth]{Figure4/Figure4-figure_supplement1.png}
  \caption{figure supplement 1}
  \label{fig4s1}
\end{suppfigure}

Relationship between rare TE variants and gene expression rank as for
Figure 4B, for permuted TE variants.

\pagebreak

% Figure 4 - figure supplement 2

\setcounter{suppfigure}{3}

\begin{suppfigure}
  \centering
  \includegraphics[width=1\textwidth]{Figure4/Figure4-figure_supplement2.png}
  \caption{figure supplement 2}
  \label{fig4s2}
\end{suppfigure}

Relationship between rare TE variants and gene expression rank as for
Figure 4B, for TE insertions (A) and TE deletions (B) separately.

\pagebreak

% Figure 5

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.8\textwidth]{Figure5/Figure5.png}
  \caption{Effects of TE variants on local gene expression}
  \label{fig5}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Genome browser representation of RNA-seq data for genes \emph{AtRLP18}
  (AT2G15040) and a leucine-rich repeat family protein (AT2G15042) for
  Db-1, containing a TE insertion within the exon of the gene
  \emph{AtRLP18}, and for a Col-0 (not containing the TE insertion
  within the exon of \emph{AtRLP18}). Inset shows magnified view of the
  TE insertion site.
\item
  Heatmap showing \emph{AtRLP18 }and AT2G15042 RNA-seq FPKM values for
  all accessions.
\item
  Percentage of accessions with resistance to \emph{Pseudomonas syringae
  }transformed with different \emph{avr }genes, for accessions
  containing or not containing a TE insertion in* AtRLP18.*
\item
  Genome browser representation of RNA-seq data for a PPR
  protein-encoding gene (AT2G01360) and \emph{QPT }(AT2G01350), showing
  transcript abundance for these genes in accessions containing a TE
  insertion variant in the upstream region of these genes.
\item
  Heatmap representation of RNA-seq FPKM values for \emph{QPT }and a
  gene encoding a PPR protein (AT2G01360), for all accessions. Note that
  scales are different for the two heatmaps, due to the higher
  transcript abundance of \emph{QPT} compared to AT2G01360. Scale
  maximum for AT2G01350 is $3.1\times10^5$, and for AT2G01360 is $5.9\times10^4$.
\end{enumerate}

\pagebreak

% Figure 6

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.8\textwidth]{Figure6/Figure6.png}
  \caption{TE variants are associated with nearby DMR methylation levels}
  \label{fig6}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Distribution of distances from TE variants to the nearest population
  DMR, for TE deletions and TE insertion, C-DMRs and CG-DMRs.
\item
  Pearson correlation between DMR DNA methylation level and TE
  presence/absence, for all DMRs and their closest TE variant, versus
  the distance from the DMR to the TE variant (log scale). Blue lines
  show a linear regression between the correlation coefficients and the
  log10 distance to the TE variant.
\item
  Empirical cumulative distribution of Pearson correlation coefficients
  between TE presence/absence and DMR methylation level for TE
  insertions, TE deletions, C-DMRs and CG-DMRs. The Kolmogorov--Smirnov
  statistic is shown in each plot, indicated by \emph{D.}
\item
  Relationship between rare TE variant counts and nearby DMR DNA
  methylation level ranks, for TE insertions, deletions, C-DMRs, and
  CG-DMRs. Plot shows the cumulative number of rare TE variants in
  equal-sized bins of DMR methylation level ranks, from the lowest
  ranked accession (left) to the highest ranked accession (right). Lines
  indicate the fit of a quadratic model, and the corresponding R\textsuperscript{2} and p
  values are shown in each plot.
\end{enumerate}

\pagebreak

% Figure 6 - figure supplement 1

\setcounter{suppfigure}{5}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.6\textwidth]{Figure6/Figure6-figure_supplement1.png}
  \caption{figure supplement 1}
  \label{fig6s1}
\end{suppfigure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  DNA methylation density distribution at C-DMRs within 1 kb of a TE
  variant (TE-DMRs) or further than 1 kb from a TE variant
  (non-TE-DMRs), in the presence or absence of the TE, for TE insertions
  and TE deletions.
\item
  As for A, for CG-DMRs.
\end{enumerate}

\pagebreak

% Figure 6 - figure supplement 2

\setcounter{suppfigure}{5}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure6/Figure6-figure_supplement2.png}
  \caption{figure supplement 2}
  \label{fig6s2}
\end{suppfigure}

Cumulative number DMR methylation level ranks for DMRs near rare TE
variants with accessions selected at random. Lines indicate the fit of a
quadratic model, and the corresponding R\textsuperscript{2} and p values are shown in each
plot.

\pagebreak

% Figure 6 - figure supplement 3

\setcounter{suppfigure}{5}

\begin{suppfigure}
  \centering
  \includegraphics[width=1\textwidth]{Figure6/Figure6-figure_supplement3.png}
  \caption{figure supplement 3}
  \label{fig6s3}
\end{suppfigure}

Selected examples of TE insertions apparently associated with
transcriptional downregulation of nearby genes and loss of gene body CG
methylation leading to the formation of a CG-DMR.

\pagebreak

% Figure 7

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{Figure7/Figure7.png}
  \caption{Local patterns of DNA methylation surrounding TE variant sites}
  \label{fig7}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Heatmap showing DNA methylation levels in 200 bp bins flanking TE
  variant sites, +/- 2 kb from the TE insertion point. TE variants were
  grouped into pericentromeric variants (\textless{}3 Mb from a
  centromere) or variants in the chromosome arms (\textgreater{}3 Mb
  from a centromere).
\item
  Line plot showing the DNA methylation level in each sequence context
  for TE insertion sites, +/- 2 kb from the TE insertion point.
\item
  As for B, for TE deletions.
\item
  Distribution of Pearson correlation coefficients between TE
  presence/absence and DNA methylation levels in the 200 bp regions
  flanking TE variant, ordered by distance to the centromere.
\end{enumerate}

\pagebreak

% Figure 7 - figure supplement 1

\setcounter{suppfigure}{6}

\begin{suppfigure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure7/Figure7-figure_supplement1.png}
  \caption{figure supplement 1}
  \label{fig7s1}
\end{suppfigure}

Heatmap showing DNA methylation levels in 200 bp bins flanking TE
variant sites in the 12 accessions with DNA methylation data for both
leaf and bud tissue, +/- 2 kb from the TE insertion point. TE variants
were grouped into pericentromeric variants (\textless{}3 Mb from a
centromere) or variants in the chromosome arms (\textgreater{}3 Mb from
a centromere).

\pagebreak

% Figure 8

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.5\textwidth]{Figure8/Figure8.png}
  \caption{Association scan between TE variants can C-DMR methylation variation}
  \label{fig8}
\end{figure}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\item
  Significant correlations between TE insertions and C-DMR DNA
  methylation level. Points show correlations between individual TE-DMR
  pairs that were more extreme than any of 500 permutations of the DMR
  data. Top plots show the total number of significant correlations for
  each TE insertion across the whole genome.
\item
  As for (A), for TE deletions.
\end{enumerate}

\pagebreak

\begin{table}[h]
  \centering
  \begin{threeparttable}
    \caption{Mapping of paired-end reads providing evidence for TE presence/absence
      variants in the L\emph{er} reference genome}
    \label{table1}
    \begin{tabular}{@{}cccccc@{}}
      \toprule
      \textbf{}  & \textbf{Concordant} & \textbf{Discordant} & \textbf{Split} & \textbf{Unmapped} & \textbf{Total} \\ \midrule
      Col-0 mapped & 0 & 993 & 9513 & 0 & 10206 \\
      L\emph{er} mapped & 10073 & 92 & 34 & 7 & 10206 \\ \bottomrule
    \end{tabular}
    \begin{tablenotes}
      \small
    \item Note: Discordant and split read categories are not mutually exclusive, as some discordant reads may have one read in the mate pair split-mapped.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

\vspace*{1 cm}

\begin{table}[h]
  \centering
  \caption{Summary of TE variant classifications}
  \label{table2}
  \begin{tabular}{@{}ccc@{}}
    \toprule
    \textbf{TEPID call}        & \textbf{TE classification} & \textbf{Count} \\ \midrule
    \multirow{3}{*}{Insertion} & NA                         & 310            \\
    & Insertion                  & 14689          \\
    & Deletion                  & 8              \\ \midrule
    \multirow{3}{*}{Absence}   & NA                         & 1852           \\
    & Insertion                  & 388            \\
    & Deletion                   & 5848  \\ \bottomrule
  \end{tabular}
\end{table}

\vspace*{1 cm}

\begin{table}[h]
  \centering
  \caption{Percentage of DMRs within 1 kb of a TE variant}
  \label{table3}
  \begin{tabular}{@{}cccc|ccc@{}}
    \toprule
    \textbf{}              & \multicolumn{3}{c|}{\textbf{C-DMRs}}                     & \multicolumn{3}{c}{\textbf{CG-DMRs}}                    \\ \midrule
    \textbf{}              & \textbf{Observed} & \textbf{Expected} & \textbf{95\% CI} & \textbf{Observed} & \textbf{Expected} & \textbf{95\% CI} \\ \midrule
    \textbf{TE deletions}  & 8.7               & 16                & 0.0078           & 4.3               & 16                & 0.0041           \\
    \textbf{TE insertions} & 36                & 26                & 0.0089           & 9.4               & 26                & 0.0047           \\
    \textbf{NA calls}      & 3.4               & 6.2               & 0.0052           & 1.7               & 6.2               & 0.0027           \\
    \textbf{Total}         & 48                & 41                & 0.01             & 15                & 42                & 0.0054           \\ \bottomrule
  \end{tabular}
\end{table}

\end{document}

### 2016-08-12 AnalysisCode/TE_insertion_distribution.R
Add comparison of type1/type2 TE distribution

@@ -10,7 +10,7 @@ tepav <- read_tsv("../RawData/TEPID_TEPAV.tsv.gz", col_names = T)
 tepav %>%
   mutate(Absence_call = grepl("Col-0", Accessions_TE_present)) %>%
   group_by(Absence_call, Absence_classification) %>%
-  summarise(count = n()) %>% View
+  summarise(count = n()) %>%
   write_tsv("../ProcessedData/deletion_stats.tsv")
 
 # 1. Plot minor allele frequency
@@ -125,6 +125,26 @@ dev.off()
 pdf("../Plots/chr1_heatmap_scale.pdf", height=2, width=5)
 image(data.matrix(seq(100)), col = color)
 dev.off()
+
+# Now compare insertion distribution of Type 1 and Type 2 elements (reviewer request)
+# First add annotation of TE superfamily to each variant
+te <- rename(all_te, TE = X5, family = X6, superfamily = X7) %>%
+  select(TE, family, superfamily)
+
+k <- select(te, superfamily) %>% rename(s = superfamily) %>% unique()
+t <- c(1, 2, 2, 1, 2, 2, 1, 2, 2, 1, 2, 2, NA, 2, 1, 1, 1, 1)
+names(t) <- k$s
+
+annotations <- left_join(tepav, te) %>% mutate(class = t[superfamily])
+
+annotations %>%
+  filter(chromosome == "chr1", !is.na(Absence_classification), !is.na(class)) %>%
+  ggplot(., aes(start, fill=Absence_classification)) +
+  geom_density(adjust=1/7) +
+  facet_wrap(Absence_classification~class) +
+  theme_bw() +
+  theme(legend.position = "none")
+ggsave("../Plots/type_1_type_2_distribution.pdf", height = 5, width = 5, useDingbats=F)
 
 # 3. Find frequency that insertions / deletions occur in different genomic features
 # common header for most files

### 2016-08-12 AnalysisCode/TE_DMR_analysis.R
Look at non-TE-DMRs

@@ -274,19 +274,20 @@ system("gzip ../ProcessedData/c_dmr_correlations.tsv")
 system("gzip ../ProcessedData/cg_dmr_correlations.tsv")
 
 ## WHAT GENOMIC FEATURES ARE TE-DMRS AND NON-TE-DMRS FOUND IN?
-# te_c_dmr %>%
-#   filter(AbsenceClassification == "No insertion") %>%
-#   select(dmr_feature, TE_close) %>%
-#   group_by(dmr_feature) %>%
-#   mutate(total = n()) %>%
-#   group_by(dmr_feature, TE_close) %>%
-#   mutate(count = n()) %>%
-#   rowwise() %>%
-#   filter(count > 200) %>%
-#   mutate(perc = count / total * 100) %>%
-#   select(dmr_feature, TE_close, perc) %>%
-#   unique() %>%
-#   ggplot(., aes(dmr_feature, perc, fill=TE_close)) + geom_bar(stat="identity", position="dodge", color = "black") + theme_bw()
+te_c_dmr %>%
+  select(dmr_feature, TE_close, AbsenceClassification) %>%
+  group_by(dmr_feature) %>%
+  mutate(total = n()) %>%
+  group_by(dmr_feature, TE_close) %>%
+  mutate(count = n()) %>%
+  rowwise() %>%
+  filter(count > 200) %>%
+  mutate(perc = count / total * 100) %>%
+  select(dmr_feature, TE_close, perc, AbsenceClassification) %>%
+  unique() %>%
+  ggplot(., aes(dmr_feature, perc, fill=TE_close)) +
+  geom_bar(stat="identity", position="dodge", color = "black") + theme_bw() +
+  facet_wrap(~AbsenceClassification)
 
 ### C-DMRs ###
 # Make dataframe

### 2016-08-12 Paper/paper.tex
Minor formatting changes
Changed font
Update text

@@
 \setlength\bibitemsep{\baselineskip}
+
+\usepackage[scaled]{helvet}
+\renewcommand\familydefault{\sfdefault}
+\usepackage[T1]{fontenc}
 
@@
 
+\renewcommand\Authfont{\fontsize{13}{14.4}\selectfont}
+\renewcommand\Affilfont{\fontsize{10}{10.8}\selectfont}
+
 \date{}
@@
 
-Variation in the presence or absence of transposable elements (TEs) is a
-major source of genetic variation between individuals. Here, we
-identified 23,095 TE presence/absence variants between 216 wild
-Arabidopsis accessions. Most variants are rare, and we find a burden of
-rare variants associated with extremes of both gene expression and DNA
-methylation levels within the population. Of the common alleles
+Variation in the presence or absence of transposable elements (TEs) is
+a major source of genetic variation between individuals. Here, we
+identified 23,095 TE presence/absence variants between 216 Arabidopsis
+accessions. Most TE variants were rare, and we find a burden of rare
+variants associated with local extremes of both gene expression and
+DNA methylation levels within the population. Of the common alleles
 identified, two thirds were not in linkage disequilibrium with nearby
@@
 diversity. Nearly 200 common TE variants were associated with
-significantly altered expression of nearby genes, with potential
-functional consequences including decreased pathogen resistance. A major
-fraction of inter-accession DNA methylation differences were associated
-with nearby TE insertion variants, with DNA, indicating an important
-role of TE variants in facilitating epigenomic variation.
+significantly altered expression of nearby genes, and a major fraction
+of inter-accession DNA methylation differences were associated with
+nearby TE insertion variants. Overall, we find that TE variants likely
+play an important role in facilitating epigenomic and transcriptional
+variation between wild Arabidopsis accessions, indicating a strong
+genetic basis for epigenetic variation.
+
+\pagebreak
 
@@
 Transposable elements (TEs) are mobile genetic elements present in
-nearly all studied organisms, and make up a large fraction of most
+nearly all studied organisms, and comprise a large fraction of most
 eukaryotic genomes. The two classes of TEs are retrotransposons (class
@@
 \cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
-insertion may disrupt essential regions of the genome. Consequently,
-safeguards have evolved in order to suppress this activity. These
-safeguards include epigenetic transcriptional silencing mechanisms,
-chiefly involving the methylation of cytosine nucleotides (DNA
-methylation) to produce 5-methylcytosine (mC), a mark that can signal
-transcriptional silencing of the methylated locus. In
-\emph{Arabidopsis thaliana }(Arabidopsis), DNA methylation occurs in
-three DNA sequence contexts: mCG, mCHG, and mCHH, where H is any base
-but G. Establishment of DNA methylation marks can be carried out by
-two distinct pathways -- the RNA-directed DNA methylation pathway
-guided by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2
-pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
+insertion may disrupt functional regions of the genome. Consequently,
+safeguard mechanisms have evolved to suppress this activity, including
+the methylation of cytosine nucleotides (DNA methylation) to produce
+5-methylcytosine (mC), a modification that can induce transcriptional
+silencing of the methylated locus. In \emph{Arabidopsis thaliana}
+(Arabidopsis), DNA methylation occurs in all three DNA sequence
+contexts: mCG, mCHG, and mCHH, where H is any base but
+G. Establishment of DNA methylation marks can be carried out by two
+distinct pathways -- the RNA-directed DNA methylation pathway guided
+by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2 pathway
+\cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
 methylation in Arabidopsis is in the transcriptional silencing of TEs.
-Loss of DNA methylation due to mutations in genes essential for DNA
-methylation establishment and maintenance leads to expression of
-previously silent TEs, and in some cases transposition
-\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm,
-  Jeddeloh:1999gu, Zemach:2013dj}.
+Loss of DNA methylation due to mutations in genes essential for its
+establishment or maintenance leads to expression of previously silent
+TEs, and in some cases transposition \cite{Mirouze:2009km,Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu,Zemach:2013dj}.
 
@@
 transcriptional and post-transcriptional silencing of TEs can lead to
-bursts of TE activity, quickly generating new genetic diversity
+bursts of TE activity, rapidly generating new genetic diversity
 \cite{Vitte:2014he}. TEs may carry regulatory information such as
@@
 pairs affected, a large fraction of genetic differences between
-Arabidopsis accessions appear to be due to variation in TE content
+Arabidopsis accessions appears to be due to variation in TE content
 \cite{Cao:2011cf, Quadrana:2016bi}. Therefore identification of TE
@@
 The tools developed previously for identification of novel TE insertion
-sites have several limitations. They either require a library of active
-TE sequences, cannot identify TE absence variants, or are not designed
+events have several limitations. They either require a library of active
+TE sequences, cannot identify TE absence variants, are not designed
 with population studies in mind, or suffer from a high degree of
@@
 In order to further assess the sensitivity and specificity of TE variant
-discovery using TEPID, we identified TE variants in the \emph{Landsberg}
+discovery using TEPID, we identified TE variants in the Landsberg
 \emph{erecta} (L\emph{er}) accession, and compared these with the
@@
 evidence for these variants were then mapped to the L\emph{er} reference
-genome that was generated by \emph{de novo} assembly using Pacific
+genome, generated by \emph{de novo} assembly using Pacific
 Biosciences P5-C3 chemistry with a 20 kb insert library
@@
 absences in L\emph{er} genome identified previously by aligning
-full-length Col-0 TEs to the Ler reference using BLAT
+full-length Col-0 TEs to the L\emph{er} reference using BLAT
 \cite{Quadrana:2016bi}. We found that 89.6\% (173/193) of these TE
@@
 resequencing data for 216 different Arabidopsis accessions
-\cite{Schmitz:2013iu}, and identified 15,007 TE insertions and 8,088 TE
-absence variants relative to the Col-0 reference accession, totalling
-23,095 unique TE variants. In most accessions we identified 300-500 TE
-insertions (mean = 378) and 1,000-1,500 TE absence variants (mean =
-1,279), the majority of which were shared by two or more accessions
-(Figure 2 - figure supplement \hyperref[fig2s2]{2}). PCR validations were performed for a
-random subset of 10 insertions and 10 absence variants in 14 accessions
-(totalling 280 validations), and confirmed the high accuracy of TE
-variant discovery using the TEPID package, with results similar to that
-observed using simulated data and the L\emph{er} genome analysis (Figure 2 -
-figure supplement \hyperref[fig2s3]{3}). The number of TE insertions identified was
+\cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
+source data 1) and 8,088 TE absence variants (Figure 2 - source data
+2) relative to the Col-0 reference accession, totalling 23,095 unique
+TE variants. In most accessions we identified 300-500 TE insertions
+(mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
+majority of which were shared by two or more accessions (Figure 2 -
+figure supplement \hyperref[fig2s2]{2}). PCR validations were
+performed for a random subset of 10 insertions and 10 absence variants
+in 14 accessions (totalling 280 validations), and confirmed the high
+accuracy of TE variant discovery using the TEPID package, with results
+similar to that observed using simulated data and the L\emph{er}
+genome analysis (Figure 2 - figure supplement
+\hyperref[fig2s3]{3}). The number of TE insertions identified was
 positively correlated with sequencing depth of coverage, while the
 number of TE absence variants identified had no correlation with
-sequencing coverage (Figure 2 - figure supplement \hyperref[fig2s4]{4}A, B), indicating
-that the sensitivity of TE absence calls is not limited by sequencing
-depth, while TE insertion calls benefitted from high sequencing depth.
-However, accessions with low coverage gained more TE insertion calls
-during the TEPID refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
+sequencing coverage (Figure 2 - figure supplement
+\hyperref[fig2s4]{4}A, B), indicating that the sensitivity of TE
+absence calls is not limited by sequencing depth, while TE insertion
+calls benefitted from high sequencing depth.  However, accessions with
+low coverage gained more TE insertion calls during the TEPID
+refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
 indicating that these false negatives were effectively reduced by
@@
 under the expectation that the minor allele is the derived allele. We
-re-classified common TE absences relative to Col-0 as rare TE insertions
-in Col-0, and common TE insertions relative to Col-0 as true TE
-deletions in Col-0. Cases where the TE variant was at an high MAF
-(\textgreater{}20\%) were assigned NA calls, as it could not be
+re-classified common TE absences relative to Col-0 as rare TE
+insertions in Col-0, and common TE insertions relative to Col-0 as
+true TE deletions in Col-0. Cases where the TE variant was at an high
+MAF (\textgreater{}20\%) were assigned NA calls, as it could not be
 determined if these were cases where the variant was most likely to be
 due to a true TE deletion or due to a new TE insertion. While these
-classifications are not definitive, as there will be rare cases where a
-true TE deletion has spread through the population and becomes the
-common allele, we will correctly classify most TE variants. Overall, we
-found 72.3\% of the TE absence variants identified with respect to the
-Col-0 reference genome were likely due to a true TE deletion in these
-accessions, while 4.8\% were due to new insertions in Col-0 not shared
-by others in the population (Table \ref{table2}). Overall, we identified 15,077 new
-TE insertions, 5,856 true TE deletions, and 2,162 TE variants at a high
-MAF that were unable to be classified as an insertion or deletion.
+classifications are not definitive, as there will be rare cases where
+a true TE deletion has spread through the population and becomes the
+common allele, we will correctly classify most TE variants. Overall,
+we found 72.3\% of the TE absence variants identified with respect to
+the Col-0 reference genome were likely due to a true TE deletion in
+these accessions, while 4.8\% were due to new insertions in Col-0 not
+shared by others in the population (Table \ref{table2}). Overall, we
+identified 15,077 new TE insertions, 5,856 true TE deletions, and
+2,162 TE variants at a high MAF that were unable to be classified as
+an insertion or deletion (Figure 2 - source data 3).
 
@@
 insertions, as the insertions we analysed in this population were likely
-older than those used in the study by Grob et al.
-
-Among the identified TE variants, several TE superfamilies were over- or
-under-represented compared to the number expected by chance given the
-overall genomic frequency of different TE types (Figure \ref{fig2}E). In
-particular, both TE insertions and deletions in the RC/Helitron
-superfamily were less numerous than expected, with an 11.5\% depletion
-of RC/Helitron elements in the set of TE variants. In contrast, TEs
-belonging to the LTR/Gypsy superfamily were more frequently deleted than
-expected, with a 17\% enrichment in the set of TE deletions. This was
-unlikely to be due to a differing ability of our detection methods to
-identify TE variants of different lengths, as the TE variants identified
-had a similar distribution of lengths as all Arabidopsis TEs annotated
-in the Col-0 reference genome (Figure 2 - figure supplement \hyperref[fig2s6]{6}). These
-enrichments suggest that the RC/Helitron TEs have been relatively
-dormant in recent evolutionary history, while the LTR/Gypsy, which are
-highly enriched in the pericentromeric regions, are frequently lost from
-the Arabidopsis genome. At the family level, we observed similar
-patterns of TE variant enrichment or depletion (Figure 2 - figure
-supplement \hyperref[fig2s7]{7}; source data 5).
+older than those used in the \emph{KEE} study \cite{Grob:2014kg}.
+
+Among the identified TE variants, several TE superfamilies were over-
+or under-represented compared to the number expected by chance given
+the overall genomic frequency of different TE types (Figure
+\ref{fig2}E). In particular, both TE insertions and deletions in the
+RC/Helitron superfamily were less numerous than expected, with an
+11.5\% depletion of RC/Helitron elements in the set of TE variants. In
+contrast, TEs belonging to the LTR/Gypsy superfamily were more
+frequently deleted than expected, with a 17\% enrichment in the set of
+TE deletions. This was unlikely to be due to a differing ability of
+our detection methods to identify TE variants of different lengths, as
+the TE variants identified had a similar distribution of lengths as
+all Arabidopsis TEs annotated in the Col-0 reference genome (Figure 2
+- figure supplement \hyperref[fig2s6]{6}). These enrichments suggest
+that the RC/Helitron TEs have been relatively dormant in recent
+evolutionary history, while the LTR/Gypsy, which are highly enriched
+in the pericentromeric regions, are frequently lost from the
+Arabidopsis genome. At the family level, we observed similar patterns
+of TE variant enrichment or depletion (Figure 2 - figure supplement
+\hyperref[fig2s7]{7}; source data 4).
 
 We further examined Arabidopsis (Col-0) DNA sequencing data from a
-transgenerational stress experiment to investigate the possible minimum
-number of generations required for TE variants to arise
-\cite{Jiang:2014ih}. We identified a single potential TE insertion in a
-sample following 10 generations of single-seed descent under high
-salinity stress conditions, and no TE variants in the control
-single-seed descent set. However, without experimental validation it
-remains unclear if this represents a true variant. Therefore, we
-conclude that TE variants may arise at a rate less than 1 insertion in
-30 generations under laboratory conditions. Further experimental work
-will be required to precisely determine the rate of transposition in
-Arabidopsis.
+transgenerational stress experiment to investigate the possible
+minimum number of generations required for TE variants to arise
+\cite{Jiang:2014ih}. In one of the three replicates, we identified a
+single potential TE insertion in a sample following 10 generations of
+single-seed descent under high salinity stress conditions, and no TE
+variants in the control single-seed descent set. However, without
+experimental validation it remains unclear if this represents a true
+variant. Therefore, we conclude that TE variants may arise at a rate
+less than 1 insertion in 30 generations under laboratory
+conditions. Further experimental work will be required to precisely
+determine the rate of transposition in Arabidopsis.
 
@@
 represent a previously unassessed source of genetic variation between
-accessions. SNPs that were previously identified between the accessions
-\cite{Schmitz:2013iu} were compared to the presence/absence of
-individual TE variants. For the testable TE variants in the population,
-the nearest 300 flanking SNPs upstream and downstream of the TE variant
-site were analyzed for local linkage disequilibrium (LD, r\textsuperscript{2}; see
-methods). TE variants were classified as being either `low', `mid', or
-`high' LD variants by comparing ranked r\textsuperscript{2} values of TE variant to SNPs
-against the median ranked r\textsuperscript{2} value for all between SNP comparisons
-(SNP-SNP) to account for regional variation in the extent of SNP-SNP LD
-(Figure \ref{fig3}A, B) due to recombination rate variation or selection
-\cite{Horton:2012jh}. The majority (61\%) of testable TE variants had
-low LD with nearby SNPs, and represent a source of genetic diversity not
-previously assessed by current SNP-based genotype calling methods
-(Figure \ref{fig3}C). 29\% of TE variants displayed high levels of LD and are
-tagged by nearby SNPs, while only 10\% had intermediate levels of LD. We
-observed a positive correlation between TE variant MAF and LD state,
-with variants of a high minor allele frequency more often classified as
-high-LD (Figure \ref{fig3}D). While the proportion of TE variants classified as
-high, mid, or low-LD was mostly the same for both TE insertions and TE
+accessions. SNPs that were previously identified between the
+accessions \cite{Schmitz:2013iu} were compared to the presence/absence
+of individual TE variants. For the testable TE variants in the
+population, the nearest 300 flanking SNPs upstream and downstream of
+the TE variant site were analyzed for local linkage disequilibrium
+(LD, r\textsuperscript{2}; see methods). TE variants were classified
+as being either `low', `mid', or `high' LD variants by comparing
+ranked r\textsuperscript{2} values of TE variant to SNPs against the
+median ranked r\textsuperscript{2} value for all between SNP
+comparisons (SNP-SNP) to account for regional variation in the extent
+of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
+variation or selection \cite{Horton:2012jh}. The majority (61\%) of
+testable TE variants had low LD with nearby SNPs, and represent a
+source of genetic diversity not previously assessed by current
+SNP-based genotype calling methods (Figure \ref{fig3}C). 29\% of TE
+variants displayed high levels of LD and are tagged by nearby SNPs,
+while only 10\% had intermediate levels of LD. We observed a positive
+correlation between TE variant MAF and LD state, with variants of a
+high minor allele frequency more often classified as high-LD (Figure
+\ref{fig3}D). While the proportion of TE variants classified as high,
+mid, or low-LD was mostly the same for both TE insertions and TE
 deletions, TE variants with a high MAF (\textgreater{}20\%) that were
 unable to be classified as either true deletions or as new insertions
-had a much higher proportion of high-LD variants (Figure \ref{fig3}E). This was
-consistent with the observation that the more common alleles were more
-often in a high-LD state. TE variants displayed a similar distribution
-over chromosome one regardless of linkage classification (Figure 3 -
-figure supplement \hyperref[fig3s1]{1}). Overall, this analysis revealed an abundance of
-previously unexplored genetic variation that exists amongst Arabidopsis
-accessions caused by the presence or absence of TEs, and illustrates the
-importance of identifying TE variants alongside other genetic diversity
-such as SNPs.
+had a much higher proportion of high-LD variants (Figure
+\ref{fig3}E). This was consistent with the observation that the more
+common alleles were more often in a high-LD state. TE variants
+displayed a similar distribution over chromosome one regardless of
+linkage classification (Figure 3 - figure supplement
+\hyperref[fig3s1]{1}). Overall, this analysis revealed an abundance of
+previously unexplored genetic variation that exists amongst
+Arabidopsis accessions caused by the presence or absence of TEs, and
+illustrates the importance of identifying TE variants alongside other
+genetic diversity such as SNPs.
 
@@
 insertions, but not TE deletions. CG-DMR methylation levels seem to be
-mostly independent from the presence/absence of common TEs variants,
+mostly independent from the presence/absence of common TE variants,
 while rare TE variants have an impact on DNA methylation levels at both
@@
 
-\subsection{Genome-wide analysis of TE variant association with methylation
-levels highlights distant and local control}
+\subsection{Genome-wide association scan highlights
+  distant and local control of DNA methylation}
 
@@
 accession labels. TE-DMR associations were deemed significant if they
-had a real association more extreme than any of the 500 permutations (p
+had an association more extreme than any of the 500 permutations (p
 \textless{} 1/500). A band of significant associations was observed for
@@
 bodies and DNase I hypersensitivity sites (putative regulatory regions),
-indicating that the more deleterious TE insertions may have been removed
+consistent with the more deleterious TE insertions have being removed
 from this population through selection. Of those TE variants found in
@@
 split and discordant reads separated \textgreater{}20 kb from one
-another are first removed, as 99.9\% of Arabidopsis TEs are shorter than
-20 kb, and this removes split reads due to larger structural variants
-not related to TE diversity. Col-0 reference annotation TEs that
-are located within the genomic region
-spanned by the split and discordant reads are then identified. TE
-absence variants are predicted where at least 80\% of the TE sequence is
-spanned by a split or discordant read, and the sequencing depth within
-the spanned region is \textless{}10\% the sequencing depth of the 2 kb
-flanking sequence, and there are a minimum number of split and
-discordant reads present, determined by the sequencing depth (1/10
-coverage; Figure \ref{fig1}). A threshold of 80\% TE sequence spanned by split
-or discordant reads is used, as opposed to 100\%, to account for
-misannotation of TE sequence boundaries in the Col-0 reference TE
-annotation, as well as TE fragments left behind by DNA TEs during
-cut-paste transposition (TE footprints) that may affect the mapping of
-reads around annotated TE borders \cite{Plasterk:1991vd}. This was found
-to improve TE absence detection using simulated data. Furthermore, the
-coverage within the spanned region may be more than 10\% that of the
-flanking sequence, but in such cases twice as many split and discordant
-reads are required. If multiple TEs are spanned by the split and
-discordant reads, and the above requirements are met, multiple TEs in
-the same region can be identified as absent with respect to the
-reference genome. Absence variants in non-Col-0 accessions are
-subsequently recategorized as TE insertions present in the Col-0 genome
-but absent from a given wild accession.
+another are first removed, as 99.9\% of Arabidopsis TEs are shorter
+than 20 kb, and this removes split reads due to larger structural
+variants not related to TE diversity (Figure 2 - figure supplement
+\hyperref{fig2s6}). Col-0 reference annotation TEs that are located
+within the genomic region spanned by the split and discordant reads
+are then identified. TE absence variants are predicted where at least
+80\% of the TE sequence is spanned by a split or discordant read, and
+the sequencing depth within the spanned region is \textless{}10\% the
+sequencing depth of the 2 kb flanking sequence, and there are a
+minimum number of split and discordant reads present, determined by
+the sequencing depth (1/10 coverage; Figure \ref{fig1}). A threshold
+of 80\% TE sequence spanned by split or discordant reads is used, as
+opposed to 100\%, to account for misannotation of TE sequence
+boundaries in the Col-0 reference TE annotation, as well as TE
+fragments left behind by DNA TEs during cut-paste transposition (TE
+footprints) that may affect the mapping of reads around annotated TE
+borders \cite{Plasterk:1991vd}. Furthermore, the coverage within the
+spanned region may be more than 10\% that of the flanking sequence,
+but in such cases twice as many split and discordant reads are
+required. If multiple TEs are spanned by the split and discordant
+reads, and the above requirements are met, multiple TEs in the same
+region can be identified as absent with respect to the reference
+genome. Absence variants in non-Col-0 accessions are subsequently
+recategorized as TE insertions present in the Col-0 genome but absent
+from a given wild accession.
 
@@
 
-We ran the TEPID, including the insertion refinement step, on previously
+We ran TEPID, including the insertion refinement step, on previously
 published sequencing data for 216 different Arabidopsis populations
-(NCBI SRA SRA012474; Schmitz et al. 2013), mapping to the TAIR10
+(NCBI SRA SRA012474; \cite{Schmitz:2013iu}), mapping to the TAIR10
 reference genome and using the TAIR9 TE annotation. The `--mask' option
-was also used to mask the mitochondrial and plastid genomes. We also ran
+was set to mask the mitochondrial and plastid genomes. We also ran
 TEPID using previously published transgenerational data for salt stress
-and control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), using
+and control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), again using
 the `--mask' option to mask mitochondrial and plastid genomes, and the
@@
 similar accessions was kept, leading to a total of 187 accessions for
-comparison. For each TE variant with minor allele frequency greater than
+comparison. For each TE variant with a minor allele frequency greater than
 3\%, the nearest 300 upstream and 300 downstream SNPs with a minor
@@
 reproduce this analysis can be found at
-\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er }TE
-variants are available in Figure 1 - source data 1 and 2. TE variants
-identified among the 216 wild Arabidopsis accessions resequenced by
-Schmitz et al. (2013) are available in Figure 2 - source data 1, 2 and
-3.
+\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er
+}TE variants are available in Figure 1 - source data 1 and 2. TE
+variants identified among the 216 wild Arabidopsis accessions
+resequenced by Schmitz et al. (2013) are available in Figure 2 -
+source data 1, 2 and 3. Source data is available on Dryad
+(\url{http://dx.doi.org/10.5061/dryad.187b3}).
 
@@
 
-Minor allele frequency distribution for TE variants.
+Number of accessions sharing TE variants identified by TEPID.
 
@@
   }transformed with different \emph{avr }genes, for accessions
-  containing or not containing a TE insertion in* AtRLP18.*
+  containing or not containing a TE insertion in \emph{AtRLP18}.
 \item
@@
   Distribution of distances from TE variants to the nearest population
-  DMR, for TE deletions and TE insertion, C-DMRs and CG-DMRs.
+  DMR, for TE deletions and TE insertions, C-DMRs and CG-DMRs.
 \item
@@
   \includegraphics[width=0.5\textwidth]{Figure8/Figure8.png}
-  \caption{Association scan between TE variants can C-DMR methylation variation}
+  \caption{Association scan between TE variants and C-DMR methylation variation}
   \label{fig8}

### 2016-08-17 AnalysisCode/TE_DMR_analysis.R
Code fix, regenerated DMR overlap estimates

@@ -122,7 +122,7 @@ bootstrapping <- cbind(bootstrap_c_dmr, bootstrap_c_dmr_insertions_1kb, bootstra
 # What is the percentage of all C-DMRs with a TE variant within 1 kb
 close_c_dmrs <- te_c_dmr %>% 
   group_by(AbsenceClassification) %>%
-  filter(distance > 1000) %>%
+  filter(distance < 1000) %>%
   select(dmr_chr, dmr_start, dmr_stop, AbsenceClassification) %>% 
   unique() %>%
   summarise(count = n())
@@ -288,6 +288,26 @@ te_c_dmr %>%
   ggplot(., aes(dmr_feature, perc, fill=TE_close)) +
   geom_bar(stat="identity", position="dodge", color = "black") + theme_bw() +
   facet_wrap(~AbsenceClassification)
+
+# How many C-DMRs are distant from a TE variant, but close to a fixed TE?
+# Add distance of each C-DMR to closest TE. Find percentage of C-DMRs not close to a TE variant
+# that are within 1 kb of a TE (will be a lot)
+c_dmr_tepav_distance <- read_tsv("../ProcessedData/c_dmr_tepav_distance.tsv.gz",
+                                 col_names = c("dmr_chr", "dmr_start", "dmr_stop", "dmr_tepav_distance"))
+c_dmr_te_distance <- read_tsv("../ProcessedData/c_dmrs_closest_te.tsv.gz",
+                              col_names = c("dmr_chr", "dmr_start", "dmr_stop", "dmr_te", "dmr_te_distance"))
+
+dmr_distances <- left_join(c_dmr_tepav_distance, c_dmr_te_distance)
+
+dmr_distances %>%
+  filter(dmr_tepav_distance > 1000) %>%
+  ggplot(., aes(dmr_te_distance)) +
+  geom_histogram(bins = 100) +
+  geom_vline(xintercept = 1000)
+
+dmr_distances %>%
+  filter(dmr_tepav_distance > 1000, dmr_te_distance < 1000) %>%
+  count()
 
 ### C-DMRs ###
 # Make dataframe

### 2016-08-17 ProcessingCode/add_dmr_features.sh

@@ -1,27 +1,44 @@
 # Alter DMR files with the genomic features that they intersect
+# Add distance from each DMR to the closest TE
 
 bedtools intersect -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -wb \
-| cut -f1,2,3,7 \
-> c_dmr_intersections.bed
+    | cut -f1,2,3,7 \
+	  > c_dmr_intersections.bed
 
 bedtools intersect -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -v \
-| awk 'BEGIN {FS=OFS="\t"} {print $0, "intergenic"}' \
-| cat c_dmr_intersections.bed - \
-| sort -k1,1 -k2,2n \
-| gzip \
-> ../ProcessedData/c_dmr_intersections.tsv.gz
+    | awk 'BEGIN {FS=OFS="\t"} {print $0, "intergenic"}' \
+    | cat c_dmr_intersections.bed - \
+    | sort -k1,1 -k2,2n \
+    | gzip \
+	  > ../ProcessedData/c_dmr_intersections.tsv.gz
 
 rm c_dmr_intersections.bed
 
 bedtools intersect -a ../RawData/cg_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -wb \
-| cut -f1,2,3,7 \
-> cg_dmr_intersections.bed
+    | cut -f1,2,3,7 \
+	  > cg_dmr_intersections.bed
 
 bedtools intersect -a ../RawData/cg_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -v \
-| awk 'BEGIN {FS=OFS="\t"} {print $0, "intergenic"}' \
-| cat cg_dmr_intersections.bed - \
-| sort -k1,1 -k2,2n \
-| gzip \
-> ../ProcessedData/cg_dmr_intersections.tsv.gz
+    | awk 'BEGIN {FS=OFS="\t"} {print $0, "intergenic"}' \
+    | cat cg_dmr_intersections.bed - \
+    | sort -k1,1 -k2,2n \
+    | gzip \
+	  > ../ProcessedData/cg_dmr_intersections.tsv.gz
 
 rm cg_dmr_intersections.bed
+
+# distance from each C-DMR to closest TE
[…]

### 2016-08-17 ProcessingCode/bootstrap_te_variants.sh

@@ -27,7 +27,7 @@ for (( i=1; i <= 10000; i++ )); do
     > ../ProcessedData/random_coords.bed
 
     # c-dmrs
-    bedtools closest -b ../ProcessedData/random_coords.bed -a ../RawData/c_dmrs.tsv.gz -d -wb \
+    bedtools closest -wb -d -t first -b ../ProcessedData/random_coords.bed -a ../RawData/c_dmrs.tsv.gz \
     | awk 'BEGIN {FS=OFS="\t"} {if ($7 < 1000) print $1, $2, $3}' - \
     | sort -k1,1 -k2,2n - \
     | uniq \
@@ -35,7 +35,7 @@ for (( i=1; i <= 10000; i++ )); do
     >> ../ProcessedData/random_selections_c_dmr_rep_10k.txt
 
     #cg-dmrs
-    bedtools closest -b ../ProcessedData/random_coords.bed -a ../RawData/cg_dmrs.tsv.gz -d -wb \
+    bedtools closest -wb -d -t first -b ../ProcessedData/random_coords.bed -a ../RawData/cg_dmrs.tsv.gz \
     | awk 'BEGIN {FS=OFS="\t"} {if ($7 < 1000) print $1, $2, $3}' - \
     | sort -k1,1 -k2,2n - \
     | uniq \
@@ -49,7 +49,7 @@ for (( i=1; i <= 10000; i++ )); do
     > ../ProcessedData/random_coords_insertions.bed
 
     # c-dmrs
-    bedtools closest -b ../ProcessedData/random_coords_insertions.bed -a ../RawData/c_dmrs.tsv.gz -d -wb \
+    bedtools closest -wb -d -t first -b ../ProcessedData/random_coords_insertions.bed -a ../RawData/c_dmrs.tsv.gz \
     | awk 'BEGIN {FS=OFS="\t"} {if ($7 < 1000) print $1, $2, $3}' - \
     | sort -k1,1 -k2,2n - \
     | uniq \
@@ -57,7 +57,7 @@ for (( i=1; i <= 10000; i++ )); do
     >> ../ProcessedData/random_selections_c_dmr_rep_10k_insertions.txt
 
     #cg-dmrs
-    bedtools closest -b ../ProcessedData/random_coords_insertions.bed -a ../RawData/cg_dmrs.tsv.gz -d -wb \
[…]

### 2016-08-17 ProcessingCode/closest_DMRs.sh

@@ -1,9 +1,9 @@
-gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
-| bedtools closest -b - -a ../ProcessedData/c_dmr_intersections.tsv.gz -d -wb \
+gzip -dc ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
+| bedtools closest -d -t first -b - -a ../ProcessedData/c_dmr_intersections.tsv.gz -wb \
 | gzip - \
 > ../ProcessedData/TE_C_DMR_distances.bed.gz
 
-gunzip -c ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
-| bedtools closest -b - -a ../ProcessedData/cg_dmr_intersections.tsv.gz -d -wb \
+gzip -dc ../RawData/TEPID_TEPAV.tsv.gz | sed 1d \
+| bedtools closest -d -t first -b - -a ../ProcessedData/cg_dmr_intersections.tsv.gz -wb \
 | gzip - \
 > ../ProcessedData/TE_CG_DMR_distances.bed.gz

### 2016-08-18 AnalysisCode/TE_DMR_analysis.R
Fixed errors in calculating TE/DMR overlaps

@@ -152,13 +152,13 @@ overlaps <- data_frame(TEPAV = c("TE deletions", "TE insertions", "NA calls", "T
                                 close_c_dmrs[2,2][[1]]/n_cdmr * 100,
                                 close_c_dmrs[3,2][[1]]/n_cdmr * 100,
                                 sum(close_c_dmrs[,2]/n_cdmr * 100)),
-                       expectC = c(means$c_dmr_dels, means$c_dmr_ins, means$c_dmr_na, means$c_dmr_all),
+                       expectC = c(means$c_dmr_dels, means$c_dmr_ins, means$c_dmr_na, sum(means$c_dmr_ins, means$c_dmr_na, means$c_dmr_dels)),
                        CIC = c(conf$c_dmr_dels, conf$c_dmr_ins, conf$c_dmr_na, conf$c_dmr_all),
                        observeCG = c(close_cg_dmrs[1,2][[1]]/n_cgdmr * 100,
                                  close_cg_dmrs[2,2][[1]]/n_cgdmr * 100,
                                  close_cg_dmrs[3,2][[1]]/n_cgdmr * 100,
                                  sum(close_cg_dmrs[,2]/n_cgdmr * 100)),
-                       expectCG = c(means$cg_dmr_dels, means$cg_dmr_ins, means$cg_dmr_na, means$cg_dmr_all),
+                       expectCG = c(means$cg_dmr_dels, means$cg_dmr_ins, means$cg_dmr_na, sum(means$cg_dmr_ins, means$cg_dmr_na, means$cg_dmr_dels)),
                        CICG = c(conf$cg_dmr_dels, conf$cg_dmr_ins, conf$cg_dmr_na, conf$cg_dmr_all)
                        )
 
@@ -305,9 +305,17 @@ dmr_distances %>%
   geom_histogram(bins = 100) +
   geom_vline(xintercept = 1000)
 
+# How many DMRs are near a fixed TE but distant from a variable TE?
 dmr_distances %>%
-  filter(dmr_tepav_distance > 1000, dmr_te_distance < 1000) %>%
+  filter(dmr_tepav_distance > 1000, dmr_te_distance < 1000) %>% 
   count()
+
+# Which DMRs are distant from TEs of any type?
[…]

### 2016-08-18 ProcessingCode/add_dmr_features.sh

@@ -1,11 +1,13 @@
 # Alter DMR files with the genomic features that they intersect
 # Add distance from each DMR to the closest TE
 
-bedtools intersect -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -wb \
+bedtools intersect -wa -wb -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz \
     | cut -f1,2,3,7 \
+    | sort -k1,1 -k2,2n \
+    | bedtools merge -i - -c 4 -o distinct \
 	  > c_dmr_intersections.bed
 
-bedtools intersect -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -v \
+bedtools intersect -wa -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -v \
     | awk 'BEGIN {FS=OFS="\t"} {print $0, "intergenic"}' \
     | cat c_dmr_intersections.bed - \
     | sort -k1,1 -k2,2n \
@@ -14,11 +16,13 @@ bedtools intersect -a ../RawData/c_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/
 
 rm c_dmr_intersections.bed
 
-bedtools intersect -a ../RawData/cg_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -wb \
+bedtools intersect -wa -wb -a ../RawData/cg_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz \
     | cut -f1,2,3,7 \
+    | sort -k1,1 -k2,2n \
+    | bedtools merge -i - -c 4 -o distinct \
 	  > cg_dmr_intersections.bed
 
-bedtools intersect -a ../RawData/cg_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -v \
+bedtools intersect -wa -a ../RawData/cg_dmrs.tsv.gz -b ../ProcessedData/GeneFeatures/genomic_features.bed.gz -v \
     | awk 'BEGIN {FS=OFS="\t"} {print $0, "intergenic"}' \
     | cat cg_dmr_intersections.bed - \
     | sort -k1,1 -k2,2n \

### 2016-08-18 Paper/paper.tex
Updates to text

@@
 
+\usepackage{cmbright}
+\usepackage{sansmath}
+\sansmath
 \usepackage[scaled]{helvet}
 \renewcommand\familydefault{\sfdefault}
+\usepackage[math]{blindtext}
+
 \usepackage[T1]{fontenc}
@@
 
-Variation in the presence or absence of transposable elements (TEs) is
-a major source of genetic variation between individuals. Here, we
+Variation in the presence or absence of transposable elements (TEs) is a
+major source of genetic variation between individuals. Here, we
 identified 23,095 TE presence/absence variants between 216 Arabidopsis
 accessions. Most TE variants were rare, and we find a burden of rare
-variants associated with local extremes of both gene expression and
-DNA methylation levels within the population. Of the common alleles
+variants associated with local extremes of gene expression and DNA
+methylation levels within the population. Of the common alleles
 identified, two thirds were not in linkage disequilibrium with nearby
-SNPs, implicating these TE variants as a source of novel genetic
-diversity. Nearly 200 common TE variants were associated with
-significantly altered expression of nearby genes, and a major fraction
-of inter-accession DNA methylation differences were associated with
-nearby TE insertion variants. Overall, we find that TE variants likely
-play an important role in facilitating epigenomic and transcriptional
-variation between wild Arabidopsis accessions, indicating a strong
-genetic basis for epigenetic variation.
+SNPs, implicating these variants as a source of novel genetic diversity.
+Nearly 200 common TE variants were associated with significantly altered
+expression of nearby genes, and a major fraction of inter-accession DNA
+methylation differences were associated with nearby TE insertions.
+Overall, this demonstrates that TE variants are a rich source of genetic
+diversity that likely plays an important role in facilitating epigenomic
+and transcriptional differences between individuals, and indicates a
+strong genetic basis for epigenetic variation.
 
@@
 nearly all studied organisms, and comprise a large fraction of most
-eukaryotic genomes. The two classes of TEs are retrotransposons (class
-I elements), which transpose via an RNA intermediate requiring a
-reverse transcription reaction, and DNA transposons (class II
-elements), which transpose via either a cut-paste or, in the case of
-Helitrons, a rolling circle mechanism with no RNA intermediate
-\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
-insertion may disrupt functional regions of the genome. Consequently,
-safeguard mechanisms have evolved to suppress this activity, including
-the methylation of cytosine nucleotides (DNA methylation) to produce
-5-methylcytosine (mC), a modification that can induce transcriptional
-silencing of the methylated locus. In \emph{Arabidopsis thaliana}
-(Arabidopsis), DNA methylation occurs in all three DNA sequence
-contexts: mCG, mCHG, and mCHH, where H is any base but
-G. Establishment of DNA methylation marks can be carried out by two
-distinct pathways -- the RNA-directed DNA methylation pathway guided
-by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2 pathway
-\cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
-methylation in Arabidopsis is in the transcriptional silencing of TEs.
-Loss of DNA methylation due to mutations in genes essential for its
-establishment or maintenance leads to expression of previously silent
-TEs, and in some cases transposition \cite{Mirouze:2009km,Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu,Zemach:2013dj}.
+eukaryotic genomes. The two classes of TEs are retrotransposons (type I
+elements), which transpose via an RNA intermediate requiring a reverse
+transcription reaction, and DNA transposons (type II elements), which
+transpose via either a cut-paste or, in the case of Helitrons, a rolling
+circle mechanism with no RNA intermediate \cite{Wicker:2007en}. TE
+activity poses mutagenic potential as a TE insertion may disrupt
+functional regions of the genome. Consequently, safeguard mechanisms
+have evolved to suppress this activity, including the methylation of
+cytosine nucleotides (DNA methylation) to produce 5-methylcytosine (mC),
+a modification that can induce transcriptional silencing of the
+methylated locus. In \emph{Arabidopsis thaliana }(Arabidopsis), DNA
+methylation occurs in all three DNA sequence contexts: mCG, mCHG, and
+mCHH, where H is any base but G. Establishment of DNA methylation marks
+can be carried out by two distinct pathways -- the RNA-directed DNA
+methylation pathway guided by 24 nucleotide (nt) small RNAs (smRNAs),
+and the DDM1/CMT2 pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major
+function of DNA methylation in Arabidopsis is in the transcriptional
+silencing of TEs. Loss of DNA methylation due to mutations in genes
+essential for its establishment or maintenance leads to expression of
+previously silent TEs, and in some cases transposition
+\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu, Zemach:2013dj}.
 
@@
 events have several limitations. They either require a library of active
-TE sequences, cannot identify TE absence variants, are not designed
-with population studies in mind, or suffer from a high degree of
+TE sequences, cannot identify TE absence variants, are not designed with
+population studies in mind, or suffer from a high degree of
 false-negatives
@@
 able to accurately identify TE presence/absence variants with respect to
-the Col-0 reference genome. We applied our TE analysis method to
-existing genome resequencing data for 216 different wild Arabidopsis
-accessions, and identified widespread TE variation amongst these
-accessions \cite{Schmitz:2013iu}.
+the Col-0 reference genome. We applied our TE variant identification
+method to existing genome resequencing data for 216 different
+Arabidopsis accessions \cite{Schmitz:2013iu}, identifying widespread TE
+variation amongst these accessions and enabling exploration of TE
+diversity and links to gene regulation and epigenomic variation.
 
@@
 respect to a reference TE annotation (Figure \ref{fig1}; see methods). This
-typically takes 5-10 minutes using Arabidopsis sequencing data at 20-40x
-coverage (excluding the read mapping step). After TE variant discovery
-has been performed, TEPID then includes a second refinement step
-designed for population studies. This examines each region of the genome
-where there was a TE insertion identified in any member of the group,
-and checks for evidence of this insertion in all other members of the
-group. In this way, TEPID leverages TE variant information for a group
-of related samples to correct false negative calls within the group.
-Testing of TEPID using simulated TE variants in the Arabidopsis genome
-showed that it was able to reliably detect simulated TE variants at
-sequencing coverage levels commonly used in genomics studies (Figure 1 -
-figure supplement \ref{fig1s1}).
+typically takes 5-10 minutes per accession for Arabidopsis genomic DNA
+sequencing data at 20-40x coverage, excluding the read mapping step.
+After TE variant discovery has been performed, TEPID then includes a
+second refinement step designed for population studies. This examines
+each region of the genome where there was a TE insertion identified in
+any of the analyzed samples, and checks for evidence of this insertion
+in all other samples. In this way, TEPID leverages TE variant
+information for a group of related samples to reduce false negative
+calls within the group. Testing of TEPID using simulated TE variants in
+the Arabidopsis genome showed that it was able to reliably detect
+simulated TE variants at sequencing coverage levels commonly used in
+genomics studies (Figure 1 - figure supplement \ref{fig1s1}).
 
@@
 evidence for these variants were then mapped to the L\emph{er} reference
-genome, generated by \emph{de novo} assembly using Pacific
-Biosciences P5-C3 chemistry with a 20 kb insert library
-\cite{Chin:2013iw}, using the same alignment parameters as was used to
-map reads to the Col-0 reference genome. This resulted in 98.7\% of
-reads being aligned concordantly to the L\emph{er} reference, whereas
-100\% aligned discordantly or as split reads to the Col-0 reference
-genome (Table \ref{table1}). To find whether reads mapped to homologous regions in
-both the Col-0 and L\emph{er} reference genomes, we conducted a blast
-search \cite{Camacho:2009fc} using the DNA sequence between read pair
-mapping locations in the L\emph{er }genome against the Col-0 genome, and
-found the top blast result for 80\% of reads providing evidence for TE
+genome, generated by \emph{de novo} assembly using Pacific Biosciences
+P5-C3 chemistry with a 20 kb insert library \cite{Chin:2013iw}, using
+the same alignment parameters as were used to map reads to the Col-0
+reference genome. This resulted in 98.7\% of reads being aligned
+concordantly to the L\emph{er} reference, whereas 100\% aligned
+discordantly or as split reads to the Col-0 reference genome (Table \ref{table1}).
+To find whether reads mapped to homologous regions in both the Col-0 and
+L\emph{er} reference genomes, we conducted a blast search
+\cite{Camacho:2009fc} using the DNA sequence between read pair mapping
+locations in the L\emph{er }genome against the Col-0 genome, and found
+the top blast result for 80\% of reads providing evidence for TE
 insertions, and 89\% of reads providing evidence for TE absence variants
 in L\emph{er}, to be located within 200 bp of the TE variant
-reported by TEPID. We conclude that reads providing evidence for TE
-variants map discordantly or as split reads when mapped to the Col-0
-reference genome, but map concordantly to homologous regions of the
-L\emph{er} \emph{de novo} assembled reference genome, indicating that
-structural variation is present at the sites identified by TEPID, and
-that this is resolved in the \emph{de novo} assembled genome.
+reported by TEPID. Thus, reads providing evidence for TE variants map
+discordantly or as split reads when mapped to the Col-0 reference
+genome, but map concordantly to homologous regions of the L\emph{er} \emph{de
+novo} assembled reference genome, indicating that structural variation
+is present at the sites identified by TEPID, and that this is resolved
+in the \emph{de novo} assembled genome.
 
@@
 absences were also identified using TEPID, indicating a false negative
-rate of around 10\% for TE absence calls. To determine the rate of false
-negative TE insertion calls, we ran TEPID using 90 bp paired-end Col-0
-reads (Col-0 control samples from \cite{Jiang:2014ih}), aligning reads
-to the L\emph{er} PacBio assembly. As TEPID requires a high-quality TE
-annotation to discover TE variants, which is not available for the
-L\emph{er} assembly, we simply looked for discordant and split read
-evidence at the known Col-0-specific TE insertion sites
+rate of \textasciitilde{}10\% for TE absence calls. To determine the
+rate of false negative TE insertion calls, we ran TEPID using 90 bp
+paired-end Col-0 reads (Col-0 control samples from \cite{Jiang:2014ih}),
+aligning reads to the L\emph{er }PacBio assembly. As TEPID requires a
+high-quality TE annotation to discover TE variants, which is not
+available for the L\emph{er }assembly, we looked for discordant and
+split read evidence at the known Col-0-specific TE insertion sites
 \cite{Quadrana:2016bi}, and found evidence reaching the TEPID threshold
 for a TE insertion call to be made at 89.6\% (173/193) of these sites,
-indicating a false negative rate of 10\%. However, we note that this
-estimate does not take into account the TEPID refinement step used on
-large populations, and so the false negative rate for samples analyzed
-in the population from Schmitz et al. (2013) is likely to be slightly
+indicating a false negative rate of \textasciitilde{}10\%. However, it should be noted
+that this estimate does not take into account the TEPID refinement step
+used on large populations, and so the false negative rate for samples
+analyzed in the population from Schmitz et al. (2013) is likely to be
 lower than this estimate, as each accession gained on average 4\% more
@@
 
-We used TEPID to analyze previously published 100 bp paired-end genome
-resequencing data for 216 different Arabidopsis accessions
+TEPID was used to analyze previously published 100 bp paired-end
+genome resequencing data for 216 different Arabidopsis accessions
 \cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
@@
 2) relative to the Col-0 reference accession, totalling 23,095 unique
-TE variants. In most accessions we identified 300-500 TE insertions
+TE variants. In most accessions TEPID identified 300-500 TE insertions
 (mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
@@
 figure supplement \hyperref[fig2s2]{2}).  PCR validations were
-performed for a random subset of 10 insertions and 10 absence variants
-in 14 accessions (totalling 280 validations), and confirmed the high
-accuracy of TE variant discovery using the TEPID package, with results
-similar to that observed using simulated data and the L\emph{er}
-genome analysis (Figure 2 - figure supplement
+performed for a random subset of 10 insertion and 10 absence variants
+in 14 accessions (totalling 280 validations), confirming the high
+accuracy of TE variant discovery using the TEPID package, with a false
+positive rate for both TE insertion and TE absence identification of
+\textasciitilde{}9\%, similar to that observed using simulated data
+and the L\emph{er} genome analysis (Figure 2 - figure supplement
 \hyperref[fig2s3]{3}). The number of TE insertions identified was
@@
 absence calls is not limited by sequencing depth, while TE insertion
-calls benefitted from high sequencing depth.  However, accessions with
-low coverage gained more TE insertion calls during the TEPID
-refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
-indicating that these false negatives were effectively reduced by
-leveraging TE variant information for the whole population.
+identification benefits from high sequencing depth.  However,
+accessions with low coverage gained more TE insertion calls during the
+TEPID refinement step (Figure 2 - figure supplement
+\hyperref[fig2s4]{4}C), indicating that these false negatives were
+effectively reduced by leveraging TE variant information for the whole
+population.
 
@@
 comparisons and classify each variant as a new TE insertion or true
-deletion of an ancestral TE in the population. To do this, we examined
-the minor allele frequency (MAF) of each variant in the population,
-under the expectation that the minor allele is the derived allele. We
-re-classified common TE absences relative to Col-0 as rare TE
-insertions in Col-0, and common TE insertions relative to Col-0 as
-true TE deletions in Col-0. Cases where the TE variant was at an high
-MAF (\textgreater{}20\%) were assigned NA calls, as it could not be
-determined if these were cases where the variant was most likely to be
-due to a true TE deletion or due to a new TE insertion. While these
-classifications are not definitive, as there will be rare cases where
-a true TE deletion has spread through the population and becomes the
-common allele, we will correctly classify most TE variants. Overall,
-we found 72.3\% of the TE absence variants identified with respect to
-the Col-0 reference genome were likely due to a true TE deletion in
-these accessions, while 4.8\% were due to new insertions in Col-0 not
-shared by others in the population (Table \ref{table2}). Overall, we
-identified 15,077 new TE insertions, 5,856 true TE deletions, and
-2,162 TE variants at a high MAF that were unable to be classified as
-an insertion or deletion (Figure 2 - source data 3).
-
-TE insertions and deletions were distributed throughout chromosome 1 in
-a pattern that was similar to the distribution of all Col-0 TEs (Figure
-\ref{fig2}A). TE deletions and common TE variants were found in similar
-chromosomal regions, as deletion variants represent the rare loss of
-common variants. TE deletions and common variants were more highly
-enriched in the pericentromeric regions than rare variants or TE
-insertions. The distribution of rare TE variants and TE insertions was
-somewhat similar to that observed for regions of the genome previously
-identified as being differentially methylated in all DNA methylation
-contexts (mCG, mCHG, mCHH) between the wild accessions (population
-C-DMRs), while population CG-DMRs, differentially methylated in the mCG
-context, less frequently overlapped with all types of TE variants
-identified \cite{Schmitz:2013iu}. Furthermore, TE variants were depleted
-within genes and DNase I hypersensitivity sites \cite{Sullivan:2014ii},
-while they were enriched in gene flanking regions and within other
-annotated TEs or pseudogenes (Figure \ref{fig2}B). We found that TE deletions and
-common TE variants were enriched within the set of TE variants found in
-gene bodies (Figure \ref{fig2}C, D). We did not find any significant enrichment
-of TE variants within the \emph{KNOT ENGAGED ELEMENT (KEE)} regions,
-previously identified as regions that may act as a ``TE sink''
-\cite{Grob:2014kg} (Figure 2 - figure supplement \hyperref[fig2s5]{5}). This may indicate
-that these regions do not act as a ``TE sink'' as has been previously
-proposed, or that the ``TE sink'' activity is restricted to very recent
-insertions, as the insertions we analysed in this population were likely
-older than those used in the \emph{KEE} study \cite{Grob:2014kg}.
+deletion of an ancestral TE in the population. To do this,the minor
+allele frequency (MAF) of each variant in the population was examined,
+under the expectation that the minor allele is the derived allele.
+Common TE absences relative to Col-0 were re-classified as TE insertions
+in Col-0, and common TE insertions relative to Col-0 as true TE
+deletions in Col-0. Cases where the TE variant had a high MAF
+(\textgreater{}20\%) were assigned NA calls, as it could not be
+determined if these were cases where the variant was most likely to be a
+true TE deletion or a new TE insertion. While these classifications are
+not definitive, as there will be rare cases where a true TE deletion has
+spread through the population and becomes the common allele, it will
+correctly classify most TE variants. Overall, 72.3\% of the TE absence
+variants identified with respect to the Col-0 reference genome were
+likely due to a true TE deletion in these accessions, while 4.8\% were
+due to insertions in Col-0 not shared by other accessions in the
+population (Table \ref{table2}). Overall, we identified 15,077 TE insertions, 5,856
+true TE deletions, and 2,162 TE variants at a high MAF that were unable
+to be classified as an insertion or deletion (Figure 2 - source data 3).
+
+TE insertions and deletions were distributed throughout chromosome 1
+in a pattern that was similar to the distribution of all Col-0 TEs
+(Figure \ref{fig2}A). TE deletions and common TE variants were found
+in similar chromosomal regions, as deletion variants represent the
+rare loss of common variants. TE deletions and common variants were
+more highly enriched in the pericentromeric regions than rare variants
+or TE insertions. Among TE deletions, type II elements were slightly
+less biased towards the centromeres in comparison to the distribution
+of type I elements (Figure 2 - figure supplement
+\hyperref[fig2s5]{5}). The distribution of rare TE variants and TE
+insertions was similar to that observed for regions of the genome
+previously identified as being differentially methylated in all DNA
+methylation contexts (mCG, mCHG, mCHH) between the wild accessions
+(population C-DMRs), while population CG-DMRs, differentially
+methylated in the mCG context, less frequently overlapped with all
+types of TE variants identified \cite{Schmitz:2013iu}. Furthermore, TE
+variants were depleted within genes and DNase I hypersensitivity sites
+\cite{Sullivan:2014ii}, while they were enriched in gene flanking
+regions and within other annotated TEs or pseudogenes (Figure
+\ref{fig2}B). TE deletions and common TE variants were enriched within
+the set of TE variants found in gene bodies (Figure 2C, D). No
+significant enrichment was found for TE variants within the \emph{KNOT
+  ENGAGED ELEMENT (KEE) }regions, previously identified as regions
+that may act as a ``TE sink'' \cite{Grob:2014kg} (Figure 2 - figure
+supplement \hyperref[fig2s6]{6}). This may indicate that these regions
+do not act as a ``TE sink'' as has been previously proposed, or that
+the ``TE sink'' activity is restricted to very recent insertions, as
+the insertions we analysed in this population were likely older than
+those used in the \emph{KEE} study \cite{Grob:2014kg}.
 
@@
 TE deletions. This was unlikely to be due to a differing ability of
-our detection methods to identify TE variants of different lengths, as
+the detection method to identify TE variants of different lengths, as
 the TE variants identified had a similar distribution of lengths as
 all Arabidopsis TEs annotated in the Col-0 reference genome (Figure 2
-- figure supplement \hyperref[fig2s6]{6}). These enrichments suggest
+- figure supplement \hyperref[fig2s7]{7}). These enrichments suggest
 that the RC/Helitron TEs have been relatively dormant in recent
@@
 of TE variant enrichment or depletion (Figure 2 - figure supplement
-\hyperref[fig2s7]{7}; source data 4).
+\hyperref[fig2s8]{8}; source data 4).
 
 We further examined Arabidopsis (Col-0) DNA sequencing data from a
-transgenerational stress experiment to investigate the possible
-minimum number of generations required for TE variants to arise
-\cite{Jiang:2014ih}. In one of the three replicates, we identified a
-single potential TE insertion in a sample following 10 generations of
-single-seed descent under high salinity stress conditions, and no TE
-variants in the control single-seed descent set. However, without
-experimental validation it remains unclear if this represents a true
-variant. Therefore, we conclude that TE variants may arise at a rate
-less than 1 insertion in 30 generations under laboratory
-conditions. Further experimental work will be required to precisely
-determine the rate of transposition in Arabidopsis.
+transgenerational stress experiment to investigate the possible minimum
+number of generations required for TE variants to arise
+\cite{Jiang:2014ih}. In one of the three replicates subjected to high
+salinity stress conditions, we identified a single potential TE
+insertion in a sample following 10 generations of single-seed descent,
+while no TE variants were identified in any of the three control
+single-seed descent replicate sets. However, without experimental
+validation it remains unclear if this represents a true variant.
+Therefore, we conclude that TE variants may arise at a rate less than 1
+insertion in 60 generations under laboratory conditions. Further
+experimental work will be required to precisely determine the rate of
+transposition in Arabidopsis.
 
@@
 Although thousands of TE variants were identified, they may be linked
-(i.e. `tagged') by the previously identified single nucleotide
-polymorphisms (SNPs), or unlinked from SNPs across the accessions. We
-tested how often common TE variants (\textgreater{}3 \% MAF; 7
+to the previously identified single nucleotide polymorphisms (SNPs),
+or unlinked from SNPs across the accessions. We tested how frequently
+common TE variants (\textgreater{}3 \% MAF, \textgreater{}7
 accessions) were linked to adjacent SNPs to determine when they would
@@
 of individual TE variants.  For the testable TE variants in the
-population, the nearest 300 flanking SNPs upstream and downstream of
-the TE variant site were analyzed for local linkage disequilibrium
-(LD, r\textsuperscript{2}; see methods). TE variants were classified
-as being either `low', `mid', or `high' LD variants by comparing
-ranked r\textsuperscript{2} values of TE variant to SNPs against the
-median ranked r\textsuperscript{2} value for all between SNP
-comparisons (SNP-SNP) to account for regional variation in the extent
-of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
+population, the nearest 300 flanking SNPs upstream and 300 SNPs
+downstream of the TE variant site were analyzed for local linkage
+disequilibrium (LD, r\textsuperscript{2}; see methods). TE variants
+were classified as being either `low', `mid', or `high' LD variants by
+comparing ranked r\textsuperscript{2} values of TE variant to SNPs
+against the median ranked r\textsuperscript{2} value for all between
+SNP comparisons (SNP-SNP) to account for regional variation in the
+extent of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
 variation or selection \cite{Horton:2012jh}. The majority (61\%) of
 testable TE variants had low LD with nearby SNPs, and represent a
-source of genetic diversity not previously assessed by current
-SNP-based genotype calling methods (Figure \ref{fig3}C). 29\% of TE
-variants displayed high levels of LD and are tagged by nearby SNPs,
-while only 10\% had intermediate levels of LD. We observed a positive
-correlation between TE variant MAF and LD state, with variants of a
-high minor allele frequency more often classified as high-LD (Figure
+source of genetic diversity not previously assessed by SNP-based
+genotype calling methods (Figure \ref{fig3}C).  29\% of TE variants
+displayed high levels of LD and are tagged by nearby SNPs, while only
+10\% had intermediate levels of LD. We observed a positive correlation
+between TE variant MAF and LD state, with variants of a high minor
+allele frequency more often classified as high-LD (Figure
 \ref{fig3}D). While the proportion of TE variants classified as high,
@@
 unable to be classified as either true deletions or as new insertions
-had a much higher proportion of high-LD variants (Figure
-\ref{fig3}E). This was consistent with the observation that the more
-common alleles were more often in a high-LD state. TE variants
-displayed a similar distribution over chromosome one regardless of
-linkage classification (Figure 3 - figure supplement
-\hyperref[fig3s1]{1}). Overall, this analysis revealed an abundance of
-previously unexplored genetic variation that exists amongst
-Arabidopsis accessions caused by the presence or absence of TEs, and
-illustrates the importance of identifying TE variants alongside other
-genetic diversity such as SNPs.
+had a much higher proportion of high-LD variants (Figure \ref{fig3}E).
+This was consistent with the observation that the more common alleles
+were more often in a high-LD state.  TE variants displayed a similar
+distribution over chromosome 1 regardless of linkage classification
+(Figure 3 - figure supplement \hyperref[fig3s1]{1}). Overall, this
+analysis revealed an abundance of previously uncharacterized genetic
+variation that exists amongst Arabidopsis accessions caused by the
+presence or absence of TEs, and illustrates the importance of
+identifying TE variants alongside other genetic diversity such as
+SNPs.
 
@@
 
-To determine whether the TE variants identified affected nearby gene
-expression, we compared the steady state transcript abundance within
-mature leaf tissue between accessions with and without TE insertions
-or deletions, for genes with TE variants located in the 2 kb gene
-upstream region, 5' UTR, exon, intron, 3' UTR or 2 kb downstream
-region (Figure \ref{fig4}A). While the steady state transcript
-abundance of most genes appeared to be unaffected by the presence of a
-TE, 196 genes displayed significant differences in transcript
-abundance linked with the presence of a TE variant, indicating a role
-for these variants in the local regulation of gene expression (1\%
-false discovery rate; greater than 2-fold change in transcript
-abundance; Figure \ref{fig4}A, Figure 4 - source data 1). We did
-not find any functional category enrichments in this set of
-differentially expressed genes. As rare TE variants with a MAF less
+To determine whether the newly discovered TE variants may affect
+nearby gene expression, the steady state transcript abundance within
+mature leaf tissue was compared between accessions with and without TE
+insertions or deletions, for genes with TE variants located in the 2
+kb gene upstream region, 5' UTR, exons, introns, 3' UTR or 2 kb
+downstream region (Figure \ref{fig4}A). While the steady state
+transcript abundance of most genes appeared to be unaffected by the
+presence of a TE, 196 genes displayed significant differences in
+transcript abundance linked with the presence of a TE variant,
+indicating a role for these variants in the local regulation of gene
+expression (1\% false discovery rate; \textgreater{}2-fold change in
+transcript abundance; Figure \ref{fig4}A, Figure 4 - source data
+1). No functional category enrichments in this set of differentially
+expressed genes were identified. As rare TE variants with a MAF less
 than 3\% may also be associated with a difference in transcript
 abundance, but were unable to be statistically tested due to their
-rarity, we performed a burden test for enrichment of rare variants in
-the extremes of expression \cite{Zhao:2016gc}. Briefly, this method
+rarity, a burden test for enrichment of rare variants in the extremes
+of expression was performed \cite{Zhao:2016gc}. Briefly, this method
 counts the frequency of rare variants within each gene expression rank
@@
 population to determine whether an enrichment of rare variants exists
-within any gene expression rank. We found a strong enrichment for gene
-expression extremes for TE variants in the all gene features tested
-(Figure \ref{fig4}B). While TE variants in gene upstream regions
-showed a strong enrichment of both high and low gene expression ranks,
-TE variants in exons or gene downstream regions seemed to have a
-stronger enrichment for low expression ranks than high
+within any gene expression rank. A strong enrichment for gene
+expression extremes was observed for TE variants in all gene features
+tested (Figure \ref{fig4}B). While TE variants in gene upstream
+regions showed a strong enrichment of both high and low gene
+expression ranks, TE variants in exons or gene downstream regions had
+a stronger enrichment for low expression ranks than high
 ranks. Randomization of the accession names removed these enrichments
@@
 supplement \hyperref[fig4s2]{2}). This rare variant analysis further
-indicated that TE variants may alter the transcript abundance of
+indicates that TE variants may alter the transcript abundance of
 nearby genes.
@@
 nearby gene expression. Furthermore, gene-level transcript abundance
-measurements may fail to identify the potential positional effect of
-TE variants upon transcription. To more closely examine changes in
+measurements may fail to identify potential positional effects of TE
+variants upon transcription. To more closely examine changes in
 transcript abundance associated with TE variants among the accessions,
@@
 that appear to have an impact on transcriptional patterns beyond
-changes in total transcript abundance from a nearby gene. For example,
-the presence of a TE insertion within an exon of \emph{AtRLP18
-}(AT2G15040) was associated with truncation of the transcripts at the
-TE insertion site in accessions possessing the TE variant, as well as
-silencing of a downstream gene encoding a leucine-rich repeat protein
-(AT2G15042) (Figure \ref{fig5}A, B). Both genes had significantly
-lower transcript abundance in accessions containing the TE insertion
-(p \textless{} $5.8\times10^{-10}$, Mann-Whitney U
-test). \emph{AtRLP18 }is reported to be involved in bacterial
+simply a change in total transcript abundance of a nearby gene. For
+example, the presence of a TE insertion within an exon of
+\emph{AtRLP18 }(AT2G15040) was associated with truncation of the
+transcripts at the TE insertion site in accessions possessing the TE
+variant, as well as silencing of a downstream gene encoding a
+leucine-rich repeat protein (AT2G15042) (Figure \ref{fig5}A, B). Both
+genes had significantly lower transcript abundance in accessions
+containing the TE insertion (p \textless{} $5.8 \times10^{-10}$, Mann-Whitney U
+test). \emph{AtRLP18 }has been reported to be involved in bacterial
 resistance, with the disruption of this gene by T-DNA insertion
@@
 bacterial plant pathogen \emph{Pseudomonas syringae}
-\cite{Wang:2008km}. We examined pathogen resistance phenotype data
-\cite{Aranzana:2005cq}, and found that accessions containing the TE
+\cite{Wang:2008km}. Examination of pathogen resistance phenotype data
+\cite{Aranzana:2005cq} revealed that accessions containing the TE
 insertion in the \emph{AtRLP18} exon were more often sensitive to
@@
 
-We also observed some TE variants associated with increased expression
-of nearby genes. For example, the presence of a TE within the upstream
+Some TE variants were also associated with increased expression of
+nearby genes. For example, the presence of a TE within the upstream
 region of a gene encoding a pentatricopeptide repeat (PPR) protein
-(AT2G01360) was associated with higher steady state transcript abundance
-of this gene (Figure \ref{fig5}D, E). Transcription appeared to begin at the TE
-insertion point, rather than the transcriptional start site of the gene
-(Figure \ref{fig5}D). Accessions containing the TE insertion had significantly
-higher AT2G01360 transcript abundance than the accessions without the TE
-insertion (p \textless{} $1.8\times10^{-7}$, Mann-Whitney U test). The apparent
-transcriptional activation, linked with presence of a TE belonging to
-the \emph{HELITRON1} family, indicates that this element may carry
-sequences or other regulatory information that has altered the
-expression of genes downstream of the TE insertion site. Importantly,
-this variant was classified as a low-LD TE insertion, as it is not in LD
-with surrounding SNPs, and therefore the associated changes in gene
-transcript abundance would not be identified using only SNP data. This
-TE variant was also upstream of \emph{QPT }(AT2G01350), involved in NAD
-biosynthesis \cite{Katoh:2006if}, which did not show alterations in
-steady state transcript abundance associated with the presence of the TE
-insertion, indicating a potential directionality of regulatory elements
-carried by the TE (Figure \ref{fig5}D, E). Overall, these examples demonstrate
-that TE variants can have unpredictable, yet important, effects on the
-expression of nearby genes, and these effects may be missed by studies
-focused on genetic variation at the level of SNPs.
+(AT2G01360) was associated with higher transcript abundance of this gene
+(Figure \ref{fig5}D, E). Transcription appeared to begin at the TE insertion
+point, rather than the transcriptional start site of the gene (Figure
+\ref{fig5}D). Accessions containing the TE insertion had significantly higher
+AT2G01360 transcript abundance than the accessions without the TE
+insertion (p \textless{} $1.8 \times10^{-7}$, Mann-Whitney U test). The apparent
+transcriptional activation, linked with the presence of a TE belonging
+to the \emph{HELITRON1} family, indicates that this element may carry
+regulatory information that alters the expression of genes downstream of
+the TE insertion site. Importantly, this variant was classified as a
+low-LD TE insertion, as it is not in LD with surrounding SNPs, and
+therefore the associated changes in gene transcript abundance would not
+be linked to genetic differences between the accessions using only SNP
+data. This TE variant was also upstream of \emph{QPT }(AT2G01350),
+involved in NAD biosynthesis \cite{Katoh:2006if}, which did not show
+alterations in transcript abundance associated with the presence of the
+TE insertion, indicating a potential directionality of regulatory
+elements carried by the TE (Figure \ref{fig5}D, E). Overall, these examples
+demonstrate that TE variants can have unpredictable, yet important,
+effects on the expression of nearby genes, and these effects may be
+missed by studies focused on genetic variation at the level of SNPs.
 
@@
 As TEs are frequently highly methylated in Arabidopsis
-\cite{Lister:2008bh, Cokus:2008fc, Zhang:2006eo, Zilberman:2006hx}, we
-next assessed the DNA methylation state surrounding TE variant sites to
-determine whether TE variants might be responsible for some of the
-differences in DNA methylation patterns previously observed between the
-wild accessions \cite{Schmitz:2013iu}. We found that TE variants were
-often physically close to DMRs (Figure \ref{fig6}A). Furthermore, C-DMRs were
-more often close to a TE insertion than expected, while they were rarely
-near a TE deletion (Table \ref{table3}). CG-DMRs were rarely close to TE insertions
-or TE deletions. Overall, we found 48\% of the 13,482 previously
-reported population C-DMRs were located within 1 kb of a TE variant
-(predominantly TE insertions), while only 15\% of CG-DMRs were within 1
-kb of a TE variant (Table \ref{table3}). For C-DMRs, this was significantly more
-than expected by chance, while it was significantly less than expected
-for CG-DMRs (p \textless{} $1\times10^{-4}$, determined by resampling 10,000
-times). To determine if DMR methylation levels were dependent on the
-presence/absence of nearby TE variants, we calculated Pearson
-correlation coefficients between the DNA methylation level at each DMR
-and the presence/absence of the nearest TE variant. We observed a
-negative correlation between the distance from a C-DMR to the nearest TE
-insertion and the correlation between the DNA methylation level at the
-C-DMR with the presence/absence of the TE insertion (Figure \ref{fig6}B). This
-suggests a distance-dependent effect of TE insertion presence on C-DMR
-methylation. In contrast, we found no such relationship for TE deletions
-on C-DMRs, or for insertions or deletions on CG-DMRs (Figure \ref{fig6}B). DNA
-methylation levels at C-DMRs located within 1 kb of a TE insertion
-(TE-DMRs) were more often positively correlated with the
-presence/absence of a TE insertion than the DNA methylation levels at
-C-DMRs further than 1 kb from a TE insertion (non-TE-DMRs). This was
-evident from the distribution of correlations for non-TE-DMRs being
-centred around zero, whereas for TE-DMRs this distribution was skewed to
-the right (Figure \ref{fig6}C, \emph{D}=0.23). For TE deletions, we did not
-observe such a difference in the distributions of correlation
-coefficients between TE-DMRs and non-TE-DMRs, nor for CG-DMRs and their
-nearby TE insertions or deletions (Figure \ref{fig6}C, \emph{D=}0.07-0.10).
-Furthermore, DNA methylation levels were often higher in the presence of
-the nearby TE insertion, while this relationship was generally not
-observed for C-DMRs further than 1 kb from a TE variant, for TE
-deletions, or for CG-DMRs (Figure 6 - figure supplement \hyperref[fig6s1]{1}).
+\cite{Zhang:2006eo, Zilberman:2006hx, Cokus:2008fc, Lister:2008bh},
+the DNA methylation state surrounding TE variant sites was assessed to
+determine whether TE variants might be responsible for differences in
+DNA methylation patterns previously observed between the wild
+accessions \cite{Schmitz:2013iu}. TE variants were often physically
+close to DMRs (Figure \ref{fig6}A). Furthermore, C-DMRs were more
+often close to a TE variant than expected, whereas CG-DMRs were rarely
+close to TE insertions or TE deletions (Table \ref{table3}). Overall,
+54\% of the 13,482 previously reported population C-DMRs were located
+within 1 kb of a TE variant (predominantly TE insertions), while only
+15\% of CG-DMRs were within 1 kb of a TE variant (Table
+\ref{table3}). For C-DMRs, this was significantly more than expected
+by chance, while it was significantly less than expected for CG-DMRs
+(p \textless{} $1 \times10^{-4}$, determined by resampling 10,000 times). Of
+the C-DMRs that were not close to a TE variant, 3,701 (27\% of all
+C-DMRs) were within 1 kb of a non-variable TE. Thus, 81\% of C-DMRs
+are within 1 kb of a TE when considering both fixed and variable TEs
+in the population. Of the remaining 19\% of C-DMRs, most were found in
+genes or intergenic regions.
+
+To determine whether DMR methylation levels were dependent on the
+presence/absence of nearby TE variants, Pearson correlation coefficients
+were calculated between the DNA methylation level at each DMR and the
+presence/absence of the nearest TE variant. A negative correlation was
+observed between the distance from a C-DMR to the nearest TE insertion
+and the correlation between the DNA methylation level at the C-DMR with
+the presence/absence of the TE insertion (Figure \ref{fig6}B). This suggests a
+distance-dependent effect of TE insertion presence on C-DMR methylation.
+In contrast, no such relationship was found for TE deletions on C-DMRs,
+or for insertions or deletions on CG-DMRs (Figure \ref{fig6}B). DNA methylation
+levels at C-DMRs located within 1 kb of a TE insertion (TE-DMRs) were
+more often positively correlated with the presence/absence of a TE
+insertion than the DNA methylation levels at C-DMRs further than 1 kb
+from a TE insertion (non-TE-DMRs). This was evident from the
+distribution of correlations for non-TE-DMRs being centred around zero,
+whereas for TE-DMRs this distribution was skewed to the right (Figure
+\ref{fig6}C, \emph{D}=0.24). For TE deletions, such a difference was not observed
+in the distributions of correlation coefficients between TE-DMRs and
+non-TE-DMRs, nor for CG-DMRs and their nearby TE insertions or deletions
+(Figure \ref{fig6}C, \emph{D=}0.07-0.10). Furthermore, DNA methylation levels
+were often higher in the presence of the nearby TE insertion, while this
+relationship was generally not observed for C-DMRs further than 1 kb
+from a TE variant, for TE deletions, or for CG-DMRs (Figure 6 - figure
+supplement \hyperref[fig6s1]{1}).
 
 As the above correlations between TE presence/absence and DMR
-methylation level rely on the TE variants having a high MAF, this
-precludes an analysis of the effect of rare variants on DMR methylation
-levels. To determine the effect that these rare TE variants may have on
-DMR methylation levels, we performed a burden test for enrichment of DMR
-methylation extremes at TE-DMRs, similar to as was done to test the
-effect of rare variants on gene expression. We found a strong enrichment
-for high C-DMR and CG-DMR methylation level ranks for TE insertions,
-while TE deletions were associated with both high and low extremes of
-DNA methylation levels at C-DMRs, and less so at CG-DMRs (Figure \ref{fig6}D).
-This further indicates that the presence of a TE insertion is associated
-with higher C-DMR methylation levels, while TE deletions seem to have
-more variable effects on DMR methylation levels. This enrichment was
-completely absent after repeating the analysis with randomized accession
-names (Figure 6 - figure supplement \hyperref[fig6s2]{2}). We also observed a slight
-enrichment for low DMR methylation ranks for TE insertions near CG-DMRs,
-indicating that the insertion of a TE was sometimes associated with
-reduced CG methylation in nearby regions (\textless{}1 kb from the TE).
-We examined these TE insertions in a genome browser, and found that some
-TE insertions were associated with decreased transcript abundance of
-nearby genes, with a corresponding loss of gene body methylation,
-offering a potential explanation for the decreased CG methylation
-observed near some TE insertions (Figure 6 - figure supplement \hyperref[fig6s3]{3}).
-
-We next examined levels of DNA methylation in regions flanking all TE
-variants regardless of the presence or absence of a population DMR call.
-While DNA methylation levels around pericentromeric TE insertions and
-deletions (\textless{}3 Mb from a centromere) seemed to be unaffected by
-the presence of a TE insertion (Figure \ref{fig7}A), TE insertions in the
-chromosome arms were associated with an increase in DNA methylation
-levels in all contexts (Figure \ref{fig7}A, B). In contrast, TE deletions in the
+methylation level rely on the TE variants having a sufficiently high
+MAF, this precludes analysis of the effect of rare variants on DMR
+methylation levels. To determine the effect that these rare TE
+variants may have on DMR methylation levels, a burden test for
+enrichment of DMR methylation extremes at TE-DMRs was performed,
+similar to the analysis undertaken to test the effect of rare variants
+on gene expression. A strong enrichment was observed for high C-DMR
+and CG-DMR methylation level ranks for TE insertions, while TE
+deletions were associated with both high and low extremes of DNA
+methylation levels at C-DMRs, and less so at CG-DMRs (Figure
+\ref{fig6}D). This further indicates that the presence of a TE insertion
+is associated with higher C-DMR methylation levels, while TE deletions
+appear to have more variable effects on DMR methylation levels. This
+enrichment was completely absent after repeating the analysis with
+randomized accession names (Figure 6 - figure supplement
+\hyperref[fig6s2]{2}). A slight enrichment was also observed for low
+DMR methylation ranks for TE insertions near CG-DMRs, indicating that
+the insertion of a TE was sometimes associated with reduced CG
+methylation in nearby regions (\textless{}1 kb from the TE). Closer
+examination of these TE insertions revealed that some TE insertions
+were associated with decreased transcript abundance of nearby genes,
+with a corresponding loss of gene body methylation, offering a
+potential explanation for the decreased CG methylation observed near
+some TE insertions (Figure 6 - figure supplement
+\hyperref[fig6s3]{3}).
+
+To further assess the effects of TE variants upon local DNA
+methylation patterns, the levels of methylation were examined in
+regions flanking all TE variants regardless of the presence or absence
+of a population DMR call. While DNA methylation levels around
+pericentromeric TE insertions and deletions (\textless{}3 Mb from a
+centromere) seemed to be unaffected by the presence of a TE insertion
+(Figure \ref{fig7}A), TE insertions in the chromosome arms were
+associated with an increase in DNA methylation levels in all sequence
+contexts (Figure \ref{fig7}A, B). In contrast, TE deletions in the
 chromosome arms did not affect patterns of DNA methylation, as the
-flanking methylation level in all contexts seemed to remain high
-following deletion of the TE (Figure \ref{fig7}A, C). As the change in DNA
-methylation levels around TE variant sites appeared to be restricted to
-regions \textless{}200 bp from the insertion site, we correlated DNA
-methylation levels in 200 bp regions flanking TE variants with the
-presence/absence of TE the variants. DNA methylation levels were often
-positively correlated with the presence of a TE insertion when the
-insertion was distant from a centromere (Figure \ref{fig7}D). TE deletions were
-more variably correlated with local DNA methylation levels, but also
-showed a slight bias towards positive correlations for TE deletions
-distant from the centromeres. As methylome data was available for both
-leaf and bud tissue for 12 accessions, we repeated this analysis
-comparing between tissue types, but did not observe any difference in
-the patterns of methylation surrounding TE variant sites between the two
-tissues (Figure 7 - figure supplement \hyperref[fig7s1]{1}).
+flanking methylation level in all contexts appeared to remain high
+following deletion of the TE (Figure \ref{fig7}A, C). As the change in
+DNA methylation levels around most TE variant sites appeared to be
+restricted to regions \textless{}200 bp from the insertion site, DNA
+methylation levels in 200 bp regions flanking TE variants were
+correlated with the presence/absence of TE variants. DNA methylation
+levels were often positively correlated with the presence of a TE
+insertion when the insertion was distant from a centromere (Figure
+\ref{fig7}D). TE deletions were more variably correlated with local
+DNA methylation levels, but also showed a bias towards positive
+correlations for TE deletions distant from the centromeres. As
+methylome data was available for both leaf and bud tissue for 12
+accessions, this analysis was repeated comparing between tissue types,
+but no differences were observed in the patterns of methylation
+surrounding TE variant sites between the two tissues (Figure 7 -
+figure supplement \hyperref[fig7s1]{1}).
 
 These results indicate that local DNA methylation patterns are
-influenced by the differential TE content between genomes, and that the
-DNA methylation-dependent silencing of TEs may lead to formation of DMRs
-between wild Arabidopsis accessions. TE insertions appear to be
-important in defining local patterns of DNA methylation, while DNA
-methylation levels often remain elevated following a TE deletion, and so
-are independent from the presence or absence of TEs in these cases.
-Importantly, the distance from a TE insertion to the centromere appears
-to have a strong impact on whether an alteration of local DNA
-methylation patterns will occur. This is likely due to flanking
-sequences being highly methylated in the pericentromeric regions, and so
-the insertion of a TE cannot further increase levels of DNA methylation.
-Overall, a large fraction of the population C-DMRs previously identified
-between wild accessions are correlated with the presence of local TE
-insertions, but not TE deletions. CG-DMR methylation levels seem to be
-mostly independent from the presence/absence of common TE variants,
-while rare TE variants have an impact on DNA methylation levels at both
-C-DMRs and CG-DMRs.
+influenced by the differential TE content between genomes, and that
+the DNA methylation-dependent silencing of TEs may frequently lead to
+the formation of DMRs between wild Arabidopsis accessions. TE
+insertions appear to be important in defining local patterns of DNA
+methylation, while DNA methylation levels often remain elevated
+following a TE deletion, and so are independent from the presence or
+absence of TEs in these cases. Importantly, the distance from a TE
+insertion to the centromere appears to have a strong impact on whether
+an alteration of local DNA methylation patterns will occur. This is
+likely due to flanking sequences being highly methylated in the
+pericentromeric regions, and so the insertion of a TE cannot further
+increase levels of DNA methylation. Overall, a large fraction of the
+population C-DMRs previously identified between wild accessions are
+correlated with the presence of local TE variants. CG-DMR methylation
+levels appear to be mostly independent from the presence/absence of
+common TE variants, while rare TE variants have an impact on DNA
+methylation levels at both C-DMRs and CG-DMRs. This analysis was aided
+by the high sensitivity of TEPID to detect TE variants, enabling the
+identification of 8x more variants than has been reported previously
+in the population \cite{Quadrana:2016bi}, allowing a thorough
+assessment of the impact of TE variants on DNA methylation patterns.
 
@@
 
-To better quantify the above results, an association scan was conducted
-for all common TE variants (\textgreater{}3\% MAF) and all population
-C-DMRs for the 124 accessions with both DNA methylation and TE variant
-data available. To test the significance of each pairwise correlation,
-we collected bootstrap p-value estimates based on 500 permutations of
-accession labels. TE-DMR associations were deemed significant if they
-had an association more extreme than any of the 500 permutations (p
-\textless{} 1/500). A band of significant associations was observed for
-TE insertions and their nearby C-DMRs, signifying a local association
-between TE insertion presence/absence and C-DMR methylation (Figure \ref{fig8}A).
-This local association was not as strong for TE deletions (Figure \ref{fig8}B),
-consistent with our above findings. While TE variants and DNA
-methylation showed a local association, it is also possible that TE
-variation may influence DNA methylation state more broadly in the
-genome, perhaps through production of \emph{trans-}acting smRNAs or
-inactivation of genes involved in DNA methylation establishment or
-maintenance. To identify any potential enrichment of C-DMRs regulated in
-\emph{trans}, we summed the total number of significant associations for
-each TE variant across the whole genome (Figure \ref{fig8}A and B, top panels).
-At many sites, we found far more significant associations than expected
+To further investigate the effects of TE variants upon local and
+distant DNA methylation levels in the genome, an association scan was
+conducted for all common TE variants (\textgreater{}3\% MAF) and all
+population C-DMRs for the 124 accessions with both DNA methylation and
+TE variant data available. To test the significance of each pairwise
+correlation, bootstrap p-value estimates were collected based on 500
+permutations of accession labels. TE-DMR associations were deemed
+significant if they had an association more extreme than any of the
+500 permutations (p \textless{} 1/500). A band of significant
+associations was observed for TE insertions and their nearby C-DMRs,
+signifying a local association between TE insertion presence/absence
+and C-DMR methylation (Figure \ref{fig8}A).  This local association was
+not as strong for TE deletions (Figure \ref{fig8}B), consistent with our above
+findings. While TE variants and DNA methylation showed a local
+association, it is also possible that TE variation may influence DNA
+methylation states more broadly in the genome, perhaps through
+production of \emph{trans-}acting smRNAs or inactivation of genes
+involved in DNA methylation establishment or maintenance. To identify
+any potential enrichment of C-DMRs regulated in \emph{trans}, the
+total number of significant associations was summed for each TE
+variant across the whole genome (Figure \ref{fig8}A and B, top panels). At
+many sites, far more significant associations were found than expected
 due to the false positive rate alone. This suggested the existence of
 many putative \emph{trans }associations between TE variants and
-genome-wide C-DMR methylation levels. We further examined these C-DMRs
-that appeared to be associated with a TE insertion in \emph{trans},
-checking for TE insertions near these C-DMRs that were present in the
-same accessions as the \emph{trans }associated TE, as these could lead
-to a false \emph{trans }association. These were extremely rare, with
-only 4 such cases for TE insertions, and 38 cases for TE deletions, and
-so were unable to explain the high degree of \emph{trans }associations
-found.
+genome-wide C-DMR methylation levels. These C-DMRs that appeared to be
+associated with a TE insertion in \emph{trans} were further
+examined\emph{, }checking for TE insertions near these C-DMRs that
+were present in the same accessions as the \emph{trans }associated TE,
+as these could lead to a false \emph{trans }association. These were
+extremely rare, with only 4 such cases for TE insertions, and 38 cases
+for TE deletions, and so were unable to explain the high degree of
+\emph{trans }associations found. Overall, this analysis suggests that
+certain TE variants may affect DNA methylation levels more broadly in
+the genome, as their effects upon DNA methylation are not necessarily
+limited to nearby DNA sequences.
 
@@
 
-Here we discovered widespread differential TE content between wild
-Arabidopsis accessions, and explore the impact of these variants at the
-level of individual accessions. Most TE variants were due to the
-\emph{de novo }insertion of TEs, while a smaller subset was likely due
-to the deletion of ancestral TE copies, mostly around the
-pericentromeric regions. A subset (32\%) of TE variants with a minor
-allele frequency above 3\% were able to be tested for linkage with
-nearby SNPs. The majority of these TE variants were not in LD with SNPs,
-indicating that they represent genetic variants currently overlooked in
-genomic studies. We found a marked depletion of TE variants within gene
-bodies and DNase I hypersensitivity sites (putative regulatory regions),
-consistent with the more deleterious TE insertions have being removed
-from this population through selection. Of those TE variants found in
-gene bodies, TE deletions were overrepresented, indicating that the loss
-of ancestral TEs inserted within genes may be more frequent, or perhaps
-less deleterious, than the \emph{de novo }insertion of TEs into genes.
-
-The identification of a large number of TE variants in this population
-gave an opportunity to form statistically robust correlations between TE
-presence/absence and transcript abundance from nearby genes, as well as
-genome-wide patterns of DNA methylation. We were able to identify
-examples where TE variants appear to have an effect upon gene
-expression, both in the disruption of transcription and in the spreading
-or disruption of regulatory information leading to the transcriptional
-activation of genes, indicating that these TE variants can have
-important consequences upon the expression of protein coding genes
-(Figure \ref{fig5}). In one case, these changes in gene expression could be
-linked with phenotypic changes, with accessions containing a TE
-insertion more frequently sensitive to bacterial infection. Further
-experiments will be needed to establish a causal link between this TE
-insertion and the associated phenotype. An analysis of rare TE variants,
-present at a low MAF, further strengthened this relationship between TE
-presence/absence and altered transcript abundance, as we were able to
-identify a strong enrichment of rare TE variants in accessions with
-extreme gene expression ranks in the population.
-
-Perhaps most importantly, we provide evidence that differential TE
-content between genomes of Arabidopsis accessions underlies a large
-fraction of the previously reported population C-DMRs, in agreement with
-recent similar findings \cite{Quadrana:2016bi}. Thus, the frequency of
-pure epialleles, independent of underlying genetic variation, may be
-even more rare than previously anticipated \cite{Richards:2006br}. We
-did not find evidence of CG-DMR methylation level being altered by the
-presence of common TE variants, but rather rare TE variants may be more
+Here we have discovered widespread differential TE content between wild
+Arabidopsis accessions, and explored the impact of these variants upon
+transcription and DNA methylation at the level of individual accessions.
+Most TE variants were due to the \emph{de novo }insertion of TEs, while
+a smaller subset was likely due to the deletion of ancestral TE copies,
+mostly around the pericentromeric regions. A subset (32\%) of TE
+variants with a minor allele frequency above 3\% were able to be tested
+for linkage with nearby SNPs. The majority of these TE variants were not
+in LD with SNPs, indicating that they represent genetic variants
+currently overlooked in genomic studies. A marked depletion of TE
+variants within gene bodies and DNase I hypersensitivity sites (putative
+regulatory regions) is consistent with the more deleterious TE
+insertions being removed from the population through selection. Of those
+TE variants found in gene bodies, TE deletions were overrepresented,
+indicating that the loss of ancestral TEs inserted within genes may be
+more frequent, or perhaps less deleterious, than the \emph{de novo
+}insertion of TEs into genes.
+
+The high sensitivity of our TE variant detection method allowed
+identification of 8x more variants than has been reported previously
+\cite{Quadrana:2016bi}, and the identification of such a large number of
+TE variants (23,095) in this population gave an opportunity to form
+statistically robust correlations between TE presence/absence and
+transcript abundance from nearby genes, as well as genome-wide patterns
+of DNA methylation. Examples were identified where TE variants appear to
+have an effect upon gene expression, both in the disruption of
+transcription and in the spreading or disruption of regulatory
+information leading to the transcriptional activation of genes,
+indicating that these TE variants can have important consequences upon
+the expression of protein coding genes (Figure \ref{fig5}). In one case, these
+changes in gene expression could be linked with phenotypic changes, with
+accessions containing a TE insertion more frequently sensitive to
+bacterial infection. Further experiments will be needed to establish a
+causal link between this TE insertion and the associated phenotype. An
+analysis of rare TE variants, present at a low MAF, further strengthened
+this relationship between TE presence/absence and altered transcript
+abundance, as a strong enrichment of rare TE variants in accessions with
+extreme gene expression ranks in the population was identified.
+
+Importantly, we provide evidence that differential TE content between
+genomes of Arabidopsis accessions underlies a large fraction of the
+previously reported population C-DMRs. Thus, the frequency of pure
+epialleles, independent of underlying genetic variation, may be even
+more rare than previously anticipated \cite{Richards:2006br}. Overall,
+81\% of all C-DMRs were within 1 kb of a TE, when considering both fixed
+and variable TEs in the population, a much higher percentage than has
+been reported previously \cite{Schmitz:2013iu, Quadrana:2016bi}. We did
+not find evidence of CG-DMR methylation level being altered by the
+presence of common TE variants, although rare TE variants may be more
 important in shaping patterns of DNA methylation at some CG-DMRs, though
@@
 TE deletions in the effect that these variants have on nearby DNA
-methylation levels. While flanking DNA methylation levels appeared to
-increase following a TE insertion, the deletion of an ancestral TE was
-often not associated with a corresponding decrease in flanking DNA
-methylation levels (Figure \ref{fig7}). This indicates that high levels of DNA
-methylation, once established, may be maintained in the absence of the
-TE insertion that presumably triggered the original change in DNA
-methylation level. It is then possible that TE variants explain more of
-the variation in DNA methylation patterns than we find direct evidence
-for, if some C-DMRs were formed by the insertion of an ancestral TE that
-is now absent in all the accessions analysed here. These DMRs would then
+methylation levels. While flanking DNA methylation levels increase
+following a TE insertion, the deletion of an ancestral TE was often not
+associated with a corresponding decrease in flanking DNA methylation
+levels (Figure \ref{fig7}). This indicates that high levels of DNA methylation,
+once established, may be maintained in the absence of the TE insertion
+that presumably triggered the original change in DNA methylation level.
+It is then possible that TE variants explain more of the inter-accession
+variation in DNA methylation patterns than we find direct evidence for,
+if some C-DMRs were formed by the insertion of an ancestral TE that is
+now absent in all the accessions analysed here. These DMRs would then
 represent the epigenetic ``scars'' of past TE insertions.
 
-Finally, we performed a genome-wide scan of common TE variant
-association with C-DMR methylation levels, and found further evidence of
-a strong local association between TE insertion presence/absence and
-C-DMR methylation level (Figure \ref{fig8}). We were also able to identify some
-TE variants that appeared to be associated with changes in DNA
-methylation levels at multiple loci throughout the genome, indicating a
-possible \emph{trans} regulation of DNA methylation levels linked to
-certain TE variants. Further experiments will be required to confirm and
-examine the role of these TE variants in determining genome-wide
-patterns of DNA methylation. Overall our results show that TE
-presence/absence variants between wild Arabidopsis accessions not only
-have important effects on nearby gene expression, but can also have a
-role in determining local patterns of DNA methylation, and explain many
-regions of differential DNA methylation previously observed in the
-population.
+Finally, a genome-wide scan of common TE variant association with C-DMR
+methylation levels provides further evidence of a strong local
+association between TE insertion presence/absence and C-DMR methylation
+level (Figure \ref{fig8}). The identification of some TE variants that appeared
+to be associated with changes in DNA methylation levels at multiple loci
+throughout the genome indicates possible \emph{trans} regulation of DNA
+methylation state linked to specific TE variants. Further experiments
+will be required to confirm and examine the role of these TE variants in
+determining genome-wide patterns of DNA methylation. Overall, our
+results show that TE presence/absence variants between wild Arabidopsis
+accessions not only have important effects on nearby gene expression,
+but can also have a role in determining local patterns of DNA
+methylation, and explain many regions of differential DNA methylation
+previously observed in the population.
 
@@
 variants not related to TE diversity (Figure 2 - figure supplement
-\hyperref{fig2s6}). Col-0 reference annotation TEs that are located
+\hyperref[fig2s7]{7}). Col-0 reference annotation TEs that are located
 within the genomic region spanned by the split and discordant reads
@@
 minimum number of split and discordant reads present, determined by
-the sequencing depth (1/10 coverage; Figure \ref{fig1}). A threshold
-of 80\% TE sequence spanned by split or discordant reads is used, as
-opposed to 100\%, to account for misannotation of TE sequence
-boundaries in the Col-0 reference TE annotation, as well as TE
-fragments left behind by DNA TEs during cut-paste transposition (TE
-footprints) that may affect the mapping of reads around annotated TE
-borders \cite{Plasterk:1991vd}. Furthermore, the coverage within the
-spanned region may be more than 10\% that of the flanking sequence,
-but in such cases twice as many split and discordant reads are
-required. If multiple TEs are spanned by the split and discordant
-reads, and the above requirements are met, multiple TEs in the same
-region can be identified as absent with respect to the reference
-genome. Absence variants in non-Col-0 accessions are subsequently
-recategorized as TE insertions present in the Col-0 genome but absent
-from a given wild accession.
+the sequencing depth (1/10 coverage; Figure \ref{fig1}). A threshold of 80\%
+TE sequence spanned by split or discordant reads is used, as opposed
+to 100\%, to account for misannotation of TE sequence boundaries in
+the Col-0 reference TE annotation, as well as TE fragments left behind
+by DNA TEs during cut-paste transposition (TE footprints) that may
+affect the mapping of reads around annotated TE borders
+\cite{Plasterk:1991vd}. Furthermore, the coverage within the spanned
+region may be more than 10\% that of the flanking sequence, but in
+such cases twice as many split and discordant reads are required. If
+multiple TEs are spanned by the split and discordant reads, and the
+above requirements are met, multiple TEs in the same region can be
+identified as absent with respect to the reference genome. Absence
+variants in non-Col-0 accessions are subsequently recategorized as TE
+insertions present in the Col-0 genome but absent from a given wild
+accession.
 
@@
 Previously published 100 bp paired end sequencing data for L\emph{er}
-(\url{http://1001genomes.org/data/MPI/MPISchneeberger2011/releases/current/Ler-1/Reads/}
-; \cite{Schneeberger:2011ft}) was downloaded and analyzed with the TEPID
+(\url{http://1001genomes.org/data/MPI/MPISchneeberger2011/releases/current/Ler-1/Reads/};
+\cite{Schneeberger:2011ft}) was downloaded and analyzed with the TEPID
 package to identify TE variants. Reads providing evidence for TE
-variants were then mapped to the \emph{de novo} assembled L\emph{er}
-genome \cite{Chin:2013iw}. To determine whether reads mapped to
+variants were then mapped to the \emph{de novo} assembled L\emph{er
+}genome \cite{Chin:2013iw}. To determine whether reads mapped to
 homologous regions of the L\emph{er} and Col-0 reference genome, the
@@
 using RepeatMasker with RepBase-derived libraries and the default
-parameters (version 4.0.5, \url{http://www.repeatmasker.org}). A blastn
-search was then conducted against the Col-0 genome using the following
-parameters: `-max-target-seqs 1', `-evalue 1e-6' \cite{Camacho:2009fc}.
-Coordinates of the top blast hit for each read location were then
-compared with the TE variant sites identified using those reads. To
-estimate false negative rates for TEPID TE absence calls, L\emph{er} TE
-absence calls were compared with a known set of Col-0-specific TE
-insertions, absent in L\emph{er }\cite{Quadrana:2016bi}. For TEPID TE
-insertion calls, we mapped Col-0 DNA sequencing reads
-\cite{Jiang:2014ih} to the L\emph{er }PacBio assembly, and identified
-sites with read evidence reaching the TEPID threshold for a TE insertion
-call to be made.
+parameters (version 4.0.5, \url{http://www.repeatmasker.org}). A
+blastn search was then conducted against the Col-0 genome using the
+following parameters: `-max-target-seqs 1', `-evalue 1e-6'
+\cite{Camacho:2009fc}.  Coordinates of the top blast hit for each read
+location were then compared with the TE variant sites identified using
+those reads. To estimate false negative rates for TEPID TE absence
+calls, L\emph{er} TE absence calls were compared with a known set of
+Col-0-specific TE insertions, absent in L\emph{er
+}\cite{Quadrana:2016bi}. For TEPID TE insertion calls, we mapped Col-0
+DNA sequencing reads \cite{Jiang:2014ih} to the L\emph{er }PacBio
+assembly, and identified sites with read evidence reaching the TEPID
+threshold for a TE insertion call to be made.
 
@@
 reference genome and using the TAIR9 TE annotation. The `--mask' option
-was set to mask the mitochondrial and plastid genomes. We also ran
-TEPID using previously published transgenerational data for salt stress
-and control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), again using
-the `--mask' option to mask mitochondrial and plastid genomes, and the
-`--strict' option for highly related samples.
+was set to mask the mitochondrial and plastid genomes. We also ran TEPID
+using previously published transgenerational data for salt stress and
+control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), again
+using the `--mask' option to mask mitochondrial and plastid genomes, and
+the `--strict' option for highly related samples.
 
@@
 1001 genomes data center
-(\url{http://1001genomes.org/data/Salk/releases/2013_24_01/}; \cite{Schmitz:2013iu}).
-This was formatted into reference (Col-0 state), alternate,
-or NA calls for each SNP. Accessions with both TE variant information
-and SNP data were selected for analysis. Hierarchical clustering of
-accessions by SNPs as well as TE variants were used to identify
-essentially clonal accessions, as these would skew minor allele
-frequency calculations. A single representative from each cluster of
-similar accessions was kept, leading to a total of 187 accessions for
-comparison. For each TE variant with a minor allele frequency greater than
-3\%, the nearest 300 upstream and 300 downstream SNPs with a minor
-allele frequency greater than 3\% were selected. Pairwise genotype
-correlations (r\textsuperscript{2} values) for all complete cases were obtained for
-SNP-SNP and SNP-TE variant states. r\textsuperscript{2} values were then ordered by
-decreasing rank and a median SNP-SNP rank value was calculated. For each
-of the 600 ranked surrounding positions, the number of times the TE rank
-was greater than the SNP-SNP median rank was calculated as a relative LD
-metric of TE to SNP. TE variants with less than 200 ranks over the
-SNP-SNP median were classified as low-LD insertions. TE variants with
-ranks between 200 and 400 were classified as mid-LD, while TE variants
-with greater than 400 ranks above their respective SNP-SNP median value
-were classified as variants in high LD with flanking SNPs.
+(\url{http://1001genomes.org/data/Salk/releases/2013_24_01/};
+\cite{Schmitz:2013iu}). This was formatted into reference (Col-0
+state), alternate, or NA calls for each SNP. Accessions with both TE
+variant information and SNP data were selected for
+analysis. Hierarchical clustering of accessions by SNPs as well as TE
+variants were used to identify essentially clonal accessions, as these
+would skew minor allele frequency calculations. A single
+representative from each cluster of similar accessions was kept,
+leading to a total of 187 accessions for comparison. For each TE
+variant with a minor allele frequency greater than 3\%, the nearest
+300 upstream and 300 downstream SNPs with a minor allele frequency
+greater than 3\% were selected. Pairwise genotype correlations
+(r\textsuperscript{2} values) for all complete cases were obtained for
+SNP-SNP and SNP-TE variant states. r\textsuperscript{2} values were
+then ordered by decreasing rank and a median SNP-SNP rank value was
+calculated. For each of the 600 ranked surrounding positions, the
+number of times the TE rank was greater than the SNP-SNP median rank
+was calculated as a relative LD metric of TE to SNP. TE variants with
+less than 200 ranks over the SNP-SNP median were classified as low-LD
+insertions. TE variants with ranks between 200 and 400 were classified
+as mid-LD, while TE variants with greater than 400 ranks above their
+respective SNP-SNP median value were classified as variants in high LD
+with flanking SNPs.
 
@@
 sequence predicted to be absent, and the other primer annealing in the
-flanking sequence (Figure 2 - figure supplement \hyperref[fig2s3]{3}). Primer
-sequences were designed that did not anneal to regions of the genome
-containing previously identified SNPs in any of the 216 accessions
-\cite{Schmitz:2013iu} or small insertions and deletions, identified
-using lumpy-sv with the default settings
+flanking sequence (Figure 2 - figure supplement
+\hyperref[fig2s3]{3}). Primer sequences were designed that did not
+anneal to regions of the genome containing previously identified SNPs
+in any of the 216 accessions \cite{Schmitz:2013iu} or small insertions
+and deletions, identified using lumpy-sv with the default settings
 \cite{Layer:2014ie}(\url{https://github.com/arq5x/lumpy-sv}), had an
@@
 accessions in the population, and tallied the ranks of accessions
-containing a rare variant. These rank counts were then binned to produce
-a histogram of the distribution of ranks. We then fit a quadratic model
-to the counts data, and calculated the R\textsuperscript{2} and p-value for the fit of
-the model.
+containing a rare variant. These rank counts were then binned to
+produce a histogram of the distribution of ranks. We then fit a
+quadratic model to the counts data, and calculated the
+R\textsuperscript{2} and p-value for the fit of the model.
 
@@
 reproduce this analysis can be found at
-\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er
-}TE variants are available in Figure 1 - source data 1 and 2. TE
-variants identified among the 216 wild Arabidopsis accessions
-resequenced by Schmitz et al. (2013) are available in Figure 2 -
-source data 1, 2 and 3. Source data is available on Dryad
+\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er }TE
+variants are available in Figure 1 - source data 1 and 2. TE variants
+identified among the 216 wild Arabidopsis accessions resequenced by
+Schmitz et al. (2013) are available in Figure 2 - source data 1, 2 and
+3. Source data is available on Dryad
 (\url{http://dx.doi.org/10.5061/dryad.187b3}).
@@
 \printbibliography
-
-%\bibliography{references}{}
-%\bibliographystyle{ieeetr}
-
-% Figure 1
 
@@
 
-Number of additional TE insertion calls made due to the TEPID refinement
-step for each accession in the population.
+Percentage of total TE insertion calls that were made due to the TEPID
+refinement step for each accession in the population.
 
@@
 % Figure 2 - figure supplement 5
-
 \setcounter{suppfigure}{1}
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{Figure2/Figure2-figure_supplement_5.png}
-  \caption{figure supplement 5. Frequency of TE insertion in the \emph{KNOT} region}
+  \includegraphics[width=0.5\textwidth]{Figure2/Figure2-figure_supplement_5.png}
+  \caption{figure supplement 5}
   \label{fig2s5}
+\end{suppfigure}
+
+Distribution of Type I and Type II elements over chromosome 1, for TE insertions and TE deletions.
+
+\pagebreak
+% Figure 2 - figure supplement 6
+
+\setcounter{suppfigure}{1}
+
+\begin{suppfigure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{Figure2/Figure2-figure_supplement_6.png}
+  \caption{figure supplement 6. Frequency of TE insertion in the \emph{KNOT} region}
+  \label{fig2s6}
 \end{suppfigure}
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{Figure2/Figure2-figure_supplement_6.png}
-  \caption{figure supplement 6. Length distribution for all Col-0 TEs and all TE variants}
-  \label{fig2s6}
+  \includegraphics[width=0.7\textwidth]{Figure2/Figure2-figure_supplement_7.png}
+  \caption{figure supplement 7. Length distribution for all Col-0 TEs and all TE variants}
+  \label{fig2s7}
 \end{suppfigure}
@@
   \centering
-  \includegraphics[width=1\textwidth]{Figure2/Figure2-figure_supplement_7.png}
-  \caption{figure supplement 7}
-  \label{fig2s7}
+  \includegraphics[width=1\textwidth]{Figure2/Figure2-figure_supplement_8.png}
+  \caption{figure supplement 8}
+  \label{fig2s8}
 \end{suppfigure}
@@
     \textbf{}              & \textbf{Observed} & \textbf{Expected} & \textbf{95\% CI} & \textbf{Observed} & \textbf{Expected} & \textbf{95\% CI} \\ \midrule
-    \textbf{TE deletions}  & 8.7               & 16                & 0.0078           & 4.3               & 16                & 0.0041           \\
-    \textbf{TE insertions} & 36                & 26                & 0.0089           & 9.4               & 26                & 0.0047           \\
-    \textbf{NA calls}      & 3.4               & 6.2               & 0.0052           & 1.7               & 6.2               & 0.0027           \\
-    \textbf{Total}         & 48                & 41                & 0.01             & 15                & 42                & 0.0054           \\ \bottomrule
+    \textbf{TE deletions}  & 17               & 16                & 0.0079           & 4.1               & 16                & 0.0041           \\
+    \textbf{TE insertions} & 28                & 26                & 0.0089           & 9.1               & 26                & 0.0047           \\
+    \textbf{NA calls}      & 8.7               & 6.2               & 0.0053           & 1.6               & 6.2               & 0.0027           \\
+    \textbf{Total}         & 54                & 48                & 0.01             & 15                & 48                & 0.0054           \\ \bottomrule
   \end{tabular}

### 2016-09-02 Paper/paper.tex
Updates to text

@@
 
-Variation in the presence or absence of transposable elements (TEs) is a
-major source of genetic variation between individuals. Here, we
+Variation in the presence or absence of transposable elements (TEs) is
+a major source of genetic variation between individuals. Here, we
 identified 23,095 TE presence/absence variants between 216 Arabidopsis
@@
 identified, two thirds were not in linkage disequilibrium with nearby
-SNPs, implicating these variants as a source of novel genetic diversity.
-Nearly 200 common TE variants were associated with significantly altered
-expression of nearby genes, and a major fraction of inter-accession DNA
-methylation differences were associated with nearby TE insertions.
-Overall, this demonstrates that TE variants are a rich source of genetic
-diversity that likely plays an important role in facilitating epigenomic
-and transcriptional differences between individuals, and indicates a
-strong genetic basis for epigenetic variation.
-
-\pagebreak
+SNPs, implicating these variants as a source of novel genetic
+diversity.  Nearly 200 common TE variants were associated with
+significantly altered expression of nearby genes, and a major fraction
+of inter-accession DNA methylation differences were associated with
+nearby TE insertions.  Overall, this demonstrates that TE variants are
+a rich source of genetic diversity that likely plays an important role
+in facilitating epigenomic and transcriptional differences between
+individuals, and indicates a strong genetic basis for epigenetic
+variation.
 
@@
 nearly all studied organisms, and comprise a large fraction of most
-eukaryotic genomes. The two classes of TEs are retrotransposons (type I
+eukaryotic genomes. The two types of TEs are retrotransposons (type I
 elements), which transpose via an RNA intermediate requiring a reverse
 transcription reaction, and DNA transposons (type II elements), which
-transpose via either a cut-paste or, in the case of Helitrons, a rolling
-circle mechanism with no RNA intermediate \cite{Wicker:2007en}. TE
-activity poses mutagenic potential as a TE insertion may disrupt
-functional regions of the genome. Consequently, safeguard mechanisms
-have evolved to suppress this activity, including the methylation of
-cytosine nucleotides (DNA methylation) to produce 5-methylcytosine (mC),
-a modification that can induce transcriptional silencing of the
-methylated locus. In \emph{Arabidopsis thaliana }(Arabidopsis), DNA
-methylation occurs in all three DNA sequence contexts: mCG, mCHG, and
-mCHH, where H is any base but G. Establishment of DNA methylation marks
-can be carried out by two distinct pathways -- the RNA-directed DNA
-methylation pathway guided by 24 nucleotide (nt) small RNAs (smRNAs),
-and the DDM1/CMT2 pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major
-function of DNA methylation in Arabidopsis is in the transcriptional
-silencing of TEs. Loss of DNA methylation due to mutations in genes
-essential for its establishment or maintenance leads to expression of
-previously silent TEs, and in some cases transposition
-\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu, Zemach:2013dj}.
+transpose via either a cut-paste or, in the case of Helitrons, a
+rolling circle mechanism with no RNA intermediate
+\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
+insertion may disrupt functional regions of the genome. Consequently,
+safeguard mechanisms have evolved to suppress this activity, including
+the methylation of cytosine nucleotides (DNA methylation) to produce
+5-methylcytosine (mC), a modification that can induce transcriptional
+silencing of the methylated locus. In \emph{Arabidopsis thaliana}
+(Arabidopsis), DNA methylation occurs in all three DNA sequence
+contexts: mCG, mCHG, and mCHH, where H is any base but
+G. Establishment of DNA methylation marks can be carried out by two
+distinct pathways -- the RNA-directed DNA methylation pathway guided
+by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2 pathway
+\cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
+methylation in Arabidopsis is in the transcriptional silencing of
+TEs. Loss of DNA methylation due to mutations in genes essential for
+its establishment or maintenance leads to expression of previously
+silent TEs, and in some cases transposition \cite{Mirouze:2009km,
+  Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu,
+  Zemach:2013dj}.
 
@@
 
-The tools developed previously for identification of novel TE insertion
-events have several limitations. They either require a library of active
-TE sequences, cannot identify TE absence variants, are not designed with
-population studies in mind, or suffer from a high degree of
-false-negatives
+The tools developed previously for identification of novel TE
+insertion events have several limitations. They either require a
+library of active TE sequences, cannot identify TE absence variants,
+are not designed with population studies in mind, or specifically aim
+to reduce false-positives rather than false-negatives
 \cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
 order to accurately map the locations of TE presence/absence variants
-with respect to a reference genome, we have developed a novel algorithm,
-TEPID (Transposable Element Polymorphism IDentification), which is
-designed for population studies. We tested our algorithm using both
-simulated and real Arabidopsis sequencing data, finding that TEPID is
-able to accurately identify TE presence/absence variants with respect to
-the Col-0 reference genome. We applied our TE variant identification
-method to existing genome resequencing data for 216 different
-Arabidopsis accessions \cite{Schmitz:2013iu}, identifying widespread TE
-variation amongst these accessions and enabling exploration of TE
-diversity and links to gene regulation and epigenomic variation.
+with respect to a reference genome, we have developed a novel
+algorithm, TEPID (Transposable Element Polymorphism IDentification),
+which is designed for population studies. We tested our algorithm
+using both simulated and real Arabidopsis sequencing data, finding
+that TEPID is able to accurately identify TE presence/absence variants
+with respect to the Col-0 reference genome. We applied our TE variant
+identification method to existing genome resequencing data for 216
+different Arabidopsis accessions \cite{Schmitz:2013iu}, identifying
+widespread TE variation amongst these accessions and enabling
+exploration of TE diversity and links to gene regulation and
+epigenomic variation.
 
@@
 
-In order to further assess the sensitivity and specificity of TE variant
-discovery using TEPID, we identified TE variants in the Landsberg
-\emph{erecta} (L\emph{er}) accession, and compared these with the
-L\emph{er }genome assembly created using long PacBio sequencing reads
-\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
-genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
-using TEPID, enabling identification of 446 TE insertions (Figure 1 -
-source data 1) and 758 TE absence variants (Figure 1 - source data 2)
-with respect to the Col-0 reference TE annotation. Reads providing
-evidence for these variants were then mapped to the L\emph{er} reference
-genome, generated by \emph{de novo} assembly using Pacific Biosciences
-P5-C3 chemistry with a 20 kb insert library \cite{Chin:2013iw}, using
-the same alignment parameters as were used to map reads to the Col-0
+In order to further assess the sensitivity and specificity of TE
+variant discovery using TEPID, we identified TE variants in the
+Landsberg \emph{erecta} (L\emph{er}) accession, and compared these
+with the L\emph{er }genome assembly created using long PacBio
+sequencing reads \cite{Chin:2013iw}. Previously published 100 bp
+paired-end L\emph{er} genome resequencing reads
+\cite{Schneeberger:2011ft} were first analyzed using TEPID, enabling
+identification of 446 TE insertions (Figure 1 - source data 1) and 758
+TE absence variants (Figure 1 - source data 2) with respect to the
+Col-0 reference TE annotation. Reads providing evidence for these
+variants were then mapped to the L\emph{er} reference genome,
+generated by \emph{de novo} assembly using Pacific Biosciences P5-C3
+chemistry with a 20 kb insert library \cite{Chin:2013iw}, using the
+same alignment parameters as were used to map reads to the Col-0
 reference genome. This resulted in 98.7\% of reads being aligned
 concordantly to the L\emph{er} reference, whereas 100\% aligned
-discordantly or as split reads to the Col-0 reference genome (Table \ref{table1}).
-To find whether reads mapped to homologous regions in both the Col-0 and
-L\emph{er} reference genomes, we conducted a blast search
+discordantly or as split reads to the Col-0 reference genome (Table
+\ref{table1}). To find whether reads mapped to homologous regions in both the
+Col-0 and L\emph{er} reference genomes, we conducted a blast search
 \cite{Camacho:2009fc} using the DNA sequence between read pair mapping
-locations in the L\emph{er }genome against the Col-0 genome, and found
+locations in the L\emph{er} genome against the Col-0 genome, and found
 the top blast result for 80\% of reads providing evidence for TE
-insertions, and 89\% of reads providing evidence for TE absence variants
-in L\emph{er}, to be located within 200 bp of the TE variant
+insertions, and 89\% of reads providing evidence for TE absence
+variants in L\emph{er}, to be located within 200 bp of the TE variant
 reported by TEPID. Thus, reads providing evidence for TE variants map
 discordantly or as split reads when mapped to the Col-0 reference
-genome, but map concordantly to homologous regions of the L\emph{er} \emph{de
-novo} assembled reference genome, indicating that structural variation
-is present at the sites identified by TEPID, and that this is resolved
-in the \emph{de novo} assembled genome.
+genome, but map concordantly to homologous regions of the L\emph{er}
+\emph{de novo} assembled reference genome, indicating that structural
+variation is present at the sites identified by TEPID, and that this
+is resolved in the \emph{de novo} assembled genome.
 
@@
 for a TE insertion call to be made at 89.6\% (173/193) of these sites,
-indicating a false negative rate of \textasciitilde{}10\%. However, it should be noted
-that this estimate does not take into account the TEPID refinement step
-used on large populations, and so the false negative rate for samples
-analyzed in the population from Schmitz et al. (2013) is likely to be
-lower than this estimate, as each accession gained on average 4\% more
-insertion calls following this refinement step (Figure 2 - figure
-supplement \hyperref[fig2s1]{1}).
+indicating a false negative rate of \textasciitilde{}10\%. However, it
+should be noted that this estimate does not take into account the TEPID
+refinement step used on large populations, and so the false negative
+rate for samples analyzed in the population from Schmitz et al. (2013)
+is likely to be lower than this estimate, as each accession gained on
+average 4\% more insertion calls following this refinement step (Figure
+2 - figure supplement \hyperref[fig2s1]{1}).
 
@@
 2) relative to the Col-0 reference accession, totalling 23,095 unique
-TE variants. In most accessions TEPID identified 300-500 TE insertions
-(mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
-majority of which were shared by two or more accessions (Figure 2 -
-figure supplement \hyperref[fig2s2]{2}).  PCR validations were
-performed for a random subset of 10 insertion and 10 absence variants
-in 14 accessions (totalling 280 validations), confirming the high
-accuracy of TE variant discovery using the TEPID package, with a false
-positive rate for both TE insertion and TE absence identification of
+TE variants.  A recent study focused on identifying recent TE
+insertions containing target site duplications in this population
+\cite{Quadrana:2016bi}.  Our goal was to provide a comprehensive
+assessment of TE presence/absence variation in Arabidopsis.  In most
+accessions TEPID identified 300-500 TE insertions (mean = 378) and
+1,000-1,500 TE absence variants (mean = 1,279), the majority of which
+were shared by two or more accessions (Figure 2 - figure supplement
+\hyperref[fig2s2]{2}).  PCR validations were performed for a random
+subset of 10 insertion and 10 absence variants in 14 accessions
+(totalling 280 validations), confirming the high accuracy of TE
+variant discovery using the TEPID package, with a false positive rate
+for both TE insertion and TE absence identification of
 \textasciitilde{}9\%, similar to that observed using simulated data
-and the L\emph{er} genome analysis (Figure 2 - figure supplement
+and the Ler genome analysis (Figure 2 - figure supplement
 \hyperref[fig2s3]{3}).  The number of TE insertions identified was
@@
 under the expectation that the minor allele is the derived allele.
-Common TE absences relative to Col-0 were re-classified as TE insertions
-in Col-0, and common TE insertions relative to Col-0 as true TE
-deletions in Col-0. Cases where the TE variant had a high MAF
+Common TE absences relative to Col-0 were re-classified as TE
+insertions in Col-0, and common TE insertions relative to Col-0 as
+true TE deletions in Col-0. Cases where the TE variant had a high MAF
 (\textgreater{}20\%) were assigned NA calls, as it could not be
-determined if these were cases where the variant was most likely to be a
-true TE deletion or a new TE insertion. While these classifications are
-not definitive, as there will be rare cases where a true TE deletion has
-spread through the population and becomes the common allele, it will
-correctly classify most TE variants. Overall, 72.3\% of the TE absence
-variants identified with respect to the Col-0 reference genome were
-likely due to a true TE deletion in these accessions, while 4.8\% were
-due to insertions in Col-0 not shared by other accessions in the
-population (Table \ref{table2}). Overall, we identified 15,077 TE insertions, 5,856
-true TE deletions, and 2,162 TE variants at a high MAF that were unable
-to be classified as an insertion or deletion (Figure 2 - source data 3).
+determined if these were cases where the variant was most likely to be
+a true TE deletion or a new TE insertion. While these classifications
+are not definitive, as there will be rare cases where a true TE
+deletion has spread through the population and becomes the common
+allele, it will correctly classify most TE variants. Overall, 72.3\%
+of the TE absence variants identified with respect to the Col-0
+reference genome were likely due to a true TE deletion in these
+accessions, while 4.8\% were due to insertions in Col-0 not shared by
+other accessions in the population (Table \ref{table2}). Overall, we
+identified 15,077 TE insertions, 5,856 true TE deletions, and 2,162 TE
+variants at a high MAF that were unable to be classified as an
+insertion or deletion (Figure 2 - source data 3).
 
@@
 \ref{fig2}B). TE deletions and common TE variants were enriched within
-the set of TE variants found in gene bodies (Figure 2C, D). No
-significant enrichment was found for TE variants within the \emph{KNOT
-  ENGAGED ELEMENT (KEE) }regions, previously identified as regions
-that may act as a ``TE sink'' \cite{Grob:2014kg} (Figure 2 - figure
-supplement \hyperref[fig2s6]{6}). This may indicate that these regions
-do not act as a ``TE sink'' as has been previously proposed, or that
-the ``TE sink'' activity is restricted to very recent insertions, as
-the insertions we analysed in this population were likely older than
-those used in the \emph{KEE} study \cite{Grob:2014kg}.
+the set of TE variants found in gene bodies (Figure \ref{fig2}C,
+D). No significant enrichment was found for TE variants within the
+\emph{KNOT ENGAGED ELEMENT (KEE) }regions, previously identified as
+regions that may act as a ``TE sink'' \cite{Grob:2014kg} (Figure 2 -
+figure supplement \hyperref[fig2s6]{6}). This may indicate that these
+regions do not act as a ``TE sink'' as has been previously proposed,
+or that the ``TE sink'' activity is restricted to very recent
+insertions, as the insertions we analysed in this population were
+likely older than those used in the \emph{KEE} study
+\cite{Grob:2014kg}.
 
@@
 or unlinked from SNPs across the accessions. We tested how frequently
-common TE variants (\textgreater{}3 \% MAF, \textgreater{}7
-accessions) were linked to adjacent SNPs to determine when they would
-represent a previously unassessed source of genetic variation between
-accessions.  SNPs that were previously identified between the
-accessions \cite{Schmitz:2013iu} were compared to the presence/absence
-of individual TE variants.  For the testable TE variants in the
-population, the nearest 300 flanking SNPs upstream and 300 SNPs
+common TE variants (\textgreater{}3\% MAF, \textgreater{}7 accessions)
+were linked to adjacent SNPs to determine when they would represent a
+previously unassessed source of genetic variation between accessions.
+SNPs that were previously identified between the accessions
+\cite{Schmitz:2013iu} were compared to the presence/absence of
+individual TE variants. For the testable TE variants in the
+population, the nearest flanking 300 SNPs upstream and 300 SNPs
 downstream of the TE variant site were analyzed for local linkage
@@
 comparing ranked r\textsuperscript{2} values of TE variant to SNPs
-against the median ranked r\textsuperscript{2} value for all between
-SNP comparisons (SNP-SNP) to account for regional variation in the
-extent of SNP-SNP LD (Figure \ref{fig3}A, B) due to recombination rate
-variation or selection \cite{Horton:2012jh}. The majority (61\%) of
-testable TE variants had low LD with nearby SNPs, and represent a
-source of genetic diversity not previously assessed by SNP-based
-genotype calling methods (Figure \ref{fig3}C).  29\% of TE variants
-displayed high levels of LD and are tagged by nearby SNPs, while only
-10\% had intermediate levels of LD. We observed a positive correlation
-between TE variant MAF and LD state, with variants of a high minor
-allele frequency more often classified as high-LD (Figure
-\ref{fig3}D). While the proportion of TE variants classified as high,
-mid, or low-LD was mostly the same for both TE insertions and TE
-deletions, TE variants with a high MAF (\textgreater{}20\%) that were
-unable to be classified as either true deletions or as new insertions
-had a much higher proportion of high-LD variants (Figure \ref{fig3}E).
-This was consistent with the observation that the more common alleles
-were more often in a high-LD state.  TE variants displayed a similar
-distribution over chromosome 1 regardless of linkage classification
-(Figure 3 - figure supplement \hyperref[fig3s1]{1}). Overall, this
-analysis revealed an abundance of previously uncharacterized genetic
-variation that exists amongst Arabidopsis accessions caused by the
-presence or absence of TEs, and illustrates the importance of
-identifying TE variants alongside other genetic diversity such as
-SNPs.
+against the median ranked r\textsuperscript{2} value for all between SNP comparisons
+(SNP-SNP) to account for regional variation in the extent of SNP-SNP
+LD (Figure \ref{fig3}A, B) due to recombination rate variation or
+selection \cite{Horton:2012jh}. The majority (61\%) of testable TE
+variants had low LD with nearby SNPs, and represent a source of
+genetic diversity not previously assessed by SNP-based genotype
+calling methods (Figure \ref{fig3}C).  29\% of TE variants displayed
+high levels of LD and are tagged by nearby SNPs, while only 10\% had
+intermediate levels of LD. We observed a positive correlation between
+TE variant MAF and LD state, with variants of a high minor allele
+frequency more often classified as high-LD (Figure \ref{fig3}D). While
+the proportion of TE variants classified as high, mid, or low-LD was
+mostly the same for both TE insertions and TE deletions, TE variants
+with a high MAF (\textgreater{}20\%) that were unable to be classified
+as either true deletions or as new insertions had a much higher
+proportion of high-LD variants (Figure \ref{fig3}E). This was
+consistent with the observation that the more common alleles were more
+often in a high-LD state. TE variants displayed a similar distribution
+over chromosome 1 regardless of linkage classification (Figure 3 -
+figure supplement \hyperref[fig3s1]{1}). Overall, this analysis
+revealed an abundance of previously uncharacterized genetic variation
+that exists amongst Arabidopsis accessions caused by the presence or
+absence of TEs, and illustrates the importance of identifying TE
+variants alongside other genetic diversity such as SNPs.
 
@@
 kb gene upstream region, 5' UTR, exons, introns, 3' UTR or 2 kb
-downstream region (Figure \ref{fig4}A). While the steady state
-transcript abundance of most genes appeared to be unaffected by the
-presence of a TE, 196 genes displayed significant differences in
-transcript abundance linked with the presence of a TE variant,
-indicating a role for these variants in the local regulation of gene
-expression (1\% false discovery rate; \textgreater{}2-fold change in
-transcript abundance; Figure \ref{fig4}A, Figure 4 - source data
-1). No functional category enrichments in this set of differentially
-expressed genes were identified. As rare TE variants with a MAF less
-than 3\% may also be associated with a difference in transcript
-abundance, but were unable to be statistically tested due to their
-rarity, a burden test for enrichment of rare variants in the extremes
-of expression was performed \cite{Zhao:2016gc}. Briefly, this method
-counts the frequency of rare variants within each gene expression rank
-in the population, and aggregates this information over the entire
-population to determine whether an enrichment of rare variants exists
-within any gene expression rank. A strong enrichment for gene
-expression extremes was observed for TE variants in all gene features
-tested (Figure \ref{fig4}B). While TE variants in gene upstream
-regions showed a strong enrichment of both high and low gene
-expression ranks, TE variants in exons or gene downstream regions had
-a stronger enrichment for low expression ranks than high
-ranks. Randomization of the accession names removed these enrichments
-completely (Figure 4 - figure supplement \hyperref[fig4s1]{1}), and
-there was little difference between TE insertions and TE deletions in
-the gene expression rank enrichments found (Figure 4 - figure
-supplement \hyperref[fig4s2]{2}). This rare variant analysis further
-indicates that TE variants may alter the transcript abundance of
-nearby genes.
+downstream region (Figure \ref{fig4}A). While the steady state transcript
+abundance of most genes appeared to be unaffected by the presence of a
+TE, 196 genes displayed significant differences in transcript
+abundance linked with the presence of a TE variant, indicating a role
+for these variants in the local regulation of gene expression (1\%
+false discovery rate; \textgreater{}2-fold change in transcript
+abundance; Figure \ref{fig4}A, Figure 4 - source data 1). No functional
+category enrichments in this set of differentially expressed genes
+were identified. As rare TE variants with a MAF less than 3\% may also
+be associated with a difference in transcript abundance, but were
+unable to be statistically tested due to their rarity, a burden test
+for enrichment of rare variants in the extremes of expression was
+performed \cite{Zhao:2016gc}. Briefly, this method counts the
+frequency of rare variants within each gene expression rank in the
+population, and aggregates this information over the entire population
+to determine whether an enrichment of rare variants exists within any
+gene expression rank. A strong enrichment for gene expression extremes
+was observed for TE variants in all gene features tested (Figure
+\ref{fig4}B). While TE variants in gene upstream regions showed a
+strong enrichment of both high and low gene expression ranks, TE
+variants in exons or gene downstream regions had a stronger enrichment
+for low expression ranks than high ranks. Randomization of the
+accession names removed these enrichments completely (Figure 4 -
+figure supplement \hyperref[fig4s1]{1}), and there was little
+difference between TE insertions and TE deletions in the gene
+expression rank enrichments found (Figure 4 - figure supplement
+\hyperref[fig4s2]{2}). This rare variant analysis further indicates
+that TE variants may alter the transcript abundance of nearby genes.
 
@@
 As TEs are frequently highly methylated in Arabidopsis
-\cite{Zhang:2006eo, Zilberman:2006hx, Cokus:2008fc, Lister:2008bh},
+\cite{Zhang:2006eo, Zilberman:2007hx, Cokus:2008fc, Lister:2008bh},
 the DNA methylation state surrounding TE variant sites was assessed to
@@
 by chance, while it was significantly less than expected for CG-DMRs
-(p \textless{} $1 \times10^{-4}$, determined by resampling 10,000 times). Of
-the C-DMRs that were not close to a TE variant, 3,701 (27\% of all
-C-DMRs) were within 1 kb of a non-variable TE. Thus, 81\% of C-DMRs
-are within 1 kb of a TE when considering both fixed and variable TEs
-in the population. Of the remaining 19\% of C-DMRs, most were found in
-genes or intergenic regions.
+(p \textless{} $1 \times10^{-4}$, determined by resampling 10,000
+times). Of the C-DMRs that were not close to a TE variant, 3,701 (27\%
+of all C-DMRs) were within 1 kb of a non-variable TE. Thus, 81\% of
+C-DMRs are within 1 kb of a TE when considering both fixed and
+variable TEs in the population. Of the remaining 19\% of C-DMRs, most
+were found in genes or intergenic regions.
 
 To determine whether DMR methylation levels were dependent on the
-presence/absence of nearby TE variants, Pearson correlation coefficients
-were calculated between the DNA methylation level at each DMR and the
-presence/absence of the nearest TE variant. A negative correlation was
-observed between the distance from a C-DMR to the nearest TE insertion
-and the correlation between the DNA methylation level at the C-DMR with
-the presence/absence of the TE insertion (Figure \ref{fig6}B). This suggests a
-distance-dependent effect of TE insertion presence on C-DMR methylation.
-In contrast, no such relationship was found for TE deletions on C-DMRs,
-or for insertions or deletions on CG-DMRs (Figure \ref{fig6}B). DNA methylation
-levels at C-DMRs located within 1 kb of a TE insertion (TE-DMRs) were
-more often positively correlated with the presence/absence of a TE
+presence/absence of nearby TE variants, Pearson correlation
+coefficients were calculated between the DNA methylation level at each
+DMR and the presence/absence of the nearest TE variant. A negative
+correlation was observed between the distance from a C-DMR to the
+nearest TE insertion and the correlation between the DNA methylation
+level at the C-DMR with the presence/absence of the TE insertion
+(Figure \ref{fig6}B). This suggests a distance-dependent effect of TE
+insertion presence on C-DMR methylation.  In contrast, no such
+relationship was found for TE deletions on C-DMRs, or for insertions
+or deletions on CG-DMRs (Figure \ref{fig6}B). DNA methylation levels
+at C-DMRs located within 1 kb of a TE insertion (TE-DMRs) were more
+often positively correlated with the presence/absence of a TE
 insertion than the DNA methylation levels at C-DMRs further than 1 kb
 from a TE insertion (non-TE-DMRs). This was evident from the
-distribution of correlations for non-TE-DMRs being centred around zero,
-whereas for TE-DMRs this distribution was skewed to the right (Figure
-\ref{fig6}C, \emph{D}=0.24). For TE deletions, such a difference was not observed
-in the distributions of correlation coefficients between TE-DMRs and
-non-TE-DMRs, nor for CG-DMRs and their nearby TE insertions or deletions
-(Figure \ref{fig6}C, \emph{D=}0.07-0.10). Furthermore, DNA methylation levels
-were often higher in the presence of the nearby TE insertion, while this
+distribution of correlations for non-TE-DMRs being centred around
+zero, whereas for TE-DMRs this distribution was skewed to the right
+(Figure \ref{fig6}C, \emph{D}=0.24). For TE deletions, such a
+difference was not observed in the distributions of correlation
+coefficients between TE-DMRs and non-TE-DMRs, nor for CG-DMRs and
+their nearby TE insertions or deletions (Figure \ref{fig6}C,
+\emph{D=}0.07-0.10). Furthermore, DNA methylation levels were often
+higher in the presence of the nearby TE insertion, while this
 relationship was generally not observed for C-DMRs further than 1 kb
@@
 methylation levels at C-DMRs, and less so at CG-DMRs (Figure
-\ref{fig6}D). This further indicates that the presence of a TE insertion
-is associated with higher C-DMR methylation levels, while TE deletions
-appear to have more variable effects on DMR methylation levels. This
-enrichment was completely absent after repeating the analysis with
-randomized accession names (Figure 6 - figure supplement
+\ref{fig6}D). This further indicates that the presence of a TE
+insertion is associated with higher C-DMR methylation levels, while TE
+deletions appear to have more variable effects on DMR methylation
+levels. This enrichment was completely absent after repeating the
+analysis with randomized accession names (Figure 6 - figure supplement
 \hyperref[fig6s2]{2}). A slight enrichment was also observed for low
@@
 These results indicate that local DNA methylation patterns are
-influenced by the differential TE content between genomes, and that
-the DNA methylation-dependent silencing of TEs may frequently lead to
-the formation of DMRs between wild Arabidopsis accessions. TE
-insertions appear to be important in defining local patterns of DNA
-methylation, while DNA methylation levels often remain elevated
-following a TE deletion, and so are independent from the presence or
-absence of TEs in these cases. Importantly, the distance from a TE
-insertion to the centromere appears to have a strong impact on whether
-an alteration of local DNA methylation patterns will occur. This is
-likely due to flanking sequences being highly methylated in the
-pericentromeric regions, and so the insertion of a TE cannot further
-increase levels of DNA methylation. Overall, a large fraction of the
-population C-DMRs previously identified between wild accessions are
-correlated with the presence of local TE variants. CG-DMR methylation
-levels appear to be mostly independent from the presence/absence of
-common TE variants, while rare TE variants have an impact on DNA
-methylation levels at both C-DMRs and CG-DMRs. This analysis was aided
-by the high sensitivity of TEPID to detect TE variants, enabling the
-identification of 8x more variants than has been reported previously
-in the population \cite{Quadrana:2016bi}, allowing a thorough
-assessment of the impact of TE variants on DNA methylation patterns.
-
-\subsection{Genome-wide association scan highlights
-  distant and local control of DNA methylation}
+influenced by the differential TE content between genomes, and that the
+DNA methylation-dependent silencing of TEs may frequently lead to the
+formation of DMRs between wild Arabidopsis accessions. TE insertions
+appear to be important in defining local patterns of DNA methylation,
+while DNA methylation levels often remain elevated following a TE
+deletion, and so are independent from the presence or absence of TEs in
+these cases. Importantly, the distance from a TE insertion to the
+centromere appears to have a strong impact on whether an alteration of
+local DNA methylation patterns will occur. This is likely due to
+flanking sequences being highly methylated in the pericentromeric
+regions, and so the insertion of a TE cannot further increase levels of
+DNA methylation. Overall, a large fraction of the population C-DMRs
+previously identified between wild accessions are correlated with the
+presence of local TE variants. CG-DMR methylation levels appear to be
+mostly independent from the presence/absence of common TE variants,
+while rare TE variants have an impact on DNA methylation levels at both
+C-DMRs and CG-DMRs. This analysis was aided by the high sensitivity of
+TEPID to detect TE variants, allowing a thorough assessment of the
+impact of TE variants on DNA methylation patterns.
+
+\subsection{Genome-wide association scan highlights distant and local
+control of DNA methylation}
 
@@
 signifying a local association between TE insertion presence/absence
-and C-DMR methylation (Figure \ref{fig8}A).  This local association was
-not as strong for TE deletions (Figure \ref{fig8}B), consistent with our above
-findings. While TE variants and DNA methylation showed a local
-association, it is also possible that TE variation may influence DNA
-methylation states more broadly in the genome, perhaps through
-production of \emph{trans-}acting smRNAs or inactivation of genes
-involved in DNA methylation establishment or maintenance. To identify
-any potential enrichment of C-DMRs regulated in \emph{trans}, the
-total number of significant associations was summed for each TE
-variant across the whole genome (Figure \ref{fig8}A and B, top panels). At
-many sites, far more significant associations were found than expected
-due to the false positive rate alone. This suggested the existence of
-many putative \emph{trans }associations between TE variants and
-genome-wide C-DMR methylation levels. These C-DMRs that appeared to be
-associated with a TE insertion in \emph{trans} were further
-examined\emph{, }checking for TE insertions near these C-DMRs that
+and C-DMR methylation (Figure \ref{fig8}A).  This local association
+was not as strong for TE deletions (Figure \ref{fig8}B), consistent
+with our above findings. While TE variants and DNA methylation showed
+a local association, it is also possible that TE variation may
+influence DNA methylation states more broadly in the genome, perhaps
+through production of \emph{trans-}acting smRNAs or inactivation of
+genes involved in DNA methylation establishment or maintenance. To
+identify any potential enrichment of C-DMRs regulated in \emph{trans},
+the total number of significant associations was summed for each TE
+variant across the whole genome (Figure \ref{fig8}A and B, top
+panels). At many sites, far more significant associations were found
+than expected due to the false positive rate alone. This suggested the
+existence of many putative \emph{trans }associations between TE
+variants and genome-wide C-DMR methylation levels. These C-DMRs that
+appeared to be associated with a TE insertion in \emph{trans} were
+further examined, checking for TE insertions near these C-DMRs that
 were present in the same accessions as the \emph{trans }associated TE,
@@
 variants with a minor allele frequency above 3\% were able to be tested
-for linkage with nearby SNPs. The majority of these TE variants were not
-in LD with SNPs, indicating that they represent genetic variants
-currently overlooked in genomic studies. A marked depletion of TE
-variants within gene bodies and DNase I hypersensitivity sites (putative
-regulatory regions) is consistent with the more deleterious TE
-insertions being removed from the population through selection. Of those
-TE variants found in gene bodies, TE deletions were overrepresented,
-indicating that the loss of ancestral TEs inserted within genes may be
-more frequent, or perhaps less deleterious, than the \emph{de novo
-}insertion of TEs into genes.
-
-The high sensitivity of our TE variant detection method allowed
-identification of 8x more variants than has been reported previously
-\cite{Quadrana:2016bi}, and the identification of such a large number of
-TE variants (23,095) in this population gave an opportunity to form
-statistically robust correlations between TE presence/absence and
-transcript abundance from nearby genes, as well as genome-wide patterns
-of DNA methylation. Examples were identified where TE variants appear to
-have an effect upon gene expression, both in the disruption of
-transcription and in the spreading or disruption of regulatory
-information leading to the transcriptional activation of genes,
-indicating that these TE variants can have important consequences upon
-the expression of protein coding genes (Figure \ref{fig5}). In one case, these
-changes in gene expression could be linked with phenotypic changes, with
-accessions containing a TE insertion more frequently sensitive to
-bacterial infection. Further experiments will be needed to establish a
-causal link between this TE insertion and the associated phenotype. An
-analysis of rare TE variants, present at a low MAF, further strengthened
-this relationship between TE presence/absence and altered transcript
-abundance, as a strong enrichment of rare TE variants in accessions with
-extreme gene expression ranks in the population was identified.
+for linkage with nearby SNPs. The majority of these TE variants
+exhibited only low levels of LD with nearby SNPs, indicating that they
+represent genetic variants currently overlooked in genomic studies. A
+marked depletion of TE variants within gene bodies and DNase I
+hypersensitivity sites (putative regulatory regions) is consistent with
+the more deleterious TE insertions being removed from the population
+through selection. Of those TE variants found in gene bodies, TE
+deletions were overrepresented, indicating that the loss of ancestral
+TEs inserted within genes may be more frequent, or perhaps less
+deleterious, than the \emph{de novo }insertion of TEs into genes.
+
+A previous study focused on recent TE insertions in the Arabidopsis
+population \cite{Quadrana:2016bi}, thus the extensive variation
+between accessions due to older TE insertions or TE deletions has not
+been explored. We examined both old and new TE variants (common and
+rare alleles), as well as TE deletions, allowing the detection of
+23,095 TE variants in the population. This large sample size gave an
+opportunity to form statistically robust correlations between TE
+presence/absence and transcript abundance from nearby genes, as well
+as genome-wide patterns of DNA methylation. Examples were identified
+where TE variants appear to have an effect upon gene expression, both
+in the disruption of transcription and in the spreading or disruption
+of regulatory information leading to the transcriptional activation of
+genes, indicating that these TE variants can have important
+consequences upon the expression of protein coding genes (Figure
+\ref{fig5}). In one case, these changes in gene expression could be
+linked with phenotypic changes, with accessions containing a TE
+insertion more frequently sensitive to bacterial infection. Further
+experiments will be needed to establish a causal link between this TE
+insertion and the associated phenotype. An analysis of rare TE
+variants, present at a low MAF, further strengthened this relationship
+between TE presence/absence and altered transcript abundance, as a
+strong enrichment of rare TE variants in accessions with extreme gene
+expression ranks in the population was identified.  Therefore, the
+effects of TE insertions appear to be long-lasting, as there was
+little difference between common (old) and rare (young) variants in
+the impact upon gene expression (Figure \ref{fig4}).
 
@@
 81\% of all C-DMRs were within 1 kb of a TE, when considering both fixed
-and variable TEs in the population, a much higher percentage than has
-been reported previously \cite{Schmitz:2013iu, Quadrana:2016bi}. We did
-not find evidence of CG-DMR methylation level being altered by the
-presence of common TE variants, although rare TE variants may be more
-important in shaping patterns of DNA methylation at some CG-DMRs, though
-the reasons for this distinction remain unclear. The level of local DNA
-methylation changes associated with TE variants was also related to the
-distance from a TE variant to the centromere, with variants in the
-chromosome arms being more strongly correlated with DNA methylation
-levels. This seems to be due to a higher baseline level of DNA
-methylation at the pericentromeric regions, which prevent any further
-increase in DNA methylation level following insertion of a TE.
-Furthermore, we found an important distinction between TE insertions and
-TE deletions in the effect that these variants have on nearby DNA
-methylation levels. While flanking DNA methylation levels increase
-following a TE insertion, the deletion of an ancestral TE was often not
-associated with a corresponding decrease in flanking DNA methylation
-levels (Figure \ref{fig7}). This indicates that high levels of DNA methylation,
-once established, may be maintained in the absence of the TE insertion
-that presumably triggered the original change in DNA methylation level.
-It is then possible that TE variants explain more of the inter-accession
-variation in DNA methylation patterns than we find direct evidence for,
-if some C-DMRs were formed by the insertion of an ancestral TE that is
-now absent in all the accessions analysed here. These DMRs would then
-represent the epigenetic ``scars'' of past TE insertions.
-
-Finally, a genome-wide scan of common TE variant association with C-DMR
-methylation levels provides further evidence of a strong local
-association between TE insertion presence/absence and C-DMR methylation
-level (Figure \ref{fig8}). The identification of some TE variants that appeared
-to be associated with changes in DNA methylation levels at multiple loci
-throughout the genome indicates possible \emph{trans} regulation of DNA
-methylation state linked to specific TE variants. Further experiments
-will be required to confirm and examine the role of these TE variants in
-determining genome-wide patterns of DNA methylation. Overall, our
-results show that TE presence/absence variants between wild Arabidopsis
-accessions not only have important effects on nearby gene expression,
-but can also have a role in determining local patterns of DNA
-methylation, and explain many regions of differential DNA methylation
-previously observed in the population.
+and variable TEs in the population. We did not find evidence of CG-DMR
+methylation level being altered by the presence of common TE variants,
+although rare TE variants may be more important in shaping patterns of
+DNA methylation at some CG-DMRs, though the reasons for this distinction
+remain unclear. The level of local DNA methylation changes associated
+with TE variants was also related to the distance from a TE variant to
+the centromere, with variants in the chromosome arms being more strongly
+correlated with DNA methylation levels. This seems to be due to a higher
+baseline level of DNA methylation at the pericentromeric regions, which
+prevent any further increase in DNA methylation level following
+insertion of a TE. Furthermore, we found an important distinction
+between TE insertions and TE deletions in the effect that these variants
+have on nearby DNA methylation levels. While flanking DNA methylation
+levels increase following a TE insertion, the deletion of an ancestral
+TE was often not associated with a corresponding decrease in flanking
+DNA methylation levels (Figure \ref{fig7}). This indicates that high levels of
+DNA methylation, once established, may be maintained in the absence of
+the TE insertion that presumably triggered the original change in DNA
+methylation level. It is then possible that TE variants explain more of
+the inter-accession variation in DNA methylation patterns than we find
+direct evidence for, if some C-DMRs were formed by the insertion of an
+ancestral TE that is now absent in all the accessions analysed here.
+These DMRs would then represent the epigenetic ``scars'' of past TE
+insertions.
+
+Finally, a genome-wide scan of common TE variant association with
+C-DMR methylation levels provides further evidence of a strong local
+association between TE insertion presence/absence and C-DMR
+methylation level (Figure \ref{fig8}). The identification of some TE
+variants that appeared to be associated with changes in DNA
+methylation levels at multiple loci throughout the genome indicates
+possible \emph{trans} regulation of DNA methylation state linked to
+specific TE variants. Further experiments will be required to confirm
+and examine the role of these TE variants in determining genome-wide
+patterns of DNA methylation. Overall, our results show that TE
+presence/absence variants between wild Arabidopsis accessions not only
+have important effects on nearby gene expression, but can also have a
+role in determining local patterns of DNA methylation, and explain
+many regions of differential DNA methylation previously observed in
+the population.
 
@@
 read alignments are first filtered to remove reads where the distance
-between split mapping loci is less than 5 kb, to remove split reads due
-to small indels, or split reads with a mapping quality (MAPQ) less than
-5. Split and discordant read mapping coordinates are then intersected
-using pybedtools \cite{Dale:2011cl, Quinlan:2010km} with the Col-0
-reference TE annotation, requiring 80\% overlap between TE and read
-mapping coordinates. To determine putative TE insertion sites, regions
-are then identified that contain independent discordant read pairs
-aligned in an orientation facing one another at the insertion site, with
-their mate pairs intersecting with the same TE (Figure \ref{fig1}). The total
-number of split and discordant reads intersecting the insertion site and
-the TE is then calculated, and a TE insertion predicted where the
-combined number of reads is greater than a threshold determined by the
-average sequencing depth over the whole genome (1/10 coverage if
-coverage is greater than 10, otherwise a minimum of 2 reads).
-Alternatively, in the absence of discordant reads mapped in orientations
-facing one another, the required total number of split and discordant
-reads at the insertion site linked to the inserted TE is set higher,
-requiring twice as many reads.
+between split mapping loci is less than 5 kb, to remove split reads
+due to small indels, or split reads with a mapping quality (MAPQ) less
+than 5. Split and discordant read mapping coordinates are then
+intersected using pybedtools \cite{Dale:2011cl, Quinlan:2010km} with
+the Col-0 reference TE annotation, requiring 80\% overlap between TE
+and read mapping coordinates. To determine putative TE insertion
+sites, regions are then identified that contain independent discordant
+read pairs aligned in an orientation facing one another at the
+insertion site, with their mate pairs intersecting with the same TE
+(Figure \ref{fig1}). The total number of split and discordant reads
+intersecting the insertion site and the TE is then calculated, and a
+TE insertion predicted where the combined number of reads is greater
+than a threshold determined by the average sequencing depth over the
+whole genome (1/10 coverage if coverage is greater than 10, otherwise
+a minimum of 2 reads).  Alternatively, in the absence of discordant
+reads mapped in orientations facing one another, the required total
+number of split and discordant reads at the insertion site linked to
+the inserted TE is set higher, requiring twice as many reads.
 
@@
 minimum number of split and discordant reads present, determined by
-the sequencing depth (1/10 coverage; Figure \ref{fig1}). A threshold of 80\%
-TE sequence spanned by split or discordant reads is used, as opposed
-to 100\%, to account for misannotation of TE sequence boundaries in
-the Col-0 reference TE annotation, as well as TE fragments left behind
-by DNA TEs during cut-paste transposition (TE footprints) that may
-affect the mapping of reads around annotated TE borders
-\cite{Plasterk:1991vd}. Furthermore, the coverage within the spanned
-region may be more than 10\% that of the flanking sequence, but in
-such cases twice as many split and discordant reads are required. If
-multiple TEs are spanned by the split and discordant reads, and the
-above requirements are met, multiple TEs in the same region can be
-identified as absent with respect to the reference genome. Absence
-variants in non-Col-0 accessions are subsequently recategorized as TE
-insertions present in the Col-0 genome but absent from a given wild
-accession.
+the sequencing depth (1/10 coverage; Figure \ref{fig1}). A threshold
+of 80\% TE sequence spanned by split or discordant reads is used, as
+opposed to 100\%, to account for misannotation of TE sequence
+boundaries in the Col-0 reference TE annotation, as well as TE
+fragments left behind by DNA TEs during cut-paste transposition (TE
+footprints) that may affect the mapping of reads around annotated TE
+borders \cite{Plasterk:1991vd}. Furthermore, the coverage within the
+spanned region may be more than 10\% that of the flanking sequence,
+but in such cases twice as many split and discordant reads are
+required. If multiple TEs are spanned by the split and discordant
+reads, and the above requirements are met, multiple TEs in the same
+region can be identified as absent with respect to the reference
+genome. Absence variants in non-Col-0 accessions are subsequently
+recategorized as TE insertions present in the Col-0 genome but absent
+from a given wild accession.
 
@@
 
-To test the sensitivity and specificity of TEPID, 100 TE insertions (50
-copy-paste transpositions, 50 cut-paste transpositions) and 100 TE
+To test the sensitivity and specificity of TEPID, 100 TE insertions
+(50 copy-paste transpositions, 50 cut-paste transpositions) and 100 TE
 absence variants were simulated in the Arabidopsis genome using the
-RSVSim R package, version 1.7.2 \cite{Bartenhagen:2013kl}, and synthetic
-reads generated from the modified genome at various levels of sequencing
-coverage using wgsim \cite{Li:2009ka} (\url{https://github.com/lh3/wgsim}).
-These reads were then used to calculate the true positive, false
-positive, and false negative TE variant discovery rates for TEPID at
-various sequencing depths, by running `tepid-map' and `tepid-discover'
-using the simulated reads with the default parameters (Figure 1 - figure
-supplement \hyperref[fig1s1]{1}).
+RSVSim R package, version 1.7.2 \cite{Bartenhagen:2013kl}, and
+synthetic reads generated from the modified genome at various levels
+of sequencing coverage using wgsim \cite{Li:2009ka}
+(\url{https://github.com/lh3/wgsim}).  These reads were then used to
+calculate the true positive, false positive, and false negative TE
+variant discovery rates for TEPID at various sequencing depths, by
+running `tepid-map' and `tepid-discover' using the simulated reads
+with the default parameters (Figure 1 - figure supplement
+\hyperref[fig1s1]{1}).
 
@@
 using RepeatMasker with RepBase-derived libraries and the default
-parameters (version 4.0.5, \url{http://www.repeatmasker.org}). A
-blastn search was then conducted against the Col-0 genome using the
-following parameters: `-max-target-seqs 1', `-evalue 1e-6'
-\cite{Camacho:2009fc}.  Coordinates of the top blast hit for each read
-location were then compared with the TE variant sites identified using
-those reads. To estimate false negative rates for TEPID TE absence
-calls, L\emph{er} TE absence calls were compared with a known set of
-Col-0-specific TE insertions, absent in L\emph{er
-}\cite{Quadrana:2016bi}. For TEPID TE insertion calls, we mapped Col-0
-DNA sequencing reads \cite{Jiang:2014ih} to the L\emph{er }PacBio
-assembly, and identified sites with read evidence reaching the TEPID
-threshold for a TE insertion call to be made.
+parameters (version 4.0.5, \url{http://www.repeatmasker.org}). A blastn
+search was then conducted against the Col-0 genome using the following
+parameters: `-max-target-seqs 1', `-evalue 1e-6' \cite{Camacho:2009fc}.
+Coordinates of the top blast hit for each read location were then
+compared with the TE variant sites identified using those reads. To
+estimate false negative rates for TEPID TE absence calls, L\emph{er} TE
+absence calls were compared with a known set of Col-0-specific TE
+insertions, absent in L\emph{er }\cite{Quadrana:2016bi}. For TEPID TE
+insertion calls, we mapped Col-0 DNA sequencing reads
+\cite{Jiang:2014ih} to the L\emph{er }PacBio assembly, and identified
+sites with read evidence reaching the TEPID threshold for a TE insertion
+call to be made.
 
@@
 reproduce this analysis can be found at
-\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er }TE
-variants are available in Figure 1 - source data 1 and 2. TE variants
-identified among the 216 wild Arabidopsis accessions resequenced by
-Schmitz et al. (2013) are available in Figure 2 - source data 1, 2 and
-3. Source data is available on Dryad
+\url{https://github.com/timoast/Arabidopsis-TE-variants}. L\emph{er}
+TE variants are available in Figure 1 - source data 1 and 2. TE
+variants identified among the 216 wild Arabidopsis accessions
+resequenced by Schmitz et al. (2013) are available in Figure 2 -
+source data 1, 2 and 3. Source data is available on Dryad
 (\url{http://dx.doi.org/10.5061/dryad.187b3}).

### 2016-11-14 AnalysisCode/intergenic_regions.Rmd
Add R code for getting intergenic regions

---
title: "Intergenic regions"
author: "Tim Stuart"
date: "11/11/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(GenomicRanges)
library(rtracklayer)
library(BSgenome.Athaliana.TAIR.TAIR9)
library(readr)
library(dplyr)
```

Load all the tracks

```{r}
genes <- import.gff("../RawData/TAIR10_GFF3_genes.gff.gz")
te <- import("../RawData/TAIR9_TE.bed.gz",
             format = "bed",
             extraCols = c(strand = "character",
                           TE = "character",
                           family = "character",
                           superfamily = "character"))
pseudogene <- import("../ProcessedData/GeneFeatures/pseudogene.bed.gz",
                     format = "bed",
                     extraCols = c(dot = "character",
                                   strand = "character",
                                   source = "character",
                                   type = "character",
                                   dot = "character",
                                   id = "character"))
dhs <- import("../RawData/Sullivan_DHS_PE_peaks_control.bed.gz",
              format = "bed",
              extraCols = c(dot = "character",
                            zero = "numeric",
                            second_dot = "character",
                            a = "numeric",
                            b = "numeric"))
centromeres <- import("../RawData/centromere_positions.txt",
                      format = "bed")

# make chromosome names match the genome
genes <- dropSeqlevels(genes, c("chrM", "chrC"))
genes <- renameSeqlevels(genes, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
te <- renameSeqlevels(te, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
pseudogene <- renameSeqlevels(pseudogene, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
dhs <- renameSeqlevels(dhs, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
centromeres <- renameSeqlevels(centromeres, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))

# remove metadata columns
mcols(te) <- NULL
mcols(genes) <- NULL
mcols(pseudogene) <- NULL
mcols(dhs) <- NULL
```

Now extend gene coordinates 2 kb in each direction.

```{r extend_coords}
# need to remove entries that cover the entire chromosome, they will overlap everything
chromosome_entries <- width(ranges(genes)) > 10^6

# check we have 5 chromosomes 
sum(chromosome_entries)

# now remove those entries
genes <- genes[!chromosome_entries]

# now extend 2 kb each direction
extended_genes <- genes + 2000
```

Merge all the coordinates and reduce the final genomic ranges object, then take the coordinates between.

```{r}
all_features <- reduce(c(extended_genes, te, pseudogene, dhs, centromeres))
seq_between <- gaps(all_features)
# remove any of length 1
intergenic <- seq_between[!width(seq_between) == 1]
```

Now intersect with the TE variants to find the proportion of all these ranges that contain at least 1 TE variant

First load the TEPID data:

```{r intersect_TEPID}
tepid <- read_tsv("../RawData/TEPID_TEPAV.tsv.gz")
tepid_coords <- makeGRangesFromDataFrame(tepid[,1:3])
tepid_coords <- renameSeqlevels(tepid_coords, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
```

Find the percentage of all intergenic regions that contain a TE insertion

```{r}
length(subsetByOverlaps(intergenic, tepid_coords)) / length(intergenic) * 100
```

Now try splitting up all intergenic regions longer than 1 kb. If longer than x kb, cut into x+1 equal bins, where x >= 1

```{r split_intergenic}
gr <- GRanges()
for(i in seq_along(intergenic)) {
        l <- width(intergenic[i])
        if(l < 1000) {
                gr <- c(gr, intergenic[i])
        } else {
                gr <- c(gr, unlist(tile(intergenic[i], n = floor(l / 1000))))
        }
}
```

```{r intersect_tiled_intergenic}
length(subsetByOverlaps(gr, tepid_coords)) / length(gr) * 100
```

Let's go with the original, non-cut, data.

### 2016-11-15 ProcessingCode/make_genomic_feature_files.sh
Add intergenic regions

@@ -10,6 +10,7 @@
 # Pseudogenes
 # Other Col-0 TEs
 # DNase I hypersensitivity sites
+# nongenic, non DHS, non TE, non pseudogene
 
 # Information of the allele frequency or insertion classification (insertion or deletion)
 # can be found in the ProcessedData/GeneFeatures/*_intersections.bed.gz files
@@ -177,6 +178,14 @@ wc -l ../ProcessedData/GeneFeatures/dhs_intersections.bed \
 # record total number
 gunzip -c ../RawData/Sullivan_DHS_PE_peaks_control.bed.gz \
 | wc -l \
+>> ../ProcessedData/gene_feature_counts.csv
+
+# Intergenic regions
+bedtools intersect -a tepid_tepav.tsv -b ../ProcessedData/GeneFeatures/intergenic_regions.bed -wb \
+> ../ProcessedData/GeneFeatures/intergenic_intersections.bed
+
+# record total number
+wc -l ../ProcessedData/GeneFeatures/intergenic_regions.bed \
 >> ../ProcessedData/gene_feature_counts.csv
 
 rm tepid_tepav.tsv

### 2016-11-15 AnalysisCode/intergenic_regions.Rmd
Move code for barplot

@@
 genes <- import.gff("../RawData/TAIR10_GFF3_genes.gff.gz")
+# filter so only type == gene
+genes <- genes[genes$type == "gene"]
 te <- import("../RawData/TAIR9_TE.bed.gz",
@@
 ```{r extend_coords}
-# need to remove entries that cover the entire chromosome, they will overlap everything
-chromosome_entries <- width(ranges(genes)) > 10^6
-
-# check we have 5 chromosomes 
-sum(chromosome_entries)
-
-# now remove those entries
-genes <- genes[!chromosome_entries]
-
-# now extend 2 kb each direction
 extended_genes <- genes + 2000
@@
 
-Find the percentage of all intergenic regions that contain a TE insertion
+```{r}
+fractionIntergenic <- length(subsetByOverlaps(intergenic, tepid_coords)) / length(intergenic) * 100
+```
+
+Write the intergenic regions to file
+
+```{r write}
+intergenic <- renameSeqlevels(intergenic, c("chr1", "chr2", "chr3", "chr4", "chr5"))
+write_tsv(as_data_frame(intergenic), "../ProcessedData/GeneFeatures/intergenic_regions.bed", col_names = FALSE)
+```
+
+## Repeat all gene feature intersections
+
+Need to repeat everything as some intervals were counted twice if two TEs were inserted in them.
+This is what we want for the insertion/deletion rare/common enrichments for fig 2C, D, but not for the barplot.
 
 ```{r}
-length(subsetByOverlaps(intergenic, tepid_coords)) / length(intergenic) * 100
+fractionGene <- length(subsetByOverlaps(genes, tepid_coords)) / length(genes) * 100
+fractionDHS <- length(subsetByOverlaps(dhs, tepid_coords)) / length(dhs) * 100
+fractionTE <- length(subsetByOverlaps(te, tepid_coords)) / length(te) * 100
+fractionPseudogene <- length(subsetByOverlaps(pseudogene, tepid_coords)) / length(pseudogene) * 100
 ```
 
-Now try splitting up all intergenic regions longer than 1 kb. If longer than x kb, cut into x+1 equal bins, where x >= 1
+Now need the other parts -- exon, intron, utr, upstream, downstream
 
-```{r split_intergenic}
-gr <- GRanges()
-for(i in seq_along(intergenic)) {
-        l <- width(intergenic[i])
-        if(l < 1000) {
-                gr <- c(gr, intergenic[i])
-        } else {
-                gr <- c(gr, unlist(tile(intergenic[i], n = floor(l / 1000))))
-        }
-}
+```{r read_regions}
+introns <- import("../ProcessedData/GeneFeatures/introns.bed.gz",
+                  format = "bed")
+exons <- import("../ProcessedData/GeneFeatures/exons.bed.gz",
+                format = "bed")
+utr3 <- import("../ProcessedData/GeneFeatures/utr3.bed.gz",
+               format = "bed")
+utr5 <- import("../ProcessedData/GeneFeatures/utr5.bed.gz",
+               format = "bed")
+upstream <- flank(genes, start = TRUE, both = FALSE, width = 2000)
+downstream <- flank(genes, start = FALSE, both = FALSE, width = 2000)
+
+# rename seq levels and drop metadata
+introns <- dropSeqlevels(introns, c("chrM", "chrC"))
+introns <- renameSeqlevels(introns, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
+mcols(introns) <- NULL
+
+exons <- dropSeqlevels(exons, c("chrM", "chrC"))
+exons <- renameSeqlevels(exons, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
+mcols(exons) <- NULL
+
+utr3 <- renameSeqlevels(utr3, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
+mcols(utr3) <- NULL
+
+utr5 <- dropSeqlevels(utr5, c("chrM", "chrC"))
+utr5 <- renameSeqlevels(utr5, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
+mcols(utr5) <- NULL
 ```
 
-```{r intersect_tiled_intergenic}
-length(subsetByOverlaps(gr, tepid_coords)) / length(gr) * 100
+Get the intersections with TE variants
+
+```{r}
+fractionIntron <- length(subsetByOverlaps(introns, tepid_coords)) / length(introns) * 100
+fractionExon <- length(subsetByOverlaps(exons, tepid_coords)) / length(exons) * 100
+fractionUTR3 <- length(subsetByOverlaps(utr3, tepid_coords)) / length(utr3) * 100
+fractionUTR5 <- length(subsetByOverlaps(utr5, tepid_coords)) / length(utr5) * 100
+fractionUpstream <- length(subsetByOverlaps(upstream, tepid_coords)) / length(upstream) * 100
+fractionDownstream <- length(subsetByOverlaps(downstream, tepid_coords)) / length(downstream) *100
 ```
 
-Let's go with the original, non-cut, data.
+```{r}
+all_fractions <- c(fractionDHS, fractionUpstream, fractionUTR5, fractionExon, fractionIntron,
+                   fractionUTR3, fractionDownstream, fractionPseudogene, fractionTE, fractionIntergenic)
+barplot_names <- c("DHS", "Upstream", "5' UTR", "Exon", "Intron", "3' UTR", "Downstream", "Pseudogenes", "TE", "Intergenic")
+intersection_df <- data.frame(all_fractions, barplot_names)
+```
+
+Barplot
+
+```{r}
+barplot(intersection_df$all_fractions, names.arg = intersection_df$barplot_names, las=2, ylim = c(0, 40),
+        ylab = "% contining TE insertion", main = "Genomic features")
+```
+
+Save plot
+
+```{r}
+pdf("../Plots/barplot_genomic_feature_intersections.pdf", height = 4, width = 4)
+barplot(intersection_df$all_fractions, names.arg = intersection_df$barplot_names, las=2, ylim = c(0, 40),
+        ylab = "% contining TE insertion", main = "Genomic features")
+dev.off()
+```

### 2016-11-15 AnalysisCode/TE_insertion_distribution.R

@@ -223,16 +223,9 @@ dhsIntersections <- read_tsv("../ProcessedData/GeneFeatures/dhs_intersections.be
 all_dhs <- read_tsv("../RawData/Sullivan_DHS_PE_peaks_control.bed.gz", col_names = F)
 fractionDHS <- nrow(dhsIntersections) / nrow(all_dhs) * 100
 
-# make barplot
-all_fractions <- c(fractionDHS, fractionUpstream, fractionUTR5, fractionExon, fractionIntrons,
-                   fractionUTR3, fractionDownstream, fractionPseudogenes, fractionTE)
-barplot_names <- c("DHS", "Upstream", "5' UTR", "Exon", "Intron", "3' UTR", "Downstream", "Pseudogenes", "TE")
-intersection_df <- data.frame(all_fractions, barplot_names)
-
-pdf("../Plots/barplot_genomic_feature_intersections.pdf", height = 4, width = 4)
-barplot(intersection_df$all_fractions, names.arg = intersection_df$barplot_names, las=2, ylim = c(0, 50),
-        ylab = "Percentage contining TE insertion", main = "TE variant frequency in genomic features")
-dev.off()
+# Intergenic
+intergenicIntersections <- read_tsv("../ProcessedData/GeneFeatures/intergenic_intersections.bed.gz",
+                                    col_names = FALSE)
 
 # now look at rare/common insertion/deletion frequency in each feature
 # make a single dataframe with all the features, absence classification and frequency classification
@@ -279,6 +272,11 @@ df <- dhsIntersections %>%
   select(feature, FrequencyClassification, AbsenceClassification) %>%
   rbind(df)
 
+df <- intergenicIntersections %>%
+        mutate(feature = "Intergenic", FrequencyClassification = X12, AbsenceClassification = X10) %>%
+        select(feature, FrequencyClassification, AbsenceClassification) %>%
+        rbind(df)
+
 df %>%
   group_by(feature) %>%
   mutate(rare_perc = sum(FrequencyClassification == "Rare") / n() * 100,
@@ -289,7 +287,7 @@ df %>%
   select(feature, rare_perc:del_na_perc) %>% unique(.) -> insertionStats
 
 # make plots
-feature_order <-  c("DHS", "Upstream", "5' UTR", "Exon", "Intron", "3' UTR", "Downstream", "Pseudogene", "TE")
[…]

### 2016-11-16 AnalysisCode/gene_expression_volcano.R
Fixed the allele frequency cutoff used

@@ -44,35 +44,31 @@ intron <- intron[!duplicated(intron),]
 utr5 <- utr5[!duplicated(utr5),]	
 utr3 <- utr3[!duplicated(utr3),]	
 downstream <- downstream[!duplicated(downstream),]	
-	
-find.len <- function(d) {	
-  return(length(unlist(strsplit(d, ","))))	
-}	
 
-# filter where insertion in at least 5 accessions (MAF >3%)	
+# filter where allele frequency definition is common	
 upstream %>%
-  rowwise() %>%
-  filter(find.len(pos_accessions) > 4) -> upstream.five	
+        rowwise() %>%
+        filter(X12 == "Common") -> upstream.five	
 	
 exon %>%
-  rowwise() %>%
-  filter(find.len(pos_accessions) > 4) -> exon.five	
+        rowwise() %>%
+        filter(X12 == "Common") -> exon.five	
 	
 intron %>%	
   rowwise() %>%	
-  filter(find.len(pos_accessions) > 4) -> intron.five	
+        filter(X12 == "Common") -> intron.five	
 	
 utr3 %>%	
   rowwise() %>%	
-  filter(find.len(pos_accessions) > 4) -> utr3.five	
+        filter(X12 == "Common") -> utr3.five	
 	
 utr5 %>%	
   rowwise() %>%	
-  filter(find.len(pos_accessions) > 4) -> utr5.five	
+        filter(X12 == "Common") -> utr5.five	
 	
 downstream %>%	
   rowwise() %>%	
-  filter(find.len(pos_accessions) > 4) -> downstream.five
+        filter(X12 == "Common") -> downstream.five
 
 lookupExp <- function(accessions, gene, exp) {	
   x <- vector()	
@@ -97,7 +93,7 @@ test.sig.cor <- function(genes, expression){
     ins.data <- ins.data[is.finite(ins.data)]	
     no.ins.data <- lookupExp(no.ins.acc, gene, expression)	
     no.ins.data <- no.ins.data[is.finite(no.ins.data)]	
-    if(length(ins.data) > 5 && length(no.ins.data) > 5) {	
+    if(length(ins.data) > 6 && length(no.ins.data) > 6) {	
       pval <- wilcox.test(ins.data, no.ins.data)[[3]]	
       r <- cor(c(ins.data, no.ins.data),	
           c(rep(1, length(ins.data)),	

### 2016-11-20 AnalysisCode/basho.Rmd
Add basho allele freq comparison

---
title: "basho"
author: "Tim Stuart"
date: "20/11/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basho allele frequencies

Aim: compare the estimated allele frequencies for TEs genotyped by Hollister and Gaut (2007) to my allele frequencies for the same TEs using the 1001 genomes read data.

First, read in the data we need:

```{r libraries, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(GenomicRanges)
library(rtracklayer)
```

```{r load_data}
basho <- read_csv("../RawData/Gaut_basho_supplement.csv", skip = 1)
tepav <- read_tsv("../RawData/TEPID_TEPAV.tsv.gz")
colnames(basho) <- make.names(colnames(basho))
basho_gr <- basho[,c(3:5, 8)] %>%
        makeGRangesFromDataFrame(start.field = "Testart..bp.", end.field = "Teend..bp.",
                                 keep.extra.columns = TRUE)
basho_gr <- renameSeqlevels(basho_gr, c("chr1", "chr2", "chr3", "chr4", "chr5"))

te <- import("../../TEPID/Annotation/Arabidopsis/TAIR9_TE.bed.gz",
             format = "bed",
             extraCols = c(strand = "character",
                           TE = "character",
                           family = "character",
                           superfamily = "character"))
```

For each TE set of coordinates in the basho set, find what TE overlaps and record this information. Then get the number of accessions predicted to contain/lack that TE from my data, and the genotyping results from Gaut.

```{r}
hits <- findOverlaps(te, basho_gr)
basho_hits <- as_data_frame(basho_gr[subjectHits(hits),])
```

```{r}
te_hits <- as_data_frame(te[queryHits(hits),])
```

```{r}
basho_hits$TE <- te_hits$TE
```

Now join with the TEPAV data

```{r}
dat <- left_join(basho_hits, tepav, by = "TE") %>%
        rowwise() %>%
        filter("Col-0" %in% unlist(strsplit(Accessions_TE_present, ","))) %>%
        select(Occ..Freq., Count_TE_present)

p <- ggplot(dat, aes(Occ..Freq., Count_TE_present/216)) + geom_point() +
        geom_smooth(method = "lm", se = FALSE) +
        xlab("Occupation frequency Hollister and Gaut (2007)") +
        ylab("Occupation frequency TEPID data") +
        theme_bw() +
        ggtitle("Basho TE allele frequencies")

p
```

Save plot

```{r}
ggsave("../Plots/basho_alleles.pdf", p, height = 10, width = 15, units = "cm", useDingbats = FALSE)
```

```{r}
model <- lm(dat$Occ..Freq. ~ dat$Count_TE_present)
summary(model)
```

### 2016-11-21 AnalysisCode/variants_per_accession.R
Change plot size

@@ -31,7 +31,7 @@ df %>%
         annotate("text", x = 30, y = 12, label = "Col-0") +
         xlab("True deleltions") + ylab("Number of accessions") +
         ggtitle("True deletions per accession") +
-        ggsave("../Plots/deletions_per_accession.png", height = 5, width = 10, dpi = 600)
+        ggsave("../Plots/deletions_per_accession.pdf", height = 10, width = 15, units = "cm")
 
 # insertions
 dat <- read_tsv("../RawData/TEPID_TEPAV.tsv.gz") %>%
@@ -61,4 +61,4 @@ ins_df %>%
         annotate("text", x = 410, y = 15, label = "Col-0") +
         xlab("TE insertions") + ylab("Number of accessions") +
         ggtitle("True insertions per accession") + 
-        ggsave("../Plots/insertions_per_accession.png", height = 5, width = 10, dpi = 600)
+        ggsave("../Plots/insertions_per_accession.pdf", height = 10, width = 15, units = "cm")

### 2016-11-23 ProcessingCode/simulated_TE_variants/RSVSim_commands.R
Add code for simulating TE variants
closes #13

@@ -1,0 +1,62 @@
+library(RSVSim)
+
+# load genome
+# first download genome from TAIR
+tair10 = readDNAStringSet("../../RawData/tair10.fa.gz")
+
+# load data and read into correct format
+# Must run the python code first to make these files
+del_tair_r1 <- read.table("tair_del_r1.bed", header=TRUE)
+del_tair_r2 <- read.table("tair_del_r2.bed", header=TRUE)
+del_tair_r3 <- read.table("tair_del_r3.bed", header=TRUE)
+del_tair_r4 <- read.table("tair_del_r4.bed", header=TRUE)
+del_tair_r5 <- read.table("tair_del_r5.bed", header=TRUE)
+
+ins_tair_r1 <- read.table("tair_ins_r1.bed", header=TRUE)
+ins_tair_r2 <- read.table("tair_ins_r2.bed", header=TRUE)
+ins_tair_r3 <- read.table("tair_ins_r3.bed", header=TRUE)
+ins_tair_r4 <- read.table("tair_ins_r4.bed", header=TRUE)
+ins_tair_r5 <- read.table("tair_ins_r5.bed", header=TRUE)
+
+ath_deletions_r1 <- with(del_tair_r1, GRanges(chr, IRanges(start, stop)))
+ath_deletions_r2 <- with(del_tair_r2, GRanges(chr, IRanges(start, stop)))
+ath_deletions_r3 <- with(del_tair_r3, GRanges(chr, IRanges(start, stop)))
+ath_deletions_r4 <- with(del_tair_r4, GRanges(chr, IRanges(start, stop)))
+ath_deletions_r5 <- with(del_tair_r5, GRanges(chr, IRanges(start, stop)))
+
+ath_insertions_r1 <- with(ins_tair_r1, GRanges(chr, IRanges(start, stop), chrB=dest_chr, startB=dest_start, copied=is_copied))
+ath_insertions_r2 <- with(ins_tair_r2, GRanges(chr, IRanges(start, stop), chrB=dest_chr, startB=dest_start, copied=is_copied))
+ath_insertions_r3 <- with(ins_tair_r3, GRanges(chr, IRanges(start, stop), chrB=dest_chr, startB=dest_start, copied=is_copied))
+ath_insertions_r4 <- with(ins_tair_r4, GRanges(chr, IRanges(start, stop), chrB=dest_chr, startB=dest_start, copied=is_copied))
[…]

### 2016-11-23 ProcessingCode/simulated_TE_variants/generate_sim_coords.py

@@ -1,0 +1,91 @@
+#! /usr/bin/env python
+
+import random
+
+
+class sim_data():
+
+    def __init__(self):
+        self.chromosomes = {}
+        self.transposons = {}
+
+    def read_genome_index(self, index):
+        """
+        reads and stores chromosome information from genome indexed by samtools faidx
+        """
+        with open(index, 'r') as infile:
+            for line in infile:
+                line = line.rsplit()
+                if 'super' in line[0] or 'scaffold' in line[0]:
+                    pass
+                elif line[0].startswith('chr'):
+                    self.chromosomes[line[0]] = int(line[1])
+                else:
+                    pass
+
+    def read_TE_file(self, TE_file):
+        """
+        reads and stores transposon annotation
+        """
+        with open(TE_file, 'r') as infile:
+            for line in infile:
+                line = line.rsplit()
+                self.transposons[line[4]] = {'chrom':line[0], 'start':line[1], 'stop':line[2]}
+
+    def create_deletion_coords(self, outfile, n):
+        """
+        randomly select n transposon coordinates and write to outfile
+        """
+        with open(outfile, 'w+') as dels:
+            dels.write("chr\tstart\tstop\tTE\n")
+            for x in xrange(n):
+                selection = random.choice(self.transposons.keys())
+                info = self.transposons[selection]
+                dels.write("{ch}\t{sta}\t{stp}\t{TE}\n".format(ch=info['chrom'],
+                                                               sta=info['start'],
+                                                               stp=info['stop'],
+                                                               TE=selection))
+
+    def create_insertion_coords(self, outfile, percCopied, n):
+        """
+        randomly select n transposon coordinates and randomly generate insertion points
+        save to outfile
+        """
+        n_copied = n*percCopied
+        with open(outfile, 'w+') as ins:
+            ins.write("chr\tstart\tstop\tdest_chr\tdest_start\tis_copied\n")
+            for x in xrange(n):
+                selection = random.choice(self.transposons.keys())
+                info = self.transposons[selection]
+                dest_chrom = random.choice(self.chromosomes.keys())
+                dest_start = random.randint(0, self.chromosomes[dest_chrom])
+                if n_copied > 0:
+                    copied = 'TRUE'
+                    n_copied -= 1
+                else:
+                    copied = 'FALSE'
+                ins.write("{ch}\t{sta}\t{stp}\t{dest_ch}\t{dest_sta}\t{copied}\n".format(ch=info['chrom'],
+                                                                                         sta=info['start'],
[…]

### 2016-11-23 ProcessingCode/simulated_TE_variants/var_coverage.py

@@ -1,0 +1,30 @@
+from subprocess import call
+from argparse import ArgumentParser
+import os
+
+
+parser = ArgumentParser(description='Simulate illumina reads of variable coverages from fasta file')
+parser.add_argument('-g', '--genome', help='genome fasta file', required=True)
+parser.add_argument('-d', '--distance', help='outer distance between the two ends', required=True, type=str)
+parser.add_argument('-c', '--coverages', help='comma separated list of coverages to simulate', required=True)
+parser.add_argument('-s', '--std', help='standard deviation', required=True, type=str)
+parser.add_argument('-p', '--pe', help='paired end reads', default=True, type=bool, required=False)
+parser.add_argument('-l', '--length', help='read length', required=True, type=str)
+options = parser.parse_args()
+
+
+def sim_read(options, cov):
+    name = cov+'x'
+    if options.pe is True:
+        multiplier = 2
+    else:
+        multiplier = 1
+    number_read = (int(cov) * 120000000) / (int(options.length) * multiplier)
+    if not os.path.exists('./'+name):
+        os.makedirs('./'+name)
+    call(['wgsim', '-d', options.distance, '-s', options.std, '-N', str(number_read), '-1', options.length, '-2', options.length, options.genome, './{c}/{c}_1.fastq'.format(c=name), './{c}/{c}_2.fastq'.format(c=name)])
+
+
+coverages = options.coverages.split(',')
+for cov in coverages:
+    sim_read(options, cov)
